---
title: "International Finance"
subtitle: "Session 2 — Interest Rates, Exchange Rates and the Climate Reading"
author: "Laura Contreras-Portela"
institute: "KEDGE Business School"
date: today
date-format: "MMMM YYYY"
format:
  revealjs:
    theme: [default, ../_theme/theme.scss]
    width: 1600
    height: 900
    title-slide-attributes:
      data-background-color: "#FAF7F2"
    slide-number: c/t
    show-slide-number: all
    progress: true
    history: true
    hash: true
    controls: true
    navigation-mode: linear      # right arrow always goes to the next slide
    preview-links: auto
    highlight-style: github
    code-line-numbers: true
    html-math-method: katex
    chalkboard:                     # C = draw on the slide, B = blank board
      buttons: true
      theme: whiteboard
      chalk-effect: 0
    footer: "International Finance · International Financial System · KEDGE"
    embed-resources: false          # true -> one portable file, no wifi needed
  beamer:                           # same source -> a real LaTeX PDF handout
    aspectratio: 169
    theme: metropolis
    incremental: false
    header-includes: |
      \definecolor{ifteal}{HTML}{0E5250}
      \definecolor{ifclay}{HTML}{B4573C}
      \setbeamercolor{frametitle}{bg=ifteal}
      \setbeamercolor{progress bar}{fg=ifteal}
      \setbeamercolor{title separator}{fg=ifteal}
      \setbeamercolor{structure}{fg=ifteal}
      \setbeamercolor{alerted text}{fg=ifclay}
execute:
  echo: false
  warning: false
  message: false
  fig-align: center
  fig-width: 8
  fig-height: 4.2
knitr:
  opts_chunk:
    dev: png
    dpi: 300
---

```{r setup}
source("../_R/setup.R")
```


## Today's lecture

1. The **interest rate** — the price of money  
2. The **exchange rate** — the price of a currency, and who pays to defend it  
3. **Climate and the financial system** — discussion of [Monasterolo (2020)](https://www.annualreviews.org/docserver/fulltext/resource/12/1/annurev-resource-110119-031134.pdf?expires=1788964584&id=id&accname=guest&checksum=4408CBC7A9F5E1E1ECD8BCE5667B6E7E)

## But before... a few questions

::: ask
::: {.incremental}

- What are the main ways of **channeling savings to borrowers**?
- How do firms participate in **international markets**? Why do they participate?
- Is **international trade** always good for everyone?
- What can we use as an **antidote to policy bias**?
- How do we keep track of all the **transactions between a country and the rest of the world**?
- How do we record **income from goods sold abroad**, a **foreign direct investment**, or the **purchase of U.S. dollars**?

:::
:::

. . .

So far, we have assumed that funds move from **savers** to **borrowers** at no cost.
But is money really *free*?


# 3 · The Interest Rate {.divider .center background-color="#06302F"}

## The week a number moved a pension fund {.hook background-color="#FAF7F2"}

September 2022, the UK.

After the mini-budget, **gilt yields jumped**. Pension funds suddenly needed more cash to cover their investment positions. To raise it, they sold gilts — pushing yields even higher.

The Bank of England eventually stepped in and bought gilts.

::: {.incremental}
- A change in an **interest rate** had triggered a chain reaction across the financial system.
- Interest rates affect **asset prices, borrowing costs, and financial stability**.
:::

::: {.notes}
Before defining an interest rate, let's see why we care about this number.

In September 2022, the UK government announced what became known as the **mini-budget**. Financial markets reacted very strongly.

A **gilt** is simply the British name for a government bond: the government borrows money from investors and promises to repay it in the future.

Gilt prices fell and their yields increased sharply.

Some pension funds had investment positions that required them to provide more cash when gilt prices fell. So suddenly, they needed a lot of cash.

How do you get cash quickly? **You sell assets.**

Many pension funds started selling gilts at the same time.

This created a feedback loop:

**gilt prices fall → yields rise → pension funds need cash → sell gilts → gilt prices fall further → yields rise further**

The point is that an interest rate can look like a boring number — 2%, 4%, 5% — but when it moves, it changes asset prices, borrowing costs, and potentially financial stability.

**So, what exactly is an interest rate?**
:::


. . .

::: ask
What is the interest rate, how is it fixed? 
:::




## 3 · Supply and demand for loanable funds

The interest rate is the price that clears the market for **loanable funds**
. . .

```{=html}
<div id="loanable-demo" style="
  max-width:1100px;
  margin:0 auto;
  font-family:inherit;
">

  <div style="
    display:grid;
    grid-template-columns:minmax(0, 3fr) minmax(210px, 1fr);
    gap:22px;
    align-items:stretch;
  ">

    <!-- ================================================== -->
    <!-- LEFT: GRAPH + CONTROLS                             -->
    <!-- ================================================== -->

    <div style="min-width:0;">

      <svg
        id="loanable-chart"
        viewBox="0 0 800 500"
        style="
          width:100%;
          height:auto;
          display:block;
        "
        aria-label="Interactive market for loanable funds">
      </svg>


      <!-- CONTROLS -->

      <div style="
        margin-top:2px;
        border-top:1px solid #DDD;
        padding-top:10px;
      ">

        <!-- DEMAND -->

        <div style="
          display:grid;
          grid-template-columns:115px 1fr 90px;
          gap:10px;
          align-items:center;
          margin-bottom:8px;
        ">

          <div style="
            font-weight:700;
            color:#3579D8;
            font-size:0.82em;
          ">
            Demand (D)
          </div>

          <div style="
            display:flex;
            align-items:center;
            gap:7px;
          ">

            <span style="
              font-size:0.67em;
              color:#777;
            ">
              Decrease
            </span>

            <input
              id="demand-slider"
              type="range"
              min="-2"
              max="2"
              step="0.1"
              value="0"
              style="
                flex:1;
                min-width:0;
                accent-color:#3579D8;
              "
            >

            <span style="
              font-size:0.67em;
              color:#777;
            ">
              Increase
            </span>

          </div>

          <div
            id="demand-value"
            style="
              text-align:center;
              font-weight:700;
              color:#3579D8;
              font-size:0.75em;
            ">
            Baseline
          </div>

        </div>


        <!-- SUPPLY -->

        <div style="
          display:grid;
          grid-template-columns:115px 1fr 90px;
          gap:10px;
          align-items:center;
        ">

          <div style="
            font-weight:700;
            color:#4FA56A;
            font-size:0.82em;
          ">
            Supply (S)
          </div>

          <div style="
            display:flex;
            align-items:center;
            gap:7px;
          ">

            <span style="
              font-size:0.67em;
              color:#777;
            ">
              Decrease
            </span>

            <input
              id="supply-slider"
              type="range"
              min="-2"
              max="2"
              step="0.1"
              value="0"
              style="
                flex:1;
                min-width:0;
                accent-color:#4FA56A;
              "
            >

            <span style="
              font-size:0.67em;
              color:#777;
            ">
              Increase
            </span>

          </div>

          <div
            id="supply-value"
            style="
              text-align:center;
              font-weight:700;
              color:#4FA56A;
              font-size:0.75em;
            ">
            Baseline
          </div>

        </div>

      </div>

    </div>


    <!-- ================================================== -->
    <!-- RIGHT: OUTCOMES                                    -->
    <!-- ================================================== -->

    <div style="
      border-left:1px solid #DDD;
      padding-left:24px;
      display:flex;
      flex-direction:column;
      justify-content:center;
      gap:30px;
      min-width:0;
    ">

      <!-- INTEREST RATE -->

      <div>

        <div style="
          font-size:0.78em;
          color:#666;
          margin-bottom:5px;
        ">
          Interest rate
        </div>

        <div
          id="rate-result"
          style="
            font-size:1.35em;
            font-weight:750;
          ">
          Baseline
        </div>

      </div>


      <!-- QUANTITY -->

      <div>

        <div style="
          font-size:0.78em;
          color:#666;
          margin-bottom:5px;
        ">
          Quantity of loanable funds
        </div>

        <div
          id="quantity-result"
          style="
            font-size:1.35em;
            font-weight:750;
          ">
          Baseline
        </div>

      </div>


      <!-- RESET -->

      <button
        id="loanable-reset"
        type="button"
        style="
          margin-top:5px;
          align-self:flex-start;
          padding:5px 13px;
          border:1px solid #AAA;
          border-radius:6px;
          background:white;
          cursor:pointer;
          font-size:0.72em;
          font-weight:600;
        ">
        Reset
      </button>

    </div>

  </div>

</div>


<script>
(() => {

  const root =
    document.getElementById("loanable-demo");

  if (!root ||
      root.dataset.initialized === "true") return;

  root.dataset.initialized = "true";


  const svg =
    document.getElementById("loanable-chart");

  const demandSlider =
    document.getElementById("demand-slider");

  const supplySlider =
    document.getElementById("supply-slider");

  const demandValue =
    document.getElementById("demand-value");

  const supplyValue =
    document.getElementById("supply-value");

  const rateResult =
    document.getElementById("rate-result");

  const quantityResult =
    document.getElementById("quantity-result");

  const resetButton =
    document.getElementById("loanable-reset");


  // =========================================================
  // MODEL
  // =========================================================

  const a = 9;
  const b = 0.65;

  const c = 1.2;
  const d = 0.65;


  function demandRate(q, shift) {

    return a - b * (q - shift);

  }


  function supplyRate(q, shift) {

    return c + d * (q - shift);

  }


  const Q0 =
    (a - c) / (b + d);

  const r0 =
    demandRate(Q0, 0);


  // =========================================================
  // GRAPH DIMENSIONS
  // =========================================================

  const W = 800;
  const H = 500;

  const margin = {
    top:42,
    right:55,
    bottom:62,
    left:80
  };

  const plotW =
    W - margin.left - margin.right;

  const plotH =
    H - margin.top - margin.bottom;

  const qMin = 0;
  const qMax = 12;

  const rMin = 0;
  const rMax = 10;


  const x = q =>
    margin.left +
    ((q - qMin) /
    (qMax - qMin)) *
    plotW;


  const y = rate =>
    margin.top +
    plotH -
    ((rate - rMin) /
    (rMax - rMin)) *
    plotH;


  // =========================================================
  // SVG HELPERS
  // =========================================================

  function el(name, attrs = {}) {

    const node =
      document.createElementNS(
        "http://www.w3.org/2000/svg",
        name
      );

    for (const [key, value]
         of Object.entries(attrs)) {

      node.setAttribute(key, value);

    }

    return node;

  }


  function line(x1, y1, x2, y2, attrs = {}) {

    const node = el("line", {
      x1:x1,
      y1:y1,
      x2:x2,
      y2:y2,
      ...attrs
    });

    svg.appendChild(node);

    return node;

  }


  function text(txt, tx, ty, attrs = {}) {

    const node = el("text", {
      x:tx,
      y:ty,
      ...attrs
    });

    node.textContent = txt;

    svg.appendChild(node);

    return node;

  }


  function circle(cx, cy, radius, attrs = {}) {

    const node = el("circle", {
      cx:cx,
      cy:cy,
      r:radius,
      ...attrs
    });

    svg.appendChild(node);

    return node;

  }


  // =========================================================
  // AXES
  // =========================================================

  line(
    x(0), y(0),
    x(qMax), y(0),
    {
      stroke:"#333",
      "stroke-width":1.7
    }
  );


  line(
    x(0), y(0),
    x(0), y(rMax),
    {
      stroke:"#333",
      "stroke-width":1.7
    }
  );


  text(
    "Interest rate (r)",
    x(0) + 5,
    margin.top - 14,
    {
      "font-size":19,
      "font-weight":700,
      fill:"#222"
    }
  );


  text(
    "Quantity of loanable funds (Q)",
    x(qMax / 2),
    H - 13,
    {
      "text-anchor":"middle",
      "font-size":19,
      "font-weight":700,
      fill:"#222"
    }
  );


  // =========================================================
  // BASELINE CURVES
  // =========================================================

  const baseDemand =
    el("path", {
      fill:"none",
      stroke:"#A9C3E8",
      "stroke-width":2.5,
      "stroke-dasharray":"8 7"
    });

  svg.appendChild(baseDemand);


  const baseSupply =
    el("path", {
      fill:"none",
      stroke:"#A8CFB4",
      "stroke-width":2.5,
      "stroke-dasharray":"8 7"
    });

  svg.appendChild(baseSupply);


  baseDemand.setAttribute(
    "d",
    "M " +
    x(0) + " " +
    y(demandRate(0, 0)) +
    " L " +
    x(12) + " " +
    y(demandRate(12, 0))
  );


  baseSupply.setAttribute(
    "d",
    "M " +
    x(0) + " " +
    y(supplyRate(0, 0)) +
    " L " +
    x(12) + " " +
    y(supplyRate(12, 0))
  );


  // =========================================================
  // BASELINE EQUILIBRIUM
  // =========================================================

  line(
    x(0),
    y(r0),
    x(Q0),
    y(r0),
    {
      stroke:"#999",
      "stroke-width":1.5,
      "stroke-dasharray":"6 6"
    }
  );


  line(
    x(Q0),
    y(0),
    x(Q0),
    y(r0),
    {
      stroke:"#999",
      "stroke-width":1.5,
      "stroke-dasharray":"6 6"
    }
  );


  circle(
    x(Q0),
    y(r0),
    6,
    {
      fill:"#999"
    }
  );


  text(
    "r₀",
    x(0) - 12,
    y(r0) + 6,
    {
      "text-anchor":"end",
      "font-size":19,
      "font-weight":700,
      fill:"#777"
    }
  );


  text(
    "Q₀",
    x(Q0),
    y(0) + 28,
    {
      "text-anchor":"middle",
      "font-size":19,
      "font-weight":700,
      fill:"#777"
    }
  );


  // =========================================================
  // ACTIVE CURVES
  // =========================================================

  const demandPath =
    el("path", {
      fill:"none",
      stroke:"#3579D8",
      "stroke-width":5
    });

  svg.appendChild(demandPath);


  const supplyPath =
    el("path", {
      fill:"none",
      stroke:"#4FA56A",
      "stroke-width":5
    });

  svg.appendChild(supplyPath);


  const demandLabel =
    text(
      "D₀",
      0,
      0,
      {
        "font-size":25,
        "font-style":"italic",
        "font-weight":700,
        fill:"#3579D8"
      }
    );


  const supplyLabel =
    text(
      "S₀",
      0,
      0,
      {
        "font-size":25,
        "font-style":"italic",
        "font-weight":700,
        fill:"#4FA56A"
      }
    );


  // =========================================================
  // NEW EQUILIBRIUM
  // =========================================================

  const rateGuide =
    line(
      0,0,0,0,
      {
        stroke:"#3579D8",
        "stroke-width":2,
        "stroke-dasharray":"7 6"
      }
    );


  const quantityGuide =
    line(
      0,0,0,0,
      {
        stroke:"#3579D8",
        "stroke-width":2,
        "stroke-dasharray":"7 6"
      }
    );


  const equilibrium =
    circle(
      0,
      0,
      8,
      {
        fill:"#222",
        stroke:"white",
        "stroke-width":2
      }
    );


  const rateLabel =
    text(
      "",
      0,
      0,
      {
        "font-size":19,
        "font-weight":700,
        fill:"#3579D8"
      }
    );


  const quantityLabel =
    text(
      "",
      0,
      0,
      {
        "font-size":19,
        "font-weight":700,
        fill:"#3579D8"
      }
    );


  // =========================================================
  // UPDATE
  // =========================================================

  function update() {

    const demandShift =
      Number(demandSlider.value);

    const supplyShift =
      Number(supplySlider.value);


    // -------------------------------------------------------
    // EQUILIBRIUM
    // -------------------------------------------------------

    const Q =
      (
        a -
        c +
        b * demandShift +
        d * supplyShift
      ) /
      (b + d);


    const rate =
      demandRate(
        Q,
        demandShift
      );


    // -------------------------------------------------------
    // MOVE DEMAND CURVE
    // -------------------------------------------------------

    demandPath.setAttribute(
      "d",
      "M " +
      x(0) + " " +
      y(demandRate(0, demandShift)) +
      " L " +
      x(12) + " " +
      y(demandRate(12, demandShift))
    );


    // -------------------------------------------------------
    // MOVE SUPPLY CURVE
    // -------------------------------------------------------

    supplyPath.setAttribute(
      "d",
      "M " +
      x(0) + " " +
      y(supplyRate(0, supplyShift)) +
      " L " +
      x(12) + " " +
      y(supplyRate(12, supplyShift))
    );


    // -------------------------------------------------------
    // CURVE LABELS
    // -------------------------------------------------------

    demandLabel.setAttribute(
      "x",
      x(10.7)
    );

    demandLabel.setAttribute(
      "y",
      y(
        demandRate(
          10.7,
          demandShift
        )
      ) - 10
    );


    demandLabel.textContent =
      Math.abs(demandShift) < 0.05
      ? "D₀"
      : "D₁";


    supplyLabel.setAttribute(
      "x",
      x(10.2)
    );

    supplyLabel.setAttribute(
      "y",
      y(
        supplyRate(
          10.2,
          supplyShift
        )
      ) - 10
    );


    supplyLabel.textContent =
      Math.abs(supplyShift) < 0.05
      ? "S₀"
      : "S₁";


    // -------------------------------------------------------
    // NEW EQUILIBRIUM GUIDES
    // -------------------------------------------------------

    rateGuide.setAttribute(
      "x1",
      x(0)
    );

    rateGuide.setAttribute(
      "x2",
      x(Q)
    );

    rateGuide.setAttribute(
      "y1",
      y(rate)
    );

    rateGuide.setAttribute(
      "y2",
      y(rate)
    );


    quantityGuide.setAttribute(
      "x1",
      x(Q)
    );

    quantityGuide.setAttribute(
      "x2",
      x(Q)
    );

    quantityGuide.setAttribute(
      "y1",
      y(0)
    );

    quantityGuide.setAttribute(
      "y2",
      y(rate)
    );


    equilibrium.setAttribute(
      "cx",
      x(Q)
    );

    equilibrium.setAttribute(
      "cy",
      y(rate)
    );


    rateLabel.setAttribute(
      "x",
      x(0) - 12
    );

    rateLabel.setAttribute(
      "y",
      y(rate) + 6
    );

    rateLabel.setAttribute(
      "text-anchor",
      "end"
    );


    quantityLabel.setAttribute(
      "x",
      x(Q)
    );

    quantityLabel.setAttribute(
      "y",
      y(0) + 28
    );

    quantityLabel.setAttribute(
      "text-anchor",
      "middle"
    );


    // -------------------------------------------------------
    // BASELINE / NEW EQUILIBRIUM
    // -------------------------------------------------------

    const baseline =
      Math.abs(demandShift) < 0.05 &&
      Math.abs(supplyShift) < 0.05;


    if (baseline) {

      rateGuide.setAttribute(
        "visibility",
        "hidden"
      );

      quantityGuide.setAttribute(
        "visibility",
        "hidden"
      );

      equilibrium.setAttribute(
        "visibility",
        "hidden"
      );

      rateLabel.setAttribute(
        "visibility",
        "hidden"
      );

      quantityLabel.setAttribute(
        "visibility",
        "hidden"
      );

    } else {

      rateGuide.setAttribute(
        "visibility",
        "visible"
      );

      quantityGuide.setAttribute(
        "visibility",
        "visible"
      );

      equilibrium.setAttribute(
        "visibility",
        "visible"
      );

      rateLabel.setAttribute(
        "visibility",
        "visible"
      );

      quantityLabel.setAttribute(
        "visibility",
        "visible"
      );


      rateLabel.textContent =
        "r₁";

      quantityLabel.textContent =
        "Q₁";

    }


    // -------------------------------------------------------
    // SLIDER LABELS
    // -------------------------------------------------------

    if (Math.abs(demandShift) < 0.05) {

      demandValue.textContent =
        "Baseline";

    } else if (demandShift > 0) {

      demandValue.textContent =
        "Increase";

    } else {

      demandValue.textContent =
        "Decrease";

    }


    if (Math.abs(supplyShift) < 0.05) {

      supplyValue.textContent =
        "Baseline";

    } else if (supplyShift > 0) {

      supplyValue.textContent =
        "Increase";

    } else {

      supplyValue.textContent =
        "Decrease";

    }


    // =======================================================
    // ECONOMIC INTERPRETATION
    // =======================================================

    const dUp =
      demandShift > 0.05;

    const dDown =
      demandShift < -0.05;

    const sUp =
      supplyShift > 0.05;

    const sDown =
      supplyShift < -0.05;

    const dSame =
      Math.abs(demandShift) <= 0.05;

    const sSame =
      Math.abs(supplyShift) <= 0.05;


    // Baseline

    if (dSame && sSame) {

      rateResult.textContent =
        "Baseline";

      quantityResult.textContent =
        "Baseline";

    }


    // Demand increases

    else if (dUp && sSame) {

      rateResult.innerHTML =
        "↑ Increases";

      quantityResult.innerHTML =
        "↑ Increases";

    }


    // Demand decreases

    else if (dDown && sSame) {

      rateResult.innerHTML =
        "↓ Decreases";

      quantityResult.innerHTML =
        "↓ Decreases";

    }


    // Supply increases

    else if (dSame && sUp) {

      rateResult.innerHTML =
        "↓ Decreases";

      quantityResult.innerHTML =
        "↑ Increases";

    }


    // Supply decreases

    else if (dSame && sDown) {

      rateResult.innerHTML =
        "↑ Increases";

      quantityResult.innerHTML =
        "↓ Decreases";

    }


    // Demand and supply increase

    else if (dUp && sUp) {

      rateResult.innerHTML =
        "? Ambiguous";

      quantityResult.innerHTML =
        "↑ Increases";

    }


    // Demand and supply decrease

    else if (dDown && sDown) {

      rateResult.innerHTML =
        "? Ambiguous";

      quantityResult.innerHTML =
        "↓ Decreases";

    }


    // Demand increases, supply decreases

    else if (dUp && sDown) {

      rateResult.innerHTML =
        "↑ Increases";

      quantityResult.innerHTML =
        "? Ambiguous";

    }


    // Demand decreases, supply increases

    else if (dDown && sUp) {

      rateResult.innerHTML =
        "↓ Decreases";

      quantityResult.innerHTML =
        "? Ambiguous";

    }

  }


  // =========================================================
  // CONTROLS
  // =========================================================

  demandSlider.addEventListener(
    "input",
    update
  );


  supplySlider.addEventListener(
    "input",
    update
  );


  resetButton.addEventListener(
    "click",
    () => {

      demandSlider.value = 0;

      supplySlider.value = 0;

      update();

    }
  );


  // =========================================================
  // INITIAL DRAW
  // =========================================================

  update();

})();
</script>
```


## 3 · What shifts the curves?
:::: {.columns}

::: {.column width="50%"}

### Supply of funds

::: {.incremental}

- **Wealth of fund suppliers rises**: more funds available for investment
- **Risk of securities falls**: supplying funds becomes more attractive
- **Near-term spending needs fall**: more funds available for investment
- **Monetary expansion**: more funds available
- **Economic conditions improve**: greater inflows of funds

:::

:::

::: {.column width="50%"}

### Demand for funds

::: {.incremental}

- **Utility from assets bought with borrowed funds rises**: borrowing increases
- **Non-price borrowing conditions become less restrictive**: borrowing increases
- **Economic conditions improve**: borrowing increases

:::

:::

::::

. . .

::: ask
Remember the **UK gilt crisis** from the beginning of the lecture?

Gilt prices fell and yields rose sharply. In our loanable-funds framework, **what could have shifted?**
:::

## 3 · The interest rate over the business cycle

Economic conditions can shift **both the supply and the demand for funds**

```{r}
#| fig-width: 10
#| fig-height: 4.0
tb <- fred("TB3MS")
if (is.null(tb)) {
  no_data_note("TB3MS")
} else {
  ggplot(tb, aes(date, value)) +
    recession_bands() +
    geom_line(colour = IF_COL$teal700, linewidth = .5) +
    labs(x = NULL, y = "Percent",
         title = "3-month Treasury bill, United States",
         caption = "Shaded: NBER recessions. Source: Board of Governors of the Federal Reserve System via FRED.") +
    theme_if()
}
```

::: ask
What is the relationship between the business cycle and interest rates — and which way does the causality run?
:::


::: {.notes}
Economic conditions can affect both the supply and the demand for funds.

Here I am showing the interest rate on a **3-month U.S. Treasury bill** short-term government security, useful measure of short-term market interest rates

The shaded areas are **recessions**

**What pattern do you see around recessions?**

Interest rates tend to be relatively high around the end of expansions and then fall during recessions. As the economy recovers and expands, they tend to rise again.

**Why?**

During an expansion, firms may see more profitable investment opportunities and households may want to borrow more. What happens to the **demand for funds**? Upward pressure on interest rates.

During a downturn, investment opportunities and borrowing demand tend to weaken, which can put downward pressure on rates.
Interest rates affect borrowing, investment, consumption, asset prices, and therefore economic activity.

Causality also runs in the **other direction**.

But economic activity is not the only thing changing over the business cycle.

One particularly important variable is **inflation**.

And there is another important actor here: the **central bank** and monetary policy! 

Central banks respond to economic conditions. How? 
When inflationary pressures build or the economy is running strongly, they may raise short-term interest rates. During recessions, they often cut rates to support economic activity.

So the relationship we see in this graph reflects several mechanisms operating simultaneously:

**economic conditions → supply and demand for funds → interest rates**

but also

**monetary policy → inflation →  interest rates → economic activity**
:::


## 3 · The interest rate and the CPI

```{r}
#| fig-width: 10
#| fig-height: 4.0
tb  <- fred("TB3MS")
cpi <- fred("CPIAUCSL")
if (is.null(tb) || is.null(cpi)) {
  no_data_note("TB3MS / CPIAUCSL")
} else {
  cpi$infl <- c(rep(NA, 12), diff(log(cpi$value), lag = 12) * 100)
  both <- rbind(
    data.frame(date = tb$date,  value = tb$value,  series = "Interest rate (3-m T-bill)"),
    data.frame(date = cpi$date, value = cpi$infl,  series = "Inflation (CPI, yoy)")
  )
  both <- both[!is.na(both$value) & both$date >= as.Date("1964-01-01"), ]
  ggplot(both, aes(date, value, colour = series)) +
    recession_bands(as.Date("1964-01-01")) +
    geom_hline(yintercept = 0, colour = IF_COL$ink, linewidth = .4) +
    geom_line(linewidth = .55) +
    scale_colour_manual(values = c(IF_COL$clay, IF_COL$teal700)) +
    labs(x = NULL, y = "Percent",
         caption = "Source: Board of Governors of the Federal Reserve System; BLS via FRED.") +
    theme_if()
}
```

::: ask
What is the relationship between the interest rate and the CPI?
:::

::: {.notes}
What do you see?

Interest rates and inflation tend to move together: periods of higher inflation are generally associated with higher interest rates.

But again, correlation is not causation.

Why might inflation matter for interest rates? And why might interest rates, in turn, matter for inflation?

This is where **monetary policy** becomes central.
:::


## 3 · Real and nominal

The real rate is the nominal rate adjusted for **expected** changes in the price level. It reflects the true cost of borrowing
$$i_r = i - \pi^e$$ 

```{r}
#| fig-width: 10
#| fig-height: 4
tb  <- fred("TB3MS")
cpi <- fred("CPIAUCSL")
if (is.null(tb) || is.null(cpi)) {
  no_data_note("TB3MS / CPIAUCSL")
} else {
  cpi$infl <- c(rep(NA, 12), diff(log(cpi$value), lag = 12) * 100)
  m <- merge(tb, cpi[, c("date", "infl")], by = "date")
  m <- m[!is.na(m$infl) & m$date >= as.Date("1960-01-01"), ]
  s <- rbind(
    data.frame(date = m$date, value = m$value,          series = "Nominal rate"),
    data.frame(date = m$date, value = m$value - m$infl, series = "Real rate")
  )
  ggplot(s, aes(date, value, colour = series)) +
    recession_bands(as.Date("1960-01-01")) +
    geom_hline(yintercept = 0, colour = IF_COL$ink, linewidth = .4) +
    geom_line(linewidth = .55) +
    scale_colour_manual(values = c(IF_COL$slate, IF_COL$teal700)) +
    labs(x = NULL, y = "Percent") +
    theme_if()
}
```

::: incremental

- When the real rate is low, there are greater incentives to borrow and less to lend
- The real interest rate adjusts for changes in *purchasing power* and therefore captures the economically relevant cost of borrowing and return to saving

::: 

## 3 · Risk and term structure

Different bonds offer different yields so that interest rates differ across financial assets 

::: panel-tabset
### Default risk

The risk that the issuer is unable or unwilling to make interest payments as promised. Bonds with no default risk — U.S. Treasuries, which can raise taxes or print money — are the benchmark. The **risk premium** is the spread between a bond with default risk and a default-free bond.

### Liquidity

A liquid asset is quickly and cheaply converted into cash. Demand is higher for more liquid assets, so the premium reflects **both** default risk and liquidity.

### Income Tax Considerations

Tax exemption acts like an increase in expected return. U.S. municipal bonds are exempt from federal income tax; Treasuries are exempt from state and local income tax.

### Maturity

Interest rates also differ according to **how long funds are lent**. The relationship between interest rates and maturity is the **term structure of interest rates**, summarized by the **yield curve**.

:::

::: {.notes}
Until now, our model gave us one equilibrium interest rate.

But open any financial newspaper and you will see many: government bond rates, corporate bond rates, 3-month rates, 10-year rates, mortgage rates...

Why are they different?

Because a euro lent through one financial asset is not necessarily equivalent to a euro lent through another.

Investors care about whether they will be repaid, how easily they can sell the asset, how returns are taxed, and how long their money is committed.

These characteristics generate differences in the interest rates we observe.
:::


## 3 · Yield curves

The **yield curve** describes the relationship between the **yield to maturity** and the **maturity** of debt securities with **comparable** risk, liquidity, and tax characteristics. It summarizes the relationship between **short-term and long-term interest rates**

:::: columns
::: {.column width="52%"}
```{r}
#| fig-width: 6
#| fig-height: 4.2

m <- c(0.25, 0.5, 1, 2, 3, 5, 7, 10, 20, 30)

yc <- rbind(
  data.frame(
    m = m,
    y = 1.0 + 2.0 * (1 - exp(-m / 6)),
    shape = "Upward sloping"
  ),
  data.frame(
    m = m,
    y = 4.9 + 0.02 * m,
    shape = "Flat"
  ),
  data.frame(
    m = m,
    y = 5.1 - 0.9 * (1 - exp(-m / 4)) + 0.01 * m,
    shape = "Inverted"
  )
)

ggplot(yc, aes(m, y, colour = shape)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.6) +
  scale_colour_manual(
    values = c(
      "Upward sloping" = IF_COL$teal700,
      "Flat"           = IF_COL$slate,
      "Inverted"       = IF_COL$clay
    )
  ) +
  scale_x_continuous(breaks = c(1, 5, 10, 20, 30)) +
  labs(
    x = "Maturity (years)",
    y = "Yield (%)",
    title = "Alternative shapes of the yield curve"
  ) +
  theme_if()
```
:::

::: {.column width="45%"}
::: incremental
- **Upward sloping**: Long-term yields exceed short-term yields
- **Flat**: Short- and long-term yields are approximately equal
- **Inverted**: Short-term yields exceed long-term yields
:::

::: ask
What economic information can be inferred from the **slope of the yield curve**?
:::

::: {.source}
Live curve: [U.S. Treasury daily yield curve rates](https://home.treasury.gov/resource-center/data-chart-center/interest-rates/TextView?type=daily_treasury_yield_curve).
:::
:::
::::

::: {.notes}
The yield curve isolates the **maturity dimension** of interest rates by comparing securities with similar risk, liquidity, and tax characteristics.

An upward-sloping curve is the most common configuration: yields increase with maturity. A flat curve implies little difference between short- and long-term yields, while an inverted curve implies that short-term yields exceed long-term yields.

The slope is economically informative because long-term yields depend, among other factors, on expectations about the future path of short-term interest rates.

Changes in the yield curve can therefore contain information about market expectations regarding future monetary and economic conditions.
:::


# 4 · The exchange rate {.divider .center background-color="#06302F"}

## 15 January 2015 {.hook background-color="#FAF7F2"}

For more than three years, the Swiss National Bank had maintained a **minimum exchange rate of CHF 1.20 per euro**

On 15 January 2015, it abandoned the policy.

The Swiss franc **appreciated sharply within minutes**.

::: incremental
- Exchange rates are not only **market prices** — under some regimes, they are also **policy targets**.
- Maintaining a target may require the central bank to **intervene in the foreign-exchange market**.
:::

::: ask
What does it mean for a central bank to **defend an exchange rate** — and what does it cost?
:::

::: {.notes}
We have just studied the interest rate: the price of borrowing funds.

We now turn to another fundamental price in international finance: the **exchange rate — the price of one currency in terms of another**.

The Swiss experience gives us the central question for this section.

From 2011, the Swiss National Bank prevented the franc from appreciating beyond CHF 1.20 per euro.

To maintain that exchange rate, it was prepared to intervene in the foreign-exchange market by buying foreign currency.

On 15 January 2015, it abandoned the minimum exchange rate. The franc appreciated immediately.

So an exchange-rate regime is not simply a label. It determines **whether the exchange rate is determined primarily by the market or whether the authorities commit to influencing its value**.

We will first distinguish the main exchange-rate regimes, and then study how an exchange rate is determined and how intervention works.
:::


## 4 · Exchange rate regimes
Exchange-rate regimes differ mainly in how much the exchange rate is allowed to respond to market forces** and how much authorities intervene 

<div style="font-size: 1.10em;">

| Regime | Definition | In favour | Against |
|:---|:---|:---|:---|
| **Fixed** | Rates held constant or fluctuating within very narrow boundaries | MNCs operate more easily | Government may alter the value of a currency |
| **Free float** | Rates determined by market forces, no intervention | Problems in one country spread less easily; no intervention policy needed; market efficiency enhanced | MNCs must manage the exposure; countries with severe problems may struggle |
| **Managed float** | Market forces **with** government intervention | | Governments manipulate the rate for their own benefit |
| **Pegged** | Home currency pegged to a foreign currency or unit of account | | Hard to defend against speculators in a crisis |
| **No national currency** | The country uses another currency, e.g. the U.S. dollar | | |

</div>

There is a **trade-off**: greater exchange-rate stability generally requires a stronger policy commitment to maintain that rate

::: {.notes}
Under a **fixed regime**, authorities commit to maintaining the exchange rate at, or close to, a particular value. This reduces exchange-rate uncertainty, which can facilitate international trade and investment, but maintaining the rate may require substantial intervention. HK

Under a **free float**, the exchange rate is determined primarily by supply and demand in the foreign-exchange market. The exchange rate can therefore adjust to economic shocks, but firms operating internationally face greater exchange-rate risk. US-UK

A **managed float** lies between these two: the exchange rate is market-determined, but the central bank intervenes when it considers movements excessive or undesirable. India, Colombia

Under a **peg**, the domestic currency is linked to another currency or a basket of currencies. Maintaining that link requires the authorities to defend the announced rate. DENMARK TO THE EURO

Finally, some countries give up their national currency altogether and adopt another currency. ECUADOR, PANAMA

The important point is that there is a **trade-off**: greater exchange-rate stability generally requires a stronger policy commitment to maintain that rate.

This brings us to the market itself: **what determines the exchange rate when currencies are traded?**
:::


## 4 · Exchange rate policy

The choice of exchange-rate regime determines the extent to which authorities **allow the exchange rate to adjust** or **intervene to influence it**

::: incremental
- The **spot exchange rate** is the price at which currencies are exchanged for immediate delivery. $S(j/k)$ denotes the price of one unit of currency $k$ in terms of currency $j$.

- Under managed or fixed regimes, authorities may influence or defend the exchange rate through:
    - **interest-rate policy**, affecting capital flows;
    - **foreign-exchange intervention**, buying or selling currencies;
    - **capital controls**, restricting international financial transactions.
:::

::: {.notes}
The regimes we have just seen differ in the degree to which authorities allow the exchange rate to be determined by the market.

But what does intervention actually mean?

Ultimately, the exchange rate is a **market price**. It is determined in the foreign-exchange market by the supply and demand for currencies.

Authorities can therefore influence the exchange rate either by affecting these market forces or by restricting the transactions behind them.

To understand how this works, we first need to understand the **foreign-exchange market**.
:::

## 4 · Fixed versus flexible: the pound market

At **$1.60/£**, the demand for pounds exceeds their supply

::: {.incremental}
- **Demand for £:** U.S. residents need pounds to buy British goods and assets.
- **Supply of £:** British residents supply pounds when they buy U.S. goods and assets.
- The question is: **how does the market return to equilibrium when $D_{£} > S_{£}$?**
:::

```{=html}
<div style="width:70%; display:flex; justify-content:center;">

<svg viewBox="0 0 900 520"
     style="width:72%; max-width:900px; height:auto;">

  <!-- Axes -->
  <line x1="120" y1="430" x2="830" y2="430"
        stroke="#333" stroke-width="2.5"/>
  <line x1="120" y1="430" x2="120" y2="45"
        stroke="#333" stroke-width="2.5"/>

  <!-- Axis labels -->
  <text x="480" y="500"
        text-anchor="middle"
        font-size="23"
        fill="#333">
    Quantity of British pounds
  </text>

  <text x="35" y="240"
        text-anchor="middle"
        font-size="22"
        fill="#333"
        transform="rotate(-90 35 240)">
    Dollar price per pound ($/£)
  </text>


  <!-- ========================================================= -->
  <!-- SUPPLY -->
  <!-- ========================================================= -->

  <path d="M 180 390
           C 260 375, 315 330, 355 275
           C 400 235, 440 220, 475 215
           C 535 145, 585 75, 635 45"
        fill="none"
        stroke="#176B68"
        stroke-width="4"/>

  <text x="640" y="52"
        font-size="22"
        font-weight="700"
        fill="#176B68">
    Supply (S)
  </text>


  <!-- ========================================================= -->
  <!-- ORIGINAL DEMAND D -->
  <!-- Flexible equilibrium: D = S at $1.80 -->
  <!-- ========================================================= -->

  <path d="M 250 50
           C 285 120, 350 175, 475 215
           C 585 250, 690 275, 805 290"
        fill="none"
        stroke="#3579D8"
        stroke-width="4"/>

  <text x="755" y="280"
        font-size="22"
        font-weight="700"
        fill="#3579D8">
    Demand (D)
  </text>


  <!-- ========================================================= -->
  <!-- NEW DEMAND D* -->
  <!-- Fixed equilibrium: D* = S exactly at $1.60 -->
  <!-- ========================================================= -->

  <path d="M 190 75
           C 230 145, 285 220, 355 275
           C 440 340, 565 375, 760 395"
        fill="none"
        stroke="#B66A50"
        stroke-width="3.5"
        stroke-dasharray="10 7"/>

  <text x="715" y="390"
        font-size="22"
        font-weight="700"
        fill="#B66A50">
    Demand (D*)
  </text>


  <!-- ========================================================= -->
  <!-- EXCHANGE RATE = $1.80 -->
  <!-- ========================================================= -->

  <line x1="120" y1="215" x2="475" y2="215"
        stroke="#777"
        stroke-width="1.5"
        stroke-dasharray="7 6"/>

  <text x="105" y="222"
        text-anchor="end"
        font-size="21"
        fill="#444">
    $1.80
  </text>


  <!-- ========================================================= -->
  <!-- EXCHANGE RATE = $1.60 -->
  <!-- ========================================================= -->

  <line x1="120" y1="275" x2="665" y2="275"
        stroke="#777"
        stroke-width="1.5"
        stroke-dasharray="7 6"/>

  <text x="105" y="282"
        text-anchor="end"
        font-size="21"
        fill="#444">
    $1.60
  </text>


  <!-- ========================================================= -->
  <!-- FIXED-REGIME EQUILIBRIUM: D* = S -->
  <!-- ========================================================= -->

  <circle cx="355" cy="275" r="6"
          fill="#333"/>

  <line x1="355" y1="275" x2="355" y2="430"
        stroke="#777"
        stroke-width="1.4"
        stroke-dasharray="6 6"/>

  <text x="355" y="455"
        text-anchor="middle"
        font-size="21"
        font-style="italic">
    D* = S
  </text>


  <!-- ========================================================= -->
  <!-- FLEXIBLE-REGIME EQUILIBRIUM: D = S -->
  <!-- ========================================================= -->

  <circle cx="475" cy="215" r="6"
          fill="#333"/>

  <line x1="475" y1="215" x2="475" y2="430"
        stroke="#777"
        stroke-width="1.4"
        stroke-dasharray="6 6"/>

  <text x="475" y="455"
        text-anchor="middle"
        font-size="21"
        font-style="italic">
    D = S
  </text>


  <!-- ========================================================= -->
  <!-- QUANTITY DEMANDED AT $1.60 -->
  <!-- ========================================================= -->

  <line x1="665" y1="275" x2="665" y2="430"
        stroke="#777"
        stroke-width="1.4"
        stroke-dasharray="6 6"/>

  <text x="665" y="455"
        text-anchor="middle"
        font-size="21"
        font-style="italic">
    D
  </text>


  <!-- ========================================================= -->
  <!-- TRADE DEFICIT -->
  <!-- ========================================================= -->

  <line x1="355" y1="410" x2="665" y2="410"
        stroke="#B66A50"
        stroke-width="3"/>

  <line x1="355" y1="401" x2="355" y2="419"
        stroke="#B66A50"
        stroke-width="3"/>

  <line x1="665" y1="401" x2="665" y2="419"
        stroke="#B66A50"
        stroke-width="3"/>

  <text x="510" y="395"
        text-anchor="middle"
        font-size="20"
        font-weight="600"
        fill="#B66A50">
    Trade deficit
  </text>


  <!-- ========================================================= -->
  <!-- FLEXIBLE REGIME: DOLLAR DEPRECIATES -->
  <!-- ========================================================= -->

  <line x1="150" y1="270" x2="150" y2="225"
        stroke="#3579D8"
        stroke-width="2.5"/>

  <polygon points="150,215 142,230 158,230"
           fill="#3579D8"/>

  <text x="165" y="245"
        font-size="18"
        font-weight="600"
        fill="#3579D8">
    Dollar depreciates
  </text>


  <!-- ========================================================= -->
  <!-- FIXED REGIME: CONTRACTIONARY POLICY -->
  <!-- ========================================================= -->

  <line x1="320" y1="110" x2="235" y2="110"
        stroke="#B66A50"
        stroke-width="2.5"/>

  <polygon points="225,110 242,102 242,118"
           fill="#B66A50"/>

  <text x="275" y="88"
        text-anchor="middle"
        font-size="18"
        font-weight="600"
        fill="#B66A50">
    Contractionary policy
  </text>

</svg>

</div>
```

::: incremental
-  Under the **flexible regime:** the **exchange rate adjusts** ($\$1.60$ to $\$1.80/£$).  
-  Under the **fixed regime:** the exchange rate cannot adjust, so **policy must adjust** (D to $D^*$).
:::

::: {.notes}
We are looking at the market for British pounds from the perspective of the United States.

Americans demand pounds when they want to buy British goods or assets. British residents supply pounds when they want to buy U.S. goods or assets.

Start at $1.60 per pound. At this exchange rate, the quantity of pounds demanded is greater than the quantity supplied. In this simplified example, this reflects a U.S. trade deficit.

So the question is: how can this imbalance disappear?

Under a flexible regime, the price is allowed to adjust. Excess demand for pounds causes the pound to appreciate — equivalently, the dollar depreciates — from $1.60 to $1.80 per pound. At the new exchange rate, demand equals supply.

Under a fixed regime, the exchange rate is not allowed to move. Something else must adjust. The graph illustrates contractionary policy: lower U.S. expenditure reduces imports and therefore reduces the demand for pounds. Demand shifts from D to D*, restoring equilibrium at $1.60.

The key point is that fixing the exchange rate does not eliminate adjustment. It changes **what has to adjust**.
:::


## 4 · The Mundell–Fleming model: bridging the real economy to the fx markets

But if governments change fiscal or monetary policy to influence the exchange rate, what happens to the real economy?

:::: columns
::: {.column width="52%"}
A **small open economy with perfect capital mobility**, so $i = i^{*}$.

IS — goods market, for a given $i^{*}$:

$$Y = C(Y-T) + I(i^{*}) + G + NX(e)$$

LM — money market, for a given $i^{*}$:

$$M/P = L(i^{*}, Y)$$
:::

::: {.column width="48%"}

```{=html}
<div style="width:100%; display:flex; justify-content:center;">

<svg viewBox="0 0 600 470"
     style="width:100%; max-width:600px; height:auto;">

  <!-- Axes -->
  <line x1="80" y1="390" x2="555" y2="390"
        stroke="#333" stroke-width="2.5"/>

  <line x1="80" y1="390" x2="80" y2="35"
        stroke="#333" stroke-width="2.5"/>

  <!-- Axis labels -->
  <text x="320" y="445"
        text-anchor="middle"
        font-size="25"
        font-style="italic"
        fill="#333">
    Output, Y
  </text>

  <text x="25" y="210"
        text-anchor="middle"
        font-size="25"
        font-style="italic"
        fill="#333"
        transform="rotate(-90 25 210)">
    Exchange rate, e
  </text>

  <!-- IS curve -->
  <path d="M 125 80
           C 220 135, 330 220, 505 345"
        fill="none"
        stroke="#5C6670"
        stroke-width="4"/>

  <text x="505" y="330"
        font-size="25"
        font-weight="700"
        font-style="italic"
        fill="#5C6670">
    IS
  </text>

  <!-- LM curve -->
  <line x1="335" y1="55"
        x2="335" y2="390"
        stroke="#176B68"
        stroke-width="4"/>

  <text x="350" y="75"
        font-size="25"
        font-weight="700"
        font-style="italic"
        fill="#176B68">
    LM
  </text>

  <!-- Equilibrium point -->
  <circle cx="335" cy="225"
          r="7"
          fill="#333"/>

  <!-- Equilibrium exchange rate guide -->
  <line x1="80" y1="225"
        x2="335" y2="225"
        stroke="#999"
        stroke-width="1.5"
        stroke-dasharray="7 6"/>

  <!-- Equilibrium output guide -->
  <line x1="335" y1="225"
        x2="335" y2="390"
        stroke="#999"
        stroke-width="1.5"
        stroke-dasharray="7 6"/>

  <!-- e* -->
  <text x="66" y="232"
        text-anchor="end"
        font-size="22"
        font-style="italic"
        fill="#555">
    e*
  </text>

  <!-- Y* -->
  <text x="335" y="417"
        text-anchor="middle"
        font-size="22"
        font-style="italic"
        fill="#555">
    Y*
  </text>

</svg>

</div>
```


:::
::::

## 4 · Problem 1.3 — policies in the Mundell–Fleming model

::: ask
What is the consequence of a **fiscal expansion**, a **monetary expansion** and a **trade restriction** on the exchange rate and on output — under floating and under fixed rates?
:::

```{=html}
<div style="
  width:80%;
  display:flex;
  gap:45px;
  justify-content:center;
  align-items:flex-start;
">

<!-- ====================================================== -->
<!-- FLOATING EXCHANGE RATE                                 -->
<!-- ====================================================== -->

<div style="width:48%; text-align:center;">

<div style="
  font-size:1.15em;
  font-weight:700;
  margin-bottom:4px;
">
Floating exchange rate
</div>

<svg viewBox="0 0 600 440"
     style="width:100%; height:auto;">

  <!-- Axes -->
  <line x1="80" y1="370" x2="555" y2="370"
        stroke="#333" stroke-width="2.5"/>

  <line x1="80" y1="370" x2="80" y2="35"
        stroke="#333" stroke-width="2.5"/>

  <!-- Axis labels -->
  <text x="320" y="425"
        text-anchor="middle"
        font-size="25"
        font-style="italic"
        fill="#333">
    Output, Y
  </text>

  <text x="27" y="200"
        text-anchor="middle"
        font-size="25"
        font-style="italic"
        fill="#333"
        transform="rotate(-90 27 200)">
    Exchange rate, e
  </text>

  <!-- IS -->
  <path d="M 125 70
           C 215 120, 330 205, 510 330"
        fill="none"
        stroke="#5C6670"
        stroke-width="4"/>

  <text x="505" y="315"
        font-size="25"
        font-weight="700"
        font-style="italic"
        fill="#5C6670">
    IS
  </text>

  <!-- LM -->
  <line x1="335" y1="50"
        x2="335" y2="370"
        stroke="#176B68"
        stroke-width="4"/>

  <text x="350" y="72"
        font-size="25"
        font-weight="700"
        font-style="italic"
        fill="#176B68">
    LM
  </text>

  <!-- Initial equilibrium -->
  <circle cx="335" cy="210"
          r="6"
          fill="#333"/>

  <!-- e0 -->
  <line x1="80" y1="210"
        x2="335" y2="210"
        stroke="#999"
        stroke-width="1.4"
        stroke-dasharray="7 6"/>

  <text x="65" y="217"
        text-anchor="end"
        font-size="21"
        font-style="italic"
        fill="#555">
    e₀
  </text>

  <!-- Y0 -->
  <line x1="335" y1="210"
        x2="335" y2="370"
        stroke="#999"
        stroke-width="1.4"
        stroke-dasharray="7 6"/>

  <text x="335" y="398"
        text-anchor="middle"
        font-size="21"
        font-style="italic"
        fill="#555">
    Y₀
  </text>

</svg>
</div>


<!-- ====================================================== -->
<!-- FIXED EXCHANGE RATE                                    -->
<!-- ====================================================== -->

<div style="width:48%; text-align:center;">

<div style="
  font-size:1.15em;
  font-weight:700;
  margin-bottom:4px;
">
Fixed exchange rate
</div>

<svg viewBox="0 0 600 440"
     style="width:100%; height:auto;">

  <!-- Axes -->
  <line x1="80" y1="370" x2="555" y2="370"
        stroke="#333" stroke-width="2.5"/>

  <line x1="80" y1="370" x2="80" y2="35"
        stroke="#333" stroke-width="2.5"/>

  <!-- Axis labels -->
  <text x="320" y="425"
        text-anchor="middle"
        font-size="25"
        font-style="italic"
        fill="#333">
    Output, Y
  </text>

  <text x="27" y="200"
        text-anchor="middle"
        font-size="25"
        font-style="italic"
        fill="#333"
        transform="rotate(-90 27 200)">
    Exchange rate, e
  </text>

  <!-- IS -->
  <path d="M 125 70
           C 215 120, 330 205, 510 330"
        fill="none"
        stroke="#5C6670"
        stroke-width="4"/>

  <text x="505" y="315"
        font-size="25"
        font-weight="700"
        font-style="italic"
        fill="#5C6670">
    IS
  </text>

  <!-- LM -->
  <line x1="335" y1="50"
        x2="335" y2="370"
        stroke="#176B68"
        stroke-width="4"/>

  <text x="350" y="72"
        font-size="25"
        font-weight="700"
        font-style="italic"
        fill="#176B68">
    LM
  </text>

  <!-- Initial equilibrium -->
  <circle cx="335" cy="210"
          r="6"
          fill="#333"/>

  <!-- Fixed exchange rate -->
  <line x1="80" y1="210"
        x2="540" y2="210"
        stroke="#999"
        stroke-width="1.6"
        stroke-dasharray="7 6"/>

  <text x="65" y="217"
        text-anchor="end"
        font-size="21"
        font-style="italic"
        fill="#555">
    ē
  </text>

  <!-- Y0 -->
  <line x1="335" y1="210"
        x2="335" y2="370"
        stroke="#999"
        stroke-width="1.4"
        stroke-dasharray="7 6"/>

  <text x="335" y="398"
        text-anchor="middle"
        font-size="21"
        font-style="italic"
        fill="#555">
    Y₀
  </text>

</svg>
</div>

</div>
```

Mundell–Fleming gives us the macroeconomic result: fiscal policy, monetary policy, or trade policy can change output and create appreciation or depreciation pressure


::: {.notes}
Floating: fiscal expansion appreciates e, Y unchanged; monetary expansion
depreciates e, raises Y; trade restriction appreciates e, Y unchanged.
Fixed: fiscal expansion raises Y, e fixed; monetary expansion is powerless
(reserves drain, M endogenous); trade restriction raises Y.
The whole point is the reversal of fiscal and monetary potency.

**Floating exchange rate**

Fiscal expansion shifts IS to the right. With LM vertical, output cannot increase. The exchange rate appreciates until net exports fall enough to offset the fiscal expansion. Therefore e rises and Y is unchanged.

Monetary expansion shifts LM to the right. Output rises and the exchange rate depreciates.

An import restriction initially increases net exports and shifts IS to the right. But the resulting appreciation offsets the increase in net exports. Output is unchanged.

**Fixed exchange rate**

Fiscal expansion shifts IS to the right. To prevent the exchange rate from appreciating, the central bank expands the money supply. LM shifts right and output rises.

A monetary expansion cannot be sustained independently. Pressure for depreciation forces the central bank to intervene, reversing the increase in the money supply. The exchange rate and output are unchanged.

An import restriction shifts IS to the right. To prevent appreciation, the central bank expands the money supply. LM shifts right and output rises.

The central result is the reversal in policy effectiveness:

**Floating rates → monetary policy is powerful; fiscal policy does not change output.**

**Fixed rates → fiscal policy is powerful; independent monetary policy is ineffective.**




:::


## 4 · Currency supply and demand

The Mundell–Fleming model tells us **whether the currency appreciates or depreciates**

::: ask
Who is **supplying** the domestic currency, and who is **demanding** it?
:::

<style>
#currency-flows table {
  font-size: 1.0em !important;
  width: 70%;
}

#currency-flows th {
  font-size: 1.0em !important;
  font-weight: 300;
}

#currency-flows th,
#currency-flows td {
  padding: 0.5em 0.6em !important;
  line-height: 1;
  vertical-align: middle;
}

#currency-flows td:first-child {
  font-weight: 300;
  white-space: nowrap;
}
</style>

<div id="currency-flows">

| | **Supply of domestic currency** | **Demand for domestic currency** |
|:---|:---|:---|
| **Trade** | Imports of foreign goods and services | Exports of domestic goods and services |
| **Income & transfers** | Payments to non-residents; transfers abroad | Income received from abroad; transfers received |
| **Financial flows** | Residents purchase foreign assets → **capital outflow** | Non-residents purchase domestic assets → **capital inflow** |
| **Debt flows** | Repayment of foreign debt | Foreign borrowing / credit received |
| **Central bank** | Buys foreign currency → reserves **increase** | Sells foreign currency → reserves **decrease** |

</div>

::: {.notes}
In Mundell–Fleming, we saw the exchange rate appreciate or depreciate in response to different policies.

But e does not simply move by itself. These movements come from transactions in the foreign-exchange market.

So now let's look behind the Mundell–Fleming diagram.

Who supplies the domestic currency?

Domestic residents supply it when they need foreign currency — for example, to import foreign goods, invest abroad, or make payments abroad.

Who demands the domestic currency?

Foreign residents demand it when they want to buy our goods, invest in our assets, or make payments to domestic residents.

And under a fixed exchange rate, there is one additional actor: the central bank.

If it wants to resist depreciation, it sells foreign reserves and buys domestic currency. If it wants to resist appreciation, it buys foreign currency and supplies domestic currency.

This gives us the market mechanism behind the movements we just saw in Mundell–Fleming.
:::


## 4 · Central bank intervention

::: ask
Suppose the ECB wants to keep the euro at a **fixed exchange rate $\bar e$**.

What must it do in each situation?
:::

```{=html}
<div style="
  width:80%;
  display:flex;
  gap:55px;
  justify-content:center;
  align-items:flex-start;
">

<!-- ====================================================== -->
<!-- DEMAND INCREASES                                       -->
<!-- ====================================================== -->

<div style="width:47%; text-align:center;">

<div style="font-size:1.12em; font-weight:700; margin-bottom:5px;">
Demand for euros increases
</div>

<svg viewBox="0 0 600 430"
     style="width:100%; height:auto;">

  <!-- Axes -->
  <line x1="80" y1="365" x2="555" y2="365"
        stroke="#333" stroke-width="2.5"/>
  <line x1="80" y1="365" x2="80" y2="35"
        stroke="#333" stroke-width="2.5"/>

  <!-- Axis labels -->
  <text x="320" y="415"
        text-anchor="middle"
        font-size="23"
        fill="#333">
    Quantity of euros
  </text>

  <text x="25" y="200"
        text-anchor="middle"
        font-size="23"
        fill="#333"
        transform="rotate(-90 25 200)">
    Value of the euro
  </text>

  <!-- Supply -->
  <path d="M 135 330 L 500 70"
        fill="none"
        stroke="#176B68"
        stroke-width="4"/>

  <text x="490" y="62"
        font-size="23"
        font-weight="700"
        fill="#176B68">
    S
  </text>

  <!-- Initial demand -->
  <path d="M 130 70 L 500 330"
        fill="none"
        stroke="#5C6670"
        stroke-width="4"/>

  <text x="490" y="345"
        font-size="22"
        font-weight="700"
        fill="#5C6670">
    D₁
  </text>

  <!-- New demand -->
  <path d="M 195 70 L 555 320"
        fill="none"
        stroke="#B66A50"
        stroke-width="4"
        stroke-dasharray="10 7"/>

  <text x="535" y="335"
        font-size="22"
        font-weight="700"
        fill="#B66A50">
    D₂
  </text>

  <!-- Fixed exchange rate -->
  <line x1="80" y1="200" x2="550" y2="200"
        stroke="#777"
        stroke-width="2"
        stroke-dasharray="7 6"/>

  <text x="65" y="207"
        text-anchor="end"
        font-size="22"
        font-style="italic"
        fill="#444">
    ē
  </text>

  <!-- Initial equilibrium -->
  <circle cx="315" cy="200" r="6" fill="#333"/>

</svg>

</div>


<!-- ====================================================== -->
<!-- SUPPLY INCREASES                                       -->
<!-- ====================================================== -->

<div style="width:47%; text-align:center;">

<div style="font-size:1.12em; font-weight:700; margin-bottom:5px;">
Supply of euros increases
</div>

<svg viewBox="0 0 600 430"
     style="width:100%; height:auto;">

  <!-- Axes -->
  <line x1="80" y1="365" x2="555" y2="365"
        stroke="#333" stroke-width="2.5"/>
  <line x1="80" y1="365" x2="80" y2="35"
        stroke="#333" stroke-width="2.5"/>

  <!-- Axis labels -->
  <text x="320" y="415"
        text-anchor="middle"
        font-size="23"
        fill="#333">
    Quantity of euros
  </text>

  <text x="25" y="200"
        text-anchor="middle"
        font-size="23"
        fill="#333"
        transform="rotate(-90 25 200)">
    Value of the euro
  </text>

  <!-- Initial supply -->
  <path d="M 135 330 L 500 70"
        fill="none"
        stroke="#176B68"
        stroke-width="4"/>

  <text x="490" y="62"
        font-size="22"
        font-weight="700"
        fill="#176B68">
    S₁
  </text>

  <!-- New supply -->
  <path d="M 200 330 L 555 80"
        fill="none"
        stroke="#B66A50"
        stroke-width="4"
        stroke-dasharray="10 7"/>

  <text x="535" y="72"
        font-size="22"
        font-weight="700"
        fill="#B66A50">
    S₂
  </text>

  <!-- Demand -->
  <path d="M 130 70 L 500 330"
        fill="none"
        stroke="#5C6670"
        stroke-width="4"/>

  <text x="490" y="345"
        font-size="23"
        font-weight="700"
        fill="#5C6670">
    D
  </text>

  <!-- Fixed exchange rate -->
  <line x1="80" y1="200" x2="550" y2="200"
        stroke="#777"
        stroke-width="2"
        stroke-dasharray="7 6"/>

  <text x="65" y="207"
        text-anchor="end"
        font-size="22"
        font-style="italic"
        fill="#444">
    ē
  </text>

  <!-- Initial equilibrium -->
  <circle cx="315" cy="200" r="6" fill="#333"/>

</svg>

</div>

</div>
```

::: {.notes}
Start on the left.

Demand for euros increases. Without intervention, the euro would appreciate.

But suppose the ECB wants to maintain the exchange rate at e-bar.

At the fixed exchange rate there is now excess demand for euros.

What can the ECB do?

It must **supply euros to the foreign-exchange market**. In exchange, it buys foreign currency.

So:

**ECB buys foreign currency → supplies euros → foreign reserves increase.**

Now look at the right-hand graph.

The supply of euros increases. Without intervention, the euro would depreciate.

At e-bar there is now excess supply of euros.

To maintain the fixed exchange rate, the ECB must **demand euros**. It does this by selling foreign currency from its reserves and receiving euros in exchange.

So:

**ECB sells foreign currency → buys euros → foreign reserves decrease.**

This is exactly the central-bank row in the previous table.

And it also explains the Mundell–Fleming result: under a fixed exchange rate, intervention changes the domestic money supply. Monetary policy therefore cannot be conducted independently of the exchange-rate commitment.
:::

## 4 · Foreign-exchange crisis: Thailand, 1997

::: incremental
- Before 1997, Thailand maintained the baht within a **very narrow range**, effectively limiting its movement against the U.S. dollar.
- Financial-sector weaknesses, short-term foreign debt and external imbalances increasingly undermined confidence in the baht.
- Investors began expecting a **devaluation** and selling baht → **supply of baht ↑ and demand for baht ↓**.
- To resist depreciation, the Bank of Thailand **bought baht and sold foreign currency**, progressively depleting its usable reserves.
- On **2 July 1997**, Thailand abandoned the peg and moved to a managed float. The baht depreciated sharply.
:::

::: ask
Using our previous FX-market graph: **how could the central bank defend the baht — and why couldn't it do so indefinitely?**
:::

::: {.notes}
This is the mechanism from the previous slide playing out in an actual crisis.

Thailand had maintained a tightly managed exchange rate. But by late 1996 and 1997, concerns about the financial system, external imbalances and short-term foreign debt increasingly undermined confidence in the baht.

Investors started expecting that the existing exchange rate could not be maintained.

What happens in our foreign-exchange market?

People try to sell baht and acquire foreign currencies. The supply of baht increases, while demand weakens. This creates depreciation pressure.

But the central bank is trying to maintain the exchange rate.

So what does it do?

Exactly what we saw on the previous slide: it sells foreign currency and buys baht.

The problem is that its ability to do this is limited by its foreign-exchange resources.

As speculative pressure intensified, the Bank of Thailand's usable reserves were progressively depleted. On 2 July 1997, the authorities abandoned the peg and introduced a managed float.

And this is where the exchange-rate crisis became especially damaging: substantial foreign-currency liabilities meant that depreciation increased debt burdens measured in baht.
:::

## 4 · The Asian crisis: the mechanism

:::: columns
::: {.column width="45%"}

Suppose the baht is fixed at **$0.04**, while market pressure pushes its equilibrium value toward **$0.02**.

At the fixed rate:

$$S_{\text{baht}} > D_{\text{baht}}$$

::: incremental
- The central bank must absorb the **excess supply of baht**.
- It **buys baht and sells dollars** → reserves $\downarrow$.
- Expectations of devaluation can trigger further baht sales → even greater pressure on reserves.
- Once the defense becomes unsustainable: **devalue or float**.
:::

:::

::: {.column width="55%"}
```{=html}
<div style="width:100%; display:flex; justify-content:center;">

<svg viewBox="0 0 760 520"
     style="width:100%; max-width:760px; height:auto;">

  <!-- ================= AXES ================= -->

  <line x1="110" y1="440" x2="710" y2="440"
        stroke="#333" stroke-width="2.5"/>

  <line x1="110" y1="440" x2="110" y2="45"
        stroke="#333" stroke-width="2.5"/>

  <!-- arrowheads -->
  <polygon points="710,440 697,433 697,447" fill="#333"/>
  <polygon points="110,45 103,58 117,58" fill="#333"/>

  <!-- axis labels -->
  <text x="35" y="52"
        text-anchor="middle"
        font-size="21"
        fill="#333">
    <tspan x="35" dy="0">Dollars</tspan>
    <tspan x="35" dy="23">per baht</tspan>
  </text>

  <text x="650" y="468"
        text-anchor="middle"
        font-size="21"
        fill="#333">
    <tspan x="650" dy="0">Millions of</tspan>
    <tspan x="650" dy="23">baht per month</tspan>
  </text>


  <!-- ================= PRICE LINES ================= -->

  <!-- $0.04 -->
  <line x1="110" y1="180" x2="440" y2="180"
        stroke="#777"
        stroke-width="1.4"
        stroke-dasharray="5 5"/>

  <text x="90" y="187"
        text-anchor="end"
        font-size="20"
        fill="#333">
    $0.04
  </text>

  <!-- $0.02 -->
  <line x1="110" y1="315" x2="440" y2="315"
        stroke="#777"
        stroke-width="1.4"
        stroke-dasharray="5 5"/>

  <text x="90" y="322"
        text-anchor="end"
        font-size="20"
        fill="#333">
    $0.02
  </text>


  <!-- ================= QUANTITY GUIDES ================= -->

  <!-- 100 -->
  <line x1="190" y1="180" x2="190" y2="440"
        stroke="#777"
        stroke-width="1.4"
        stroke-dasharray="5 5"/>

  <text x="190" y="466"
        text-anchor="middle"
        font-size="20"
        fill="#333">
    100
  </text>

  <!-- 400 -->
  <line x1="440" y1="180" x2="440" y2="440"
        stroke="#777"
        stroke-width="1.4"
        stroke-dasharray="5 5"/>

  <text x="440" y="466"
        text-anchor="middle"
        font-size="20"
        fill="#333">
    400
  </text>


  <!-- ================= ORIGINAL CURVES ================= -->

  <!-- S1 -->
  <!-- passes through A -->
  <line x1="150" y1="390"
        x2="390" y2="75"
        stroke="#176B68"
        stroke-width="3.5"/>

  <text x="392" y="68"
        font-size="22"
        font-weight="700"
        fill="#176B68">
    S₁
  </text>

  <!-- D1 -->
  <!-- passes through A -->
  <line x1="245" y1="75"
        x2="540" y2="405"
        stroke="#5C6670"
        stroke-width="3.5"/>

  <text x="530" y="420"
        font-size="22"
        font-weight="700"
        fill="#5C6670">
    D₁
  </text>


  <!-- ================= SHIFTED CURVES ================= -->

  <!-- D2 -->
  <!-- intersects $0.04 at Q = 100 and crosses S2 at B -->
  <line x1="120" y1="90"
        x2="395" y2="405"
        stroke="#B66A50"
        stroke-width="3.5"/>

  <text x="390" y="420"
        font-size="22"
        font-weight="700"
        fill="#B66A50">
    D₂
  </text>

  <!-- S2 -->
  <!-- intersects $0.04 at Q = 400 and crosses D2 at B -->
  <line x1="255" y1="390"
        x2="535" y2="70"
        stroke="#3579D8"
        stroke-width="3.5"/>

  <text x="535" y="64"
        font-size="22"
        font-weight="700"
        fill="#3579D8">
    S₂
  </text>


  <!-- ================= POINT A ================= -->

  <!-- initial equilibrium: S1 x D1 -->
  <circle cx="315" cy="180" r="6" fill="#333"/>

  <text x="327" y="169"
        font-size="22"
        font-weight="700"
        fill="#333">
    A
  </text>


  <!-- ================= POINT B ================= -->

  <!-- new equilibrium: S2 x D2 -->
  <circle cx="315" cy="315" r="6" fill="#333"/>

  <text x="327" y="337"
        font-size="22"
        font-weight="700"
        fill="#333">
    B
  </text>


  <!-- ================= POINT C ================= -->

  <!-- S2 at the defended $0.04 -->
  <circle cx="440" cy="180" r="6" fill="#333"/>

  <text x="452" y="197"
        font-size="22"
        font-weight="700"
        fill="#333">
    C
  </text>


  <!-- ================= CENTRAL BANK PURCHASE ================= -->

  <line x1="190" y1="145"
        x2="440" y2="145"
        stroke="#B66A50"
        stroke-width="3"/>

  <!-- left arrow -->
  <polygon points="190,145 204,137 204,153"
           fill="#B66A50"/>

  <!-- right arrow -->
  <polygon points="440,145 426,137 426,153"
           fill="#B66A50"/>

  <text x="315" y="112"
        text-anchor="middle"
        font-size="19"
        font-weight="700"
        fill="#333">
    <tspan x="315">Thailand CB</tspan>
    <tspan x="315" dy="22">buys 300 million baht</tspan>
  </text>

</svg>

</div>
```
:::
::::

::: {.notes}
This is a stylized representation of the mechanism, not the literal quantities observed in Thailand.

Suppose the central bank maintains the baht at four cents, but after a loss of confidence the market-clearing value would be closer to two cents.

At four cents, there is now excess supply of baht.

To maintain the fixed rate, the central bank has to buy that excess baht. What does it use to buy it? Its foreign-currency resources.

So it sells dollars and buys baht.

Now introduce expectations.

If investors believe the central bank will eventually be unable to maintain the rate, holding baht becomes less attractive. They have an incentive to sell before the devaluation occurs.

That creates additional selling pressure, requiring still more intervention.

This is why credibility and reserves matter under a fixed exchange-rate regime.

In Thailand, speculative attacks and intervention substantially depleted the Bank of Thailand's usable reserves. The authorities eventually abandoned the peg on 2 July 1997.

The subsequent depreciation was particularly damaging because firms and financial institutions had substantial foreign-currency exposures. When the baht depreciated, the domestic-currency burden of those liabilities increased sharply.
:::

. . .

::: keyidea
A fixed exchange rate can become a **self-reinforcing reserve problem** when expectations of devaluation generate additional capital outflows.
:::


# 5 · Climate change and the financial system {.divider .center background-color="#06302F"}

## 5 · Monasterolo (2020) — Discussion

10 groups of 6–7 students · 2 groups per question

### 1 · Prepare — 45 min

- **One question** per group
- Read the paper and prepare a **collective answer**
- Main **arguments, mechanisms, and examples** - also challenge the author's ideas
- Choose a **lead speaker**, but everyone should be ready to contribute

### 2 · Discuss — ~5 min per question

- **Group A · 2 min** — Present your answer
- **Group B · 2 min** — React and add: **complement, qualify, or challenge**
- **Both groups · 1 min** — Respond to each other; anyone can jump in


## 5 · Monasterolo (2020), *Climate Change and the Financial System*


1. Why might **investors' decisions be misaligned with climate risks**?

2. What are the main **climate-related financial risks**, and how can they affect the financial system?

3. How can **climate policy and regulation** create risks for financial assets?

4. Why is **assessing and pricing climate risk** particularly challenging?

5. What would **good practices** for investors, financial institutions, and regulators look like?



## Next class {.hook background-color="#FAF7F2"}

Next class, we will start our session on **securities markets**.

::: ask
### Coming next

- What are **securities markets**?
- How do **money markets** and **capital markets** differ?
- Who participates in these markets — and why?
- How are securities **issued, traded, and priced**?
:::

