---
title: "International Finance"
subtitle: "Session 3 — Securities Markets: Money, Bonds and Stocks"
author: "Laura Contreras-Portela"
institute: "KEDGE Business School"
date: today
date-format: "MMMM YYYY"
format:
  revealjs:
    theme: [default, ../_theme/theme.scss]
    width: 1600
    height: 900
    title-slide-attributes:
      data-background-color: "#FAF7F2"
    slide-number: c/t
    show-slide-number: all
    progress: true
    history: true
    hash: true
    controls: true
    navigation-mode: linear      # right arrow always goes to the next slide
    preview-links: auto
    highlight-style: github
    code-line-numbers: true
    html-math-method: katex
    chalkboard:                     # C = draw on the slide, B = blank board
      buttons: true
      theme: whiteboard
      chalk-effect: 0
    footer: "International Finance · Session 3 · KEDGE"
    embed-resources: false          # true -> one portable file, no wifi needed
  beamer:                           # same source -> a real LaTeX PDF handout
    aspectratio: 169
    theme: metropolis
    incremental: false
    header-includes: |
      \definecolor{ifteal}{HTML}{0E5250}
      \definecolor{ifclay}{HTML}{B4573C}
      \setbeamercolor{frametitle}{bg=ifteal}
      \setbeamercolor{progress bar}{fg=ifteal}
      \setbeamercolor{title separator}{fg=ifteal}
      \setbeamercolor{structure}{fg=ifteal}
      \setbeamercolor{alerted text}{fg=ifclay}
execute:
  echo: false
  warning: false
  message: false
  fig-align: center
  fig-width: 8
  fig-height: 4.2
knitr:
  opts_chunk:
    dev: png
    dpi: 300
---

```{r setup}
source("../_R/setup.R")
```

## A small open economy, in two equations {.hook background-color="#FAF7F2"}

:::: columns
::: {.column width="52%"}
Every market we open today sits inside one framework. A **small open economy**
with **perfect capital mobility**, so the domestic rate is pinned to the world
rate: $i = i^{*}$.

**IS** — the goods market, for a given $i^{*}$:

$$Y = C(Y-T) + I(i^{*}) + G + NX(e)$$

**LM** — the money market, for a given $i^{*}$:

$$M/P = L(i^{*}, Y)$$

::: ask
The LM curve comes out **vertical** in $(Y, e)$ space. **Why?**
:::
:::

::: {.column width="48%"}
```{r}
#| fig-width: 5.6
#| fig-height: 4.2
yy <- seq(0, 10, length.out = 100)
ggplot() +
  geom_line(data = data.frame(y = yy, e = 8.4 - .62 * yy),
            aes(y, e), colour = IF_COL$slate, linewidth = 1.1) +
  annotate("segment", x = 5.6, xend = 5.6, y = 0, yend = 9,
           colour = IF_COL$teal700, linewidth = 1.1) +
  annotate("segment", x = 0, xend = 5.6, y = 4.93, yend = 4.93,
           colour = IF_COL$muted, linetype = "22", linewidth = .4) +
  annotate("text", x = 6.1, y = 8.6, label = "LM", colour = IF_COL$teal700,
           size = 4.6, fontface = "italic") +
  annotate("text", x = 8.4, y = 3.8, label = "IS", colour = IF_COL$slate,
           size = 4.6, fontface = "italic") +
  labs(x = expression(Y), y = expression(e)) +
  coord_cartesian(xlim = c(0, 10), ylim = c(0, 9)) +
  theme_schematic()
```

$e$ = foreign currency per unit of domestic currency.
:::
::::

::: notes
Ten minutes on these three slides. Do not re-derive the model - put it up as the
machinery we are going to use, and make them operate it.

Open with the claim, so they know why it is on the board:

Whether a government can do anything at all with monetary policy depends on one
choice: fixed or floating. That is what this model tells you, and it is the
reason half of today's instruments exist.

Then put the question up and wait for an answer before you say anything.

*[wait]*

What you are listening for: the interest rate is fixed at the world rate. Look
at the LM equation - if $i^{*}$ is given and $M/P$ is given, then there is only
one level of $Y$ that clears the money market. The exchange rate does not appear
in it at all. So the curve is vertical.

Then say the consequence out loud, because everything on the next slide follows
from it:

If LM is vertical, output is determined by the money market alone. Anything that
shifts IS - government spending, taxes, a tariff - moves the exchange rate and
leaves output exactly where it was.
:::

## What the model predicts {.center}

::: ask
Fill it in. A **fiscal expansion**, a **monetary expansion**, an **import
restriction** — what does each do to the exchange rate and to output, under
**floating** and under **fixed** rates?
:::

::: {.fillin}
| Policy | Floating: $e$ | Floating: $Y$ | Fixed: $e$ | Fixed: $Y$ |
|:---|:--:|:--:|:--:|:--:|
| Fiscal expansion | | | | |
| Monetary expansion | | | | |
| Import restriction | | | | |
:::

::: {.smaller}
Press **`C`** and draw the IS and LM shift for each row before writing the sign.
:::

::: notes
Do not fill this in for them. Draw it.

*[press C for the chalkboard]*

Take the rows one at a time. For each one ask: which curve moves, and in which
direction? Then read the answer off the diagram.

The answers, so you have them:

Floating. Fiscal expansion: IS shifts right, e appreciates, Y unchanged.
Monetary expansion: LM shifts right, e depreciates, Y rises. Import
restriction: IS shifts right, e appreciates, Y unchanged.

Fixed. Fiscal expansion: Y rises, e fixed by construction. Monetary expansion:
powerless - the central bank has to sell reserves to hold the rate, which pulls
the money supply straight back in. Import restriction: Y rises.

*[pause]*

The sentence to land, and ask them to write it down:

Under floating rates monetary policy works and fiscal policy does not. Under
fixed rates it is exactly the other way round. The exchange rate regime decides
which instrument your government actually has.
:::

## The fixed-rate row, in one country {.center}

:::: columns
::: {.column width="45%"}
Thailand, 1997. The baht was **fixed at $0.04** while the equilibrium rate had
fallen to **$0.02**.

::: incremental
-   To hold the rate the central bank must **buy 300 million baht a month**,
    paying in dollars from its reserves.
-   That is the fixed-rate row of the table, happening: every intervention
    contracts the domestic money supply. **Monetary policy is not available.**
-   When the reserves run out the choice is to **float** or to **devalue**. In
    July 1997 Thailand floated, and the baht halved overnight.
:::
:::

::: {.column width="55%"}
```{r}
#| fig-width: 6.2
#| fig-height: 4.4
q <- seq(50, 500, length.out = 200)
bt <- rbind(
  data.frame(q = q, p = 0.010 + 0.00008 * q, curve = "S1 supply"),
  data.frame(q = q, p = 0.001 + 0.00008 * q, curve = "S2 shifted"),
  data.frame(q = q, p = 0.062 - 0.00008 * q, curve = "D1 demand"),
  data.frame(q = q, p = 0.050 - 0.00008 * q, curve = "D2 shifted")
)
ggplot(bt, aes(q, p, colour = curve, linetype = curve)) +
  geom_line(linewidth = .95) +
  geom_hline(yintercept = c(0.02, 0.04), colour = IF_COL$muted,
             linetype = "22", linewidth = .4) +
  annotate("segment", x = 100, xend = 400, y = 0.0415, yend = 0.0415,
           colour = IF_COL$clay, linewidth = 1,
           arrow = grid::arrow(length = unit(.018, "npc"), ends = "both", type = "closed")) +
  annotate("text", x = 250, y = 0.0465, label = "central bank buys 300",
           colour = IF_COL$clay, size = 3.5, fontface = "bold") +
  scale_colour_manual(values = c(IF_COL$slate, IF_COL$clay,
                                 IF_COL$teal700, IF_COL$teal300)) +
  scale_linetype_manual(values = c(1, 5, 1, 5)) +
  scale_x_continuous(breaks = c(100, 400)) +
  scale_y_continuous(breaks = c(0.02, 0.04), labels = c("$0.02", "$0.04")) +
  labs(x = "Millions of baht per month", y = "Dollars per baht") +
  theme_if() +
  theme(panel.grid.major.x = element_blank(),
        legend.text = element_text(size = rel(.7)))
```
:::
::::

::: notes
They have not studied this crisis, so introduce it rather than test it. Two or
three minutes, and the purpose is to make the fixed-rate column concrete.

Set it up in a sentence: Thailand promised to keep the baht at four cents. The
market thought it was worth two. The promise costs money to keep.

*[point at the gap between the two dashed lines]*

That distance is what the central bank is buying, every month, out of a finite
pile of dollars.

*[reveal the bullets one at a time]*

On the second bullet, connect it back explicitly:

This is the row we just filled in. Every time the central bank buys baht it is
taking baht out of circulation - the money supply contracts whether it wants
that or not. That is what "monetary policy is powerless under a fixed rate"
looks like from the inside.

Then the bridge into today, and do not skip it:

Notice what actually broke. Thai banks had borrowed in dollars, short, and lent
in baht, long. Halve the baht and the debts double while the assets do not move.
That mismatch - borrow short in one form, lend long in another - is the subject
of today's first market. Same structure, different costume.
:::

## Where we are

Sessions 1 and 2 established four prices and the balance-of-payments identity.
We now disaggregate the financial system by **instrument**, and ask how each
market transfers funds from surplus units to deficit units.

:::: columns
::: {.column width="50%"}
1. **Money markets** — short maturity, high liquidity
2. **Bond market** — long-term debt claims
3. **Stock market** — residual ownership claims
:::

::: {.column width="50%"}
4. **Mortgage markets** — debt secured on real property
5. **Foreign exchange** — claims in different currencies
6. **Derivatives** — claims contingent on other claims
:::
::::

::: keyidea
Each market is analysed through the same two questions: **what contractual
claim does the instrument represent**, and **what determines its price**. These
are the organising questions of the course, and of the examination.
:::

::: {.source}
Plus the reading discussion: **Sartzetakis (2021)** on green bonds.
:::

::: notes
In the first two sessions we built four prices and one accounting identity.
Today we open the box. We stop asking what the exchange rate is, and start
asking who issues the paper, who buys it, and what makes the price move.

There are six markets on this list. We do the first three today: money markets,
bonds, and stocks. Mortgages, currencies and derivatives are next session.

For every one of them I am going to ask you the same two questions. What is the
claim? And what determines its price. Those two questions are the course. They
are also the exam, so if you find yourself memorising definitions, stop, and go
back to those two.
:::

## 9 August 2007: a funding market stops {.hook background-color="#FAF7F2"}

BNP Paribas suspended redemptions on three funds, citing an evaporation of
liquidity that made it impossible to value the assets they held. Those assets
were **asset-backed commercial paper**, and the market for them had ceased to
refinance itself.

::: incremental
-   This was not an equity market event. It was a **money market** — the
    shortest, safest, largest segment of the system — that stopped rolling over.
-   The underlying practice is **maturity transformation**: funding long-dated
    assets with short-dated liabilities. It is profitable precisely because it
    is risky, and the risk is **rollover risk**.
-   The same mechanism propagated through Northern Rock, Bear Stearns and
    Lehman Brothers within thirteen months.
:::

. . .

::: ask
Market microstructure is usually treated as detail. Under what conditions does
it become the **binding constraint** on the whole financial system?
:::

::: notes
On the ninth of August 2007, BNP Paribas told investors in three of its funds
that they could not have their money back. Not that the funds had lost value —
that the bank could not say what they were worth.

The assets were asset-backed commercial paper. Very short-term corporate
borrowing. And the market for it had simply stopped.

*[pause]*

Notice what this was not. It was not a stock market crash. It was the shortest,
safest, most boring corner of the system — and it was the one that seized.

Here is why it seized. Those vehicles were borrowing for thirty days and lending
for thirty years. We call that maturity transformation, and it is profitable
exactly because it is dangerous. Every thirty days you have to find someone
willing to lend to you again. That is rollover risk. One month nobody was
willing.

Thirteen months later the same mechanism had taken down Northern Rock, Bear
Stearns, and Lehman Brothers.

So when I say market plumbing, I do not mean detail. I mean the thing that
decides whether the rest of the system works on a given morning.
:::

# 1 · Money markets {.divider .center background-color="#06302F"}

## The corporate cash-timing problem {.hook background-color="#FAF7F2"}

A firm receives EUR 40 million on the 25th of the month and disburses it on the
5th. For eleven days those funds must earn a return without being committed:
the firm requires a **liquid**, short-dated, low-risk store of value.

::: incremental
-   Receipts and disbursements are **imperfectly synchronised** — for
    corporations, for banks, and for governments alike.
-   The money market exists to price that mismatch. It is a market for
    **liquidity over short horizons**, not for long-run investment returns.
:::

::: notes
Let me make this concrete. You work in treasury. Your company is paid forty
million euros on the twenty-fifth of the month, and it pays that money out on
the fifth of the next month.

So for eleven days you are sitting on forty million euros. You cannot leave it
in a current account earning nothing. You cannot lock it away either, because
you need it on the fifth.

*[ask]* What do you do with it?

You need something short, safe, and sellable. That is the money market. Eleven
days at three percent on forty million is about thirty-six thousand euros. That
is real money for doing nothing except not being careless.

And this is not only companies. Banks have the same problem. Governments have
the same problem. Money comes in and goes out, and the two are never
synchronised.
:::

## 1 · What makes a money market instrument

::: incremental
-   **Short maturity** — under one year, and typically under 120 days. This
    limits both interest-rate risk and the horizon over which credit quality
    can deteriorate.
-   **High liquidity** — the instrument can be sold before maturity without a
    material price concession.
-   **Low default risk** — issuers are sovereigns, large banks and highly rated
    corporations.
-   **Large denominations** — USD 1 million and above. This is a **wholesale**
    market; households participate indirectly, through money market funds.
:::

. . .

::: keyidea
The four properties are not independent. Short maturity makes low default risk
credible; low default risk and standardisation are what sustain a liquid
secondary market. Both parties are managing a **liquidity-timing** problem
rather than seeking a risk-adjusted return — which is why these instruments
price close to the risk-free curve, and why small deviations from it are
informative.
:::

::: notes
Four properties. They are examinable, and they are connected, so do not learn
them as a list.

Short maturity. Under a year, usually under a hundred and twenty days. That
matters for two reasons: rates cannot move far against you in three months, and
a company's credit cannot deteriorate far in three months either.

High liquidity. You can sell before maturity without dropping the price to find
a buyer.

Low default risk. Look at who is borrowing here: governments, the largest banks,
the best-rated companies.

Large denominations. A million dollars and up. None of you will ever buy one
of these directly. You reach this market through a money market fund.

*[pause]*

Now the connection. Short maturity is *why* the low default risk is believable.
And low default risk plus standardisation is *why* there is a liquid secondary
market. Pull one property out and the others go with it.

And notice what nobody is doing here. Nobody buys commercial paper because they
believe in the company. Both sides are solving a timing problem. That is why
these instruments price so close to the risk-free rate — and why when they move
away from it, that gap tells you something is wrong.
:::

## 1 · The instruments

::: {.panel-tabset}

### Treasury bills

-   Zero-coupon obligations of the U.S. Treasury, 28 days to 12 months.
-   Issued by **auction**. *Competitive* bids specify a price and may be
    rationed; *noncompetitive* bids accept the clearing price and are filled in
    full.
-   They constitute the **benchmark risk-free asset** of the domestic market:
    every other short rate in this session is quoted relative to them.

### Fed funds

-   Uncollateralised loans of **reserve balances** between depository
    institutions, predominantly **overnight**.
-   Banks use them to satisfy reserve requirements at the close of the
    maintenance period.
-   The **federal funds rate** is the operating target of U.S. monetary policy,
    which makes it the anchor of the short end of the yield curve.

### Repos

-   A **repurchase agreement** is the sale of a security with a simultaneous
    commitment to repurchase it at a fixed price, typically within 3 to 14 days.
-   Economically it is a **collateralised loan**; the repurchase price less the
    sale price is the interest. Legally it is a sale, which is why repo
    receives privileged treatment in bankruptcy.
-   The lender applies a **haircut** — lending less than the collateral's market
    value — as protection against a fall in that value.
-   Unlike fed funds, **non-bank institutions participate**, which is what made
    the repo market both large and, in 2008, a channel for runs.

### CDs and CP

-   **Negotiable certificates of deposit**: bank-issued time deposits,
    USD 100,000 to USD 10 million, transferable before maturity.
-   **Commercial paper**: an **unsecured** promissory note issued by a
    corporation. Maturity is capped at **270 days** because that is the limit of
    the exemption from SEC registration.
-   Being unsecured, commercial paper is a direct claim on the issuer's
    creditworthiness — the feature that matters in section 1's charts.

### BAs and eurodollars

-   A **banker's acceptance** is a time draft on which a commercial bank has
    substituted its own credit for the importer's.
-   This resolves a problem of **asymmetric information** in cross-border trade:
    the exporter is paid at once, bears no exchange-rate exposure, and need not
    evaluate a counterparty it cannot observe. The importer's bank, which can
    observe it, performs that assessment.
-   **Eurodollars** are dollar-denominated deposits held at banks outside the
    United States — a response to the dollar's role as invoicing and settlement
    currency, not to any feature of U.S. regulation alone.

:::

::: notes
I am not going to read all five tabs. You have them in the handout. I want to
spend the time on two.

Treasury bills first. Zero coupon, sold at auction, and they are the benchmark.
Every other short rate you will see today is quoted relative to them.

*[switch to the repo tab]*

Now repos, because this is the one that matters.

A repo is a pawn shop for securities. I sell you my bonds today and I promise to
buy them back on Thursday at a slightly higher price. That difference is the
interest. Legally it is a sale. Economically it is a loan with collateral.

The lender does not lend the full value of the collateral. If the bonds are
worth a hundred, they lend you ninety. That ten is called the haircut, and it
protects the lender if the collateral falls in value.

*[ask]* So what happens to me if my lender wakes up one morning and decides the
haircut should be thirty instead of ten?

That is the 2008 repo run in one question. Nobody has to refuse to lend. They
only have to ask for more collateral.

One more thing before we move on. Commercial paper is unsecured — no collateral
at all. Hold on to that word. We come back to it in two slides.
:::

## 1 · The instruments side by side

| Instrument | Issuer | Buyer | Usual maturity | Secondary market |
|:---|:---|:---|:---|:--:|
| Treasury bills | U.S. government | Consumers and companies | 4, 13, 26 weeks | Excellent |
| Federal funds | Banks | Banks | 1 to 7 days | None |
| Repurchase agreements | Businesses and banks | Businesses and banks | 1 to 15 days | Good |
| Negotiable CDs | Large money-centre banks | Businesses | 14 to 120 days | Good |
| Commercial paper | Finance companies and businesses | Businesses | 1 to 270 days | Poor |
| Banker's acceptance | Banks | Businesses | 30 to 180 days | Good |
| Eurodollar deposits | Non-US banks | Businesses, governments, banks | 1 day to 1 year | Poor |

::: {.source}
The final column is a ranking of secondary-market liquidity. Adapted from
Mishkin, F.S. and Eakins, S.G., *Financial Markets and Institutions*.
:::

::: notes
Do not write this down, it is in the handout. Take thirty seconds and read the
last column.

*[wait]*

*[ask]* Why does that column vary so much? Treasury bills are excellent,
commercial paper is poor. Same kind of instrument, same kind of maturity.

The answer is standardisation. One Treasury bill is identical to another, so
anyone will buy yours. Commercial paper is a specific claim, on one company, for
one maturity — there is no queue of buyers for that. And fed funds are unsecured
overnight loans between two named banks. There is nothing to sell.

Liquidity is not a property of the paper. It is a property of how many people
want the same thing you are holding.
:::

## 1 · The rates move together

```{r}
#| fig-width: 12
#| fig-height: 4.4
ids <- c(TB3MS = "3-month Treasury bill",
         FEDFUNDS = "Federal funds",
         MPRIME = "Bank prime rate")
series <- lapply(names(ids), function(id) {
  d <- fred(id)
  if (is.null(d)) return(NULL)
  d$series <- ids[[id]]
  d
})
series <- Filter(Negate(is.null), series)

if (!length(series)) {
  no_data_note("TB3MS / FEDFUNDS / MPRIME")
} else {
  d <- do.call(rbind, series)
  d <- d[d$date >= as.Date("2000-01-01"), ]
  ggplot(d, aes(date, value, colour = series)) +
    recession_bands(from = as.Date("2000-01-01")) +
    geom_line(linewidth = .5) +
    scale_colour_manual(values = unname(unlist(
      IF_COL[c("teal700", "clay", "slate")]))) +
    labs(x = NULL, y = "Percent",
         title = "Money market rates move as one",
         caption = "Shaded: NBER recessions. Source: Federal Reserve via FRED.") +
    theme_if()
}
```

::: ask
The series co-move closely because they share a common policy anchor. Under
what conditions does the **spread** between the federal funds rate and the
Treasury bill rate widen, and what is that spread measuring?
:::

::: notes
Three rates here: Treasury bills, federal funds, the prime rate.

*[ask]* What do you notice?

They move together. All three are anchored to the same policy rate, so most of
the time they are on top of each other.

*[pause]*

Which makes the interesting question the opposite one. What does it mean when
they come apart?

The federal funds rate is banks lending to each other, unsecured. The Treasury
bill is lending to the government. When the gap between those two widens, it is
telling you that banks have stopped trusting each other.

In a normal month that gap is a few basis points. In September 2008 it went past
three hundred. Same chart, same two rates — and that spread was the clearest
real-time measure we had of the panic.
:::

## 1 · Commercial paper outstanding

```{r}
#| fig-width: 12
#| fig-height: 4.2
cp <- fred("COMPOUT")
if (is.null(cp)) {
  no_data_note("COMPOUT")
} else {
  cp$value <- cp$value / 1000
  ggplot(cp, aes(date, value)) +
    recession_bands(from = min(cp$date)) +
    geom_line(colour = IF_COL$teal700, linewidth = .55) +
    labs(x = NULL, y = "USD trillion",
         title = "Commercial paper outstanding",
         caption = "Shaded: NBER recessions. Source: Federal Reserve Board via FRED.") +
    theme_if()
}
```

::: ask
2007 marks a **structural break**, not a cyclical decline: the series steps
down and does not recover its prior level. Which class of issuer exited the
market, why could it not return, and what replaced that funding?
:::

::: notes
Look at this one before I say anything.

*[wait]*

*[ask]* What happened in 2007?

I want you to notice that this is not a recession dip. The line does not fall
and recover. It steps down and stays down. Twenty years later it has never
returned to the old level.

So who left? The issuers who disappeared were the ones funding long-term assets
with thirty-day paper — structured investment vehicles, the finance arms of
banks. When the paper stopped rolling over, they had two options: sell the
assets, or hand them back to the bank that sponsored them. Both happened, and
both made things worse.

This is the same chart as the BNP Paribas story we opened with. That is what it
looks like in the data.
:::

## 1 · Price and rate

Money market instruments are **zero-coupon**: they are purchased below face
value and redeemed at face value, so the entire return is the **discount**.
Two conventions are used to express it, and they are not interchangeable.

:::: columns
::: {.column width="50%"}
The **discount rate** — how far below face value you paid:

$$i_{discount} = \frac{F-P}{F} \times \frac{360}{n}$$
:::

::: {.column width="50%"}
The **annualised yield** — what you actually earned:

$$i_{yt} = \frac{F-P}{P} \times \frac{365}{n}$$
:::
::::

. . .

::: keyidea
One instrument, two measures. The **bank discount basis** deflates by face
value $F$ on a 360-day year — a convention inherited from pre-computational
market practice. The **annualised yield** deflates by the price actually paid,
$P$, on a 365-day year, and is therefore the true holding-period return
expressed annually. Since $P < F$, the yield always exceeds the discount rate.
Only the yield is comparable across instruments.
:::

::: notes
These instruments pay no coupon. You buy at ninety-eight, you get a hundred at
maturity, and the whole return is that gap. So there is no interest rate to
quote — there is only a discount, and we have to turn it into a rate.

The market does that two ways, and they do not give the same number.

The first one, the discount rate, divides by F — the face value, what you get at
the end — and uses a three-hundred-and-sixty-day year.

The second one, the annualised yield, divides by P — what you actually paid —
and uses three hundred and sixty-five days.

*[pause]*

Three hundred and sixty is not a mistake. It is a convention from before
calculators, when twelve months of thirty days made the arithmetic possible by
hand. We are still using it.

Now: P is always smaller than F. So dividing by P always gives you the bigger
number. The yield is always higher than the discount rate on the same
instrument.

*[ask]* If you were selling me this bill, which of the two would you quote?

Only the yield can be compared with a deposit rate or a bond yield. If you take
one thing from this slide, take that.
:::

## 1 · The two rates, drawn

```{r}
#| fig-width: 10
#| fig-height: 4.0
P <- seq(97, 100, length.out = 200)
F <- 100; n <- 90
d <- rbind(
  data.frame(P = P, rate = (F - P) / F * 360 / n, which = "Discount rate"),
  data.frame(P = P, rate = (F - P) / P * 365 / n, which = "Annualised yield")
)
ggplot(d, aes(P, rate, colour = which)) +
  geom_line(linewidth = .8) +
  scale_colour_manual(values = c(IF_COL$clay, IF_COL$slate)) +
  scale_y_continuous(labels = function(x) paste0(round(x * 100), "%")) +
  labs(x = "Current price", y = "Rate",
       title = "F = 100, n = 90 days",
       caption = "The gap between the two is the 360 / 365 convention plus the choice of denominator.") +
  theme_if()
```

::: notes
Same two formulas, drawn. Face value a hundred, ninety days.

The only thing to see is that the two lines pull apart as the price falls. When
there is almost no discount they agree. The bigger the discount, the more the
choice of denominator matters.

Thirty seconds on this slide and we move on.
:::

## 1 · Problem 2.1

::: ask
A 28-day T-bill costs **$996.73** and is worth **$1,000** at maturity.

1.  What is the **discount rate**?
2.  What is the **annualised yield**?
:::

. . .

:::: columns
::: {.column width="50%"}
$$i_{discount} = \frac{1000-996.73}{1000}\times\frac{360}{28} = 4.20\%$$
:::

::: {.column width="50%"}
$$i_{yt} = \frac{1000-996.73}{996.73}\times\frac{365}{28} = 4.28\%$$
:::
::::

::: notes
Calculators out. This is the kind of question that appears on the exam.

Twenty-eight day bill, you pay nine hundred and ninety-six seventy-three, you
get a thousand.

*[give them two minutes, then reveal]*

Discount rate, four point two zero percent. Yield, four point two eight.

*[pause]*

Eight basis points apart. And everybody in the room would describe both of those
as "the rate on this bill". On a hundred million euros, eight basis points is
eighty thousand a year. This is why the convention matters.
:::

## 1 · Problem 2.2

The Treasury auctions **$2.5 billion** of 91-day bills. Competitive bids:

| Bidder | Bid amount | Bid price |
|:--:|:--:|:--:|
| 1 | $500 million | $0.9940 |
| 2 | $750 million | $0.9901 |
| 3 | $1.5 billion | $0.9925 |
| 4 | $1 billion | $0.9936 |
| 5 | $600 million | $0.9939 |

It also received **$750 million** of noncompetitive bids.

::: ask
Who receives bills, how many, and at what price?
:::

::: notes
This one is an auction. Two and a half billion of ninety-one day bills, five
competitive bidders, and seven hundred and fifty million of noncompetitive bids.

Remember the rule from the Treasury bills tab: noncompetitive bidders take
whatever price comes out, and they are always filled.

*[ask]* So where do we start?

Noncompetitive comes off the top. Two and a half minus zero point seven five
leaves one point seven five billion for the competitive bidders.

Now sort them by price, highest first — the highest price is the lowest interest
rate, so that is the cheapest money for the Treasury. Bidder one at nine nine
four zero. Bidder five. Bidder four. Bidder three.

Work down until the one point seven five billion is gone. Bidder three only gets
part of what they asked for, and bidder two gets nothing at all.

*[ask]* Bidder two bid nine nine zero one. What did they get wrong?
:::

# 2 · The bond market {.divider .center background-color="#06302F"}

## Athens, March 2012 {.hook background-color="#FAF7F2"}

Greece restructured approximately EUR 200 billion of sovereign bonds. Private
holders accepted a reduction of roughly 53% of face value — the largest
sovereign restructuring on record. The same bonds had carried investment-grade
ratings a few years earlier.

::: incremental
-   A bond is a **contractual claim**: specified payments on specified dates,
    enforceable under a governing law.
-   Enforceability is not payment. **Credit risk** is the probability that the
    contract is not performed, and it is priced.
-   For sovereign debt the issuer can also alter the contract: Greece inserted
    **collective action clauses** retroactively, binding dissenting holders to
    the majority. This has no corporate equivalent.
:::

::: notes
March 2012. Greece restructures about two hundred billion euros of government
bonds. Private holders take a cut of roughly fifty-three percent of face value.
It is the largest sovereign restructuring there has ever been.

And a few years before that, those same bonds were rated investment grade.

*[pause]*

Everything I am about to teach you for the next ten slides assumes a bond is a
contract. Specified payments, specified dates, enforceable in court. That is
true. It is just not the same thing as being paid.

And with a government there is a further problem. Greece inserted collective
action clauses into bonds that had already been issued. Retroactively. Which
meant that if enough holders agreed to the deal, the ones who refused were bound
anyway.

*[ask]* Think about what that means. The borrower rewrote the contract after you
signed it. Can a company do that?

No. This is a risk that exists only in sovereign debt, and it is why we price it
separately.
:::

## 2 · The vocabulary

| Term | What it means |
|:---|:---|
| Coupon interest rate | The annual interest rate, usually fixed for the whole life |
| Current yield | Coupon payment divided by the current price |
| Face amount / par value | The maturity value |
| Indenture | The contract that specifies the terms of the loan agreement |
| Market rate | The rate currently in effect on securities of like risk and maturity |
| Maturity | Number of periods until the security matures |
| Yield to maturity | The yield if you buy at the current price and hold to maturity |

::: keyidea
The **coupon rate** is fixed in the indenture at issue. The **market rate** on
comparable risk and maturity varies continuously. If the contractual payments
cannot adjust, the only variable that can reconcile the two is the **price**.
The whole of bond valuation follows from that single observation.
:::

::: notes
Take a minute to read this. It is all in the handout.

*[wait, take questions]*

I only want to insist on one line of it.

The coupon rate is written into the indenture on the day the bond is issued, and
it never changes. The market rate — what investors currently demand for this
much risk and this much time — changes every day.

*[ask]* So if the payments cannot move, and what people demand has gone up, what
is left to adjust?

The price. That is it. That is the whole of bond valuation, and you just said it
yourselves.

One warning. Current yield and yield to maturity are not the same thing, and
telling me they are is the single most common way to lose marks on this part of
the exam. We come back to it in three slides.
:::

## 2 · The instruments

::: {.panel-tabset}

### Treasuries

-   **Treasury notes**: 1 to 10 years. **Treasury bonds**: 10 to 30 years.
-   Nominal default risk is negligible because the obligation is denominated in
    a currency the issuer creates. That is a statement about **nominal**
    default only.
-   Two risks remain and are examinable: **inflation risk**, which erodes the
    real value of fixed payments, and **interest-rate risk**, the capital loss
    on an existing bond when market rates rise. The latter increases with
    maturity.

### Municipals

-   Issued by local, county and state governments to fund public projects.
-   **Tax free**, which is the whole point. Compare like with like:

    $$\text{after-tax rate} = \text{taxable rate}\times(1-\text{marginal tax rate})$$

-   **General obligation** bonds are repaid from taxes; **revenue** bonds from
    user fees on the project itself.
-   They are **not** default free. Detroit, 2013.

### Corporates

-   Face value usually USD 1,000, interest paid semi-annually.
-   **Restrictive covenants** limit dividends, new debt, ratios — they exist
    because shareholder and bondholder interests conflict.
-   A **call provision** lets the issuer buy the bond back. A **conversion**
    feature lets the holder turn it into equity.
-   **Secured** bondholders have first claim on a specific asset; **unsecured**
    ones do not, and charge more for it.

### Ratings

-   Agencies map credit risk onto an ordinal scale from **Aaa/AAA** downwards.
    Below **Baa3/BBB-** an issue is **speculative grade**.
-   That threshold has effects beyond information. Many institutional mandates
    — insurers, pension funds, some trusts — **prohibit** holdings below it, so
    a one-notch downgrade can compel sales by investors with no independent
    view of the credit.
-   The result is **forced selling** into a falling market: a price movement
    caused by mandate design rather than by fundamentals.
-   Note also the **issuer-pays** model: the entity being rated selects and
    remunerates the agency. The conflict of interest this creates is central to
    the post-2008 literature on structured credit.

:::

::: notes
Four tabs. I want two of them.

*[municipals tab]*

Municipal bonds are issued by cities and states, and the interest is tax free.
That is the whole point of them, and it means you cannot compare them with a
corporate bond directly.

To compare, you take the corporate rate and multiply by one minus your marginal
tax rate. That gives you what you actually keep. We do this as a problem in a
few slides, so make a note of the formula.

And they are not safe. Detroit filed for bankruptcy in 2013 — the pensioners and
the bondholders were standing in the same queue.

*[ratings tab]*

Now ratings, and this is the one I want you to remember.

Below BBB minus, a bond stops being investment grade and becomes speculative.
Junk, if you want the market word.

Here is why that line does more than describe risk. A lot of institutions —
insurers, pension funds — are simply *not allowed* to hold anything below it.
It is written into their mandate.

*[ask]* So what happens to the price when a bond is downgraded from BBB minus to
BB plus?

It falls — and it falls because people who have no opinion at all about the
company are forced to sell it. That is forced selling, and it is a price move
caused by regulation, not by anything the company did.

One last thing. Who pays the rating agency? The company being rated chooses the
agency and pays the fee. Sit with that for a second. That conflict of interest is
at the centre of everything written about structured credit after 2008.
:::

## 2 · The rating scales

:::: columns
::: {.column width="52%"}
| Moody's | S&P | Fitch | Meaning |
|:--:|:--:|:--:|:---|
| Aaa | AAA | AAA | Best quality, least credit risk |
| Aa1-Aa3 | AA+/AA/AA- | AA+/AA/AA- | High quality, very strong capacity |
| A1-A3 | A+/A/A- | A+/A/A- | Upper-medium grade |
| Baa1-Baa3 | BBB+/BBB/BBB- | BBB+/BBB/BBB- | Medium quality, adequate capacity |
:::

::: {.column width="48%"}
| Moody's | S&P | Fitch | Meaning |
|:--:|:--:|:--:|:---|
| Ba1-Ba3 | BB+/BB/BB- | BB+/BB/BB- | Speculative |
| B1-B3 | B+/B/B- | B+/B/B- | Very speculative |
| Caa-C | CCC/CC/C | CCC/CC/C | Highly speculative, near or in default |
| — | D | DDD/DD/D | In default |
:::
::::

::: ask
The investment-grade boundary lies between Baa3 and Ba1. Ratings are ordinal
and credit quality is continuous, so why should **one notch** produce a
discontinuity in price?
:::

::: notes
You do not need to memorise this ladder. You need to know where one line is.

It is between Baa3 and Ba1 — investment grade above, speculative below.

*[ask]* Credit quality is continuous. A company does not become a different
company overnight. So why should one notch produce a jump in the price?

*[wait]*

Because of what I just said. The mandates. A pension fund that is prohibited
from holding speculative grade has to sell, whatever it thinks about the credit.
Everybody in that position sells at the same time, and there is no natural buyer
on the other side.

The market has a name for a bond that crosses that line: a fallen angel. Worth
knowing.
:::

## 2 · Rates by credit quality

```{r}
#| fig-width: 12
#| fig-height: 4.4
ids <- c(AAA = "Corporate Aaa", BAA = "Corporate Baa", GS20 = "20-year Treasury")
series <- lapply(names(ids), function(id) {
  d <- fred(id)
  if (is.null(d)) return(NULL)
  d$series <- ids[[id]]
  d
})
series <- Filter(Negate(is.null), series)

if (!length(series)) {
  no_data_note("AAA / BAA / GS20")
} else {
  d <- do.call(rbind, series)
  d <- d[d$date >= as.Date("1980-01-01"), ]
  ggplot(d, aes(date, value, colour = series)) +
    recession_bands(from = as.Date("1980-01-01")) +
    geom_line(linewidth = .5) +
    scale_colour_manual(values = unname(unlist(
      IF_COL[c("slate", "clay", "teal700")]))) +
    labs(x = NULL, y = "Percent",
         title = "The credit spread is the distance between these lines",
         caption = "Shaded: NBER recessions. Source: Moody's and the Treasury via FRED.") +
    theme_if()
}
```

::: ask
The spread is widest precisely when firms most need external finance. Identify
the two distinct mechanisms behind this, and explain why their coincidence
makes credit supply **procyclical**.
:::

::: notes
Three lines. Aaa corporates at the top of the quality range, Baa lower down, and
the twenty-year Treasury.

The distance between the Baa line and the Treasury line is the credit spread.
That is what you are paid for taking corporate risk instead of government risk.

*[point at 2008, then 2020]*

Look at what it does in a crisis. It goes from about two percentage points to
over six.

*[ask]* Now think about when a company most needs to raise money. Is it in a
boom, or in a downturn?

Exactly. The spread is widest precisely when firms most need to borrow.

Two separate things are happening at once. The probability of default genuinely
has gone up. And investors want less risk than they did last month. Both push
the same way, at the same time.

That is what we mean by credit being procyclical. Finance is hardest to get
exactly when it is most needed — and that is a policy problem, not a quirk of
the market.
:::

## 2 · The price of a bond

The price is the present value of the contractual payments, discounted at the
rate the market currently requires on claims of comparable risk and maturity:

$$P = \frac{C}{(1+r)} + \frac{C}{(1+r)^2} + \dots + \frac{C}{(1+r)^n} + \frac{F}{(1+r)^n}$$

. . .

::: keyidea
Price and yield are **inversely related**, necessarily. The payments $C$ and
$F$ are fixed by the indenture, so a higher required return can only be
delivered by a lower price.
:::

::: notes
Let us build this rather than just look at it.

Start with the simplest bond you can imagine. One year, one payment at the end.
What is it worth today? The payment, discounted one year. Everybody is fine with
that.

Now make it two years. Two payments, discounted one and two years.

Keep going and you have the expression on the slide. The only new thing is the
very last period, where you get the final coupon *and* the face value back.

*[pause]*

*[ask]* Now. The coupon is fixed in the contract. The face value is fixed in the
contract. If I demand a higher return than the last person who bought this bond,
what has to happen?

The price has to fall. It is the only variable left. Price and yield move in
opposite directions, and not as a rule of thumb — there is nothing else that can
move.
:::

## 2 · Two ways of quoting the return

The **yield to maturity** is the internal rate of return $r$ that equates the
present value of the contractual payments to the observed market price:

$$0 = -P + \sum_{t=1}^{n}\frac{C}{(1+r)^t} + \frac{F}{(1+r)^n}$$

. . .

The **current (coupon) yield** ignores any capital gain or loss at redemption:

$$r_c = \frac{C}{P}$$

::: keyidea
Two assumptions are embedded in the yield to maturity: that the bond is **held
to maturity**, and that every coupon is **reinvested at $r$**. Neither holds in
general, which is why realised returns differ from quoted yields.
:::

::: notes
The yield to maturity is the discount rate that makes the price you see and the
payments you expect balance exactly. It is an internal rate of return.

Be honest with yourselves about this equation: you cannot solve it for r by
rearranging. Beyond two periods there is no formula. In the exam I will give you
either a short maturity or a calculator. In a bank, software does it.

*[pause]*

The current yield underneath is a shortcut. Coupon divided by price. It ignores
what happens at redemption entirely.

*[ask]* Suppose I buy a bond for eighty and it repays a hundred at maturity.
Does the current yield capture that extra twenty?

No. That is why it is only an approximation, and the next slide shows you how
wrong it gets.

One more thing, and this is the part people skip. The yield to maturity assumes
two things. That you hold the bond to the end. And that you reinvest every
single coupon at that same rate r.

*[ask]* What if rates fall after you buy?

Then you are reinvesting those coupons at less than r, and what you actually
earn is below the yield you were quoted. That is reinvestment risk. It is the
mirror image of the price risk we just talked about — when rates fall your price
goes up, but your reinvestment gets worse.
:::

## 2 · Coupon yield is only an approximation

```{r}
#| fig-width: 10
#| fig-height: 4.0
Pr <- seq(50, 150, length.out = 300)
F <- 100; C <- 5; n <- 10
ytm <- sapply(Pr, function(p) {
  f <- function(r) -p + sum(C / (1 + r)^(1:n)) + F / (1 + r)^n
  tryCatch(uniroot(f, c(-0.5, 1))$root, error = function(e) NA_real_)
})
d <- rbind(
  data.frame(P = Pr, rate = ytm,    which = "Yield to maturity"),
  data.frame(P = Pr, rate = C / Pr, which = "Coupon yield")
)
ggplot(d, aes(P, rate, colour = which)) +
  geom_hline(yintercept = 0, colour = IF_COL$muted, linewidth = .3) +
  geom_vline(xintercept = 100, linetype = "dotted", colour = IF_COL$muted) +
  geom_line(linewidth = .8) +
  scale_colour_manual(values = c(IF_COL$clay, IF_COL$slate)) +
  scale_y_continuous(labels = function(x) paste0(round(x * 100), "%")) +
  labs(x = "Current price", y = NULL,
       title = "F = 100, coupon 5, 10 years to maturity",
       caption = "The two agree only at par. Away from par the coupon yield misleads.") +
  theme_if()
```

::: notes
Blue is the true yield to maturity. Red is the current yield shortcut.

Notice they only touch in one place. At a price of a hundred — at par.

*[ask]* Look to the left of par, where the bond is cheap. Which line is higher,
and why?

The true yield is higher, because if you buy at eighty and get a hundred back,
there is a capital gain the current yield is not counting.

*[ask]* And to the right of par?

The other way round. You paid a hundred and twenty, you get a hundred back, and
there is a capital loss the current yield is ignoring.

So the shortcut is not just imprecise. It is wrong in a direction you can
predict. Below par it flatters the bond. Above par it flatters it too.
:::

## 2 · Problem 2.3

::: ask
1.  A corporate bond yields **9%**, a municipal bond **6.75%**. Which do you buy?
2.  A corporate bond yields **5%**, a municipal **3.5%**, and your marginal tax
    rate is **28%**. Which now?
:::

. . .

:::: columns
::: {.column width="50%"}
**First case.** The comparison needs a tax rate to be meaningful. At a marginal
rate of 25%, the corporate is worth $9\% \times 0.75 = 6.75\%$ after tax —
exactly the municipal. Above 25% the municipal wins.
:::

::: {.column width="50%"}
**Second case.** $5\% \times (1-0.28) = 3.60\%$ after tax, against **3.5%**
tax-free. The corporate wins, narrowly.
:::
::::

::: notes
Two parts. Read the first one.

*[wait — let someone attempt an answer]*

Nine percent against six seventy-five. Most of you will say the corporate, and
I want to push back on that.

*[ask]* What do you need to know before you can answer at all?

The tax rate. Without it the question has no answer. The municipal is tax free
and the corporate is not, so you are comparing two numbers that are not
comparable. That is the lesson of part one — notice when a question is missing
something.

If you put the tax rate at twenty-five percent, nine times zero point seven five
is six seventy-five. Exactly the municipal. Above twenty-five percent, the
municipal wins.

Part two gives you the rate. Five percent, twenty-eight percent tax. Five times
zero point seven two is three point six zero, against three point five tax free.
The corporate wins — but only by ten basis points.
:::

## 2 · Problem 2.4

::: ask
1.  What is the **coupon yield** on a bond with face value $1,000, current
    price $921.01 and a coupon rate of **10.95%**?
2.  What is the **price** of a two-year, 10% coupon bond paying **semi-annually**,
    face value $1,000, if the required return is **12%**?
:::

. . .

:::: columns
::: {.column width="46%"}
$$r_c = \frac{109.50}{921.01} = 11.89\%$$
:::

::: {.column width="54%"}
Four periods, coupon 50, discount rate 6% per period:

$$P = \sum_{t=1}^{4}\frac{50}{1.06^t} + \frac{1000}{1.06^4} = 965.35$$
:::
::::

::: notes
Part one is straightforward. The coupon is ten point nine five percent of a
thousand, so a hundred and nine fifty a year. Divide by the price of nine
twenty-one. Eleven point eight nine percent.

Part two is where people lose marks, so go slowly.

*[ask]* Semi-annual coupons. What has to change in the formula?

Three things, and you need all three. Halve the coupon — fifty, not a hundred.
Halve the rate — six percent per period, not twelve. And double the number of
periods — four, not two.

Most people remember one of the three. Some remember two. Three changes.

The answer is nine sixty-five thirty-five. The bond trades below par, and it
should — the coupon is ten percent and the market wants twelve.
:::

# 3 · The stock market {.divider .center background-color="#06302F"}

## The residual claim {.hook background-color="#FAF7F2"}

A bondholder holds a claim to a fixed sum. A shareholder holds a claim to
whatever remains once all prior claims are satisfied — a quantity bounded below
by zero and unbounded above.

::: incremental
-   Equity returns arise from **dividends** and **capital appreciation**.
-   Shareholders are **residual claimants** on the firm's assets and cash flows,
    and hold **control rights** through the vote.
-   Ranking last in liquidation is what gives equity its asymmetric payoff, and
    what makes the **agency relationship** between managers and shareholders a
    central problem of corporate finance.
:::

::: notes
Here is the difference between what we have been doing and what comes next.

A bondholder is owed a number. It is written down. If the company can pay, they
get exactly that and not a cent more.

A shareholder owns what is left after everyone else has been paid. Which can be
a great deal, or can be nothing at all.

*[ask]* Imagine a company about to be liquidated. Which would you rather hold?

*[ask]* Now a company about to triple in size. Which now?

That is the whole thing. Everything that follows — why shareholders vote, why
dividends are discretionary, why we value equity by discounting dividends —
comes from being last in the queue.

And one more consequence, which you will meet properly in corporate finance:
because shareholders own the upside but are protected on the downside by limited
liability, their interests and the managers' interests are not automatically the
same. That problem has a name. It is the agency problem.
:::

## 3 · Common, preferred, and foreign

:::: columns
::: {.column width="50%"}
**Common stock** — the right to vote for directors and on certain issues, and
to receive dividends if the board declares them.

**Preferred stock** — usually **no vote**, but a **fixed dividend** paid before
common. Closer to a bond than most students expect.
:::

::: {.column width="50%"}
**Foreign stocks** are useful for diversification but awkward to buy if they are
not listed locally.

**American depository receipts (ADRs)** solve it: a bank buys the foreign
shares and issues receipts against them that trade on U.S. exchanges.
:::
::::

::: keyidea
An **index** is a summary statistic for a defined set of securities, not an
asset. Its **weighting scheme** determines what the statistic measures — and
therefore what a given movement in it means.
:::

::: notes
Common stock is what you picture: a vote, and a dividend if the board declares
one.

Preferred stock is the interesting one. Usually no vote. Fixed dividend. Paid
before common shareholders.

*[ask]* Fixed payment, paid ahead of others, no control. Is that equity, or is
that debt?

The honest answer is that it sits between the two, and that is exactly why
accountants and rating agencies treat it inconsistently. If you see it in a
balance sheet, look at how it has been classified — it is not always obvious.

On the right, ADRs. If you want to hold a Brazilian company but you can only
trade in New York, a bank buys the Brazilian shares and issues receipts against
them that trade in dollars on a US exchange.

*[pause]*

If that sounds familiar, it should. It is the same engineering as the eurodollar
deposit from an hour ago: take a claim, repackage it so it settles in the
currency and the legal system the buyer is comfortable with.

Last line. An index is not an investment. It is a summary statistic. And how you
weight it decides what it is a summary *of* — which is the next slide.
:::

## 3 · Where the trading happens

:::: columns
::: {.column width="50%"}
**Organised exchanges**

-   NYSE trades roughly 4 to 7 billion shares a day.
-   "Organised" once meant a physical floor; electronic networks have largely
    replaced that idea.
-   **Listing requirements** exclude small firms.
-   Auction market with floor specialists; around 25% of trades filled directly
    by the specialist.
:::

::: {.column width="50%"}
**Over the counter**

-   NASDAQ is the obvious example, with roughly 3,000 securities listed.
-   **Dealers** stand ready to make a market — without one, the equity would be
    hard to trade at all.
-   Multiple market makers post bid and ask prices for the same security.
:::
::::

::: ask
Dealers quote a **bid-ask spread** and profit from it. Competition constrains
that spread — but a spread of zero would not be sustainable either. What costs
must the spread compensate the dealer for, and what happens to liquidity when
those costs rise?
:::

::: notes
Two ways to organise a market.

Exchanges on the left — one central place where orders meet. The New York Stock
Exchange still uses the word organised, but the trading floor you have seen in
films is mostly theatre now. It is computers.

Over the counter on the right. No central place. Dealers hold inventory and
quote you a price to buy and a price to sell.

*[ask]* The dealer buys from you at the bid and sells to someone else at the
ask, and keeps the difference. So why does competition not push that spread to
zero?

*[wait]*

Two reasons, and they are both real costs.

First, the dealer is holding inventory. They bought your shares and now they are
exposed until they sell them.

Second — and this is the one worth thinking about — some of the people trading
with the dealer know something the dealer does not. If you are selling to me
because you know bad news is coming, I lose. I cannot tell who those people are,
so I widen the spread against everyone.

*[ask]* Now, in a crisis, when nobody knows what anything is worth. What happens
to that second cost, and what happens to the spread?

It widens. Which is the same lesson as this morning. Liquidity disappears
exactly when you need it.
:::

## 3 · Price-weighted versus value-weighted

The **DJIA** is a price-weighted average of 30 stocks. The **S&P 500** is a
market-value-weighted index of 500.

```{r}
#| fig-width: 12
#| fig-height: 3.9
dj <- fred("DJIA")
sp <- fred("SP500")
if (is.null(dj)) {
  no_data_note("DJIA")
} else {
  dj <- dj[dj$date >= as.Date("2015-01-01"), ]
  dj$index <- 100 * dj$value / dj$value[1]
  dj$series <- "Dow Jones Industrial Average"
  d <- dj[, c("date", "index", "series")]
  if (!is.null(sp)) {
    sp <- sp[sp$date >= as.Date("2015-01-01"), ]
    sp$index <- 100 * sp$value / sp$value[1]
    sp$series <- "S&P 500"
    d <- rbind(d, sp[, c("date", "index", "series")])
  }
  ggplot(d, aes(date, index, colour = series)) +
    geom_line(linewidth = .5) +
    scale_colour_manual(values = c(IF_COL$slate, IF_COL$teal500)) +
    labs(x = NULL, y = "Index, first observation = 100",
         title = "Thirty stocks by price, or five hundred by market value",
         caption = "Source: S&P Dow Jones Indices via FRED.") +
    theme_if()
}
```

::: ask
Under price weighting, a USD 400 constituent carries roughly thirteen times the
weight of a USD 30 constituent, irrespective of the firms' market values. State
the implication for a **stock split**, and hence whether the index measures
anything economically meaningful.
:::

::: notes
Two indices, two completely different ideas.

The S&P 500 weights by market value. If a company is worth two percent of the
market, it is two percent of the index. That is what you would own if you bought
the whole market.

The Dow weights by *price per share*. Not by size. By the number printed on the
share.

*[ask]* So a four-hundred-dollar share and a thirty-dollar share. Which one
moves the Dow more?

The four-hundred-dollar one — about thirteen times more — regardless of whether
the companies are the same size or one is a hundred times bigger than the other.

*[pause]*

Now here is the part that should bother you. A company does a two-for-one stock
split. Same company, same value, twice as many shares at half the price.

*[ask]* What just happened to its weight in the Dow?

It halved. Nothing happened to the company at all.

*[ask]* So why does every news programme still quote the Dow?

Because it is from 1896 and people recognise the number. That is a reason from
journalism, not from finance.
:::

## 3 · What a share is worth

The **Gordon growth model** — a special case of the dividend discount model, in
which dividends grow at a constant rate $g$ in perpetuity and $k_e > g$:

$$P = \sum_{t=1}^{\infty}\frac{Div_t}{(1+k_e)^t} = \frac{Div_1}{k_e - g}$$

:::: columns
::: {.column width="50%"}
**Geometric growth**, for backing out $g$ from two observations:

$$g = \left(\frac{\text{final}}{\text{initial}}\right)^{1/n} - 1$$
:::

::: {.column width="50%"}
**Price-to-earnings**, a relative valuation measure:

$$PE = \frac{\text{Price}}{\text{Earnings}}$$
:::
::::

::: keyidea
The denominator $k_e - g$ governs everything. As $g \to k_e$ the valuation
diverges, so near that boundary the price is **highly sensitive** to small
revisions in either parameter. This is not a defect of the model but a property
of perpetual growth — and it explains why estimates of high-growth firms
disperse so widely, and why such valuations reprice violently on modest news.
:::

::: notes
A share pays you dividends, forever. So its value is the present value of all of
them. Write that down and you get the sum on the left.

If the dividend grows at a constant rate g, that infinite sum collapses to the
expression on the right. Next year's dividend, over the required return minus
the growth rate. That is the Gordon growth model, and you have met growing
perpetuities before in corporate finance.

Two conditions, and they are not fine print. The growth rate has to be constant
forever. And k has to be bigger than g.

*[ask]* Why must k be bigger than g? What is wrong with a company that grows
faster than the required return, forever?

*[wait]*

Eventually it is bigger than the entire economy. That is your sanity check — if
you ever get g above k in the exam, you have made an arithmetic error, not
discovered a fantastic company.

Underneath: the PE ratio. Price over earnings. A market number on top, an
accounting number on the bottom.

*[ask]* A company has a PE of forty. Is it expensive, or is it expected to grow?

Both answers are right, and you cannot tell which from the ratio. That is the
limitation of every multiple.
:::

## 3 · The denominator does all the work

```{r}
#| fig-width: 12
#| fig-height: 4.0
div1 <- 2
a <- data.frame(x = seq(.06, .24, length.out = 200))
a$price <- div1 / (a$x - .05); a$panel <- "Varying the required return (g = 5%)"
b <- data.frame(x = seq(.01, .19, length.out = 200))
b$price <- div1 / (.20 - b$x); b$panel <- "Varying the growth rate (k = 20%)"
d <- rbind(a, b)
ggplot(d, aes(x, price)) +
  geom_line(colour = IF_COL$clay, linewidth = .8) +
  facet_wrap(~panel, scales = "free_x") +
  scale_x_continuous(labels = function(x) paste0(round(x * 100), "%")) +
  coord_cartesian(ylim = c(0, 250)) +
  labs(x = NULL, y = "Stock price",
       title = "Gordon growth, with the first dividend fixed at 2",
       caption = "Both curves are the same asymptote seen from two directions.") +
  theme_if() +
  theme(strip.text = element_text(colour = IF_COL$teal700, face = "bold"))
```

::: notes
Two panels, same model. On the left I vary the required return. On the right I
vary the growth rate.

Both go vertical. It is the same asymptote, seen from two directions — as g gets
close to k, the denominator goes to zero and the price explodes.

*[pause]*

This is not a flaw in the model. It is what perpetual growth means. But look at
what it implies.

Near that edge, change your growth assumption by half a percentage point and the
valuation changes enormously.

*[ask]* So two analysts, same company, almost identical assumptions. How far
apart can their target prices be?

Very far. And that is why high-growth companies reprice so violently on news
that sounds small. Nothing happened to this year's earnings. Something happened
to the assumption in the denominator.

You have watched this happen — in 2000, and again in 2021 and 2022.
:::

## 3 · Problem 2.5

::: ask
1.  A stock is expected to pay a dividend of **$0.16** and to be worth **$60**
    in a year. At a **12%** discount rate, what is it worth now?
2.  If the industry PE ratio is **16** and the firm earns **$1.13 per share**,
    what is the price?
:::

. . .

:::: columns
::: {.column width="50%"}
$$P_0 = \frac{0.16 + 60}{1.12} = 53.71$$
:::

::: {.column width="50%"}
$$P = 16 \times 1.13 = 18.08$$
:::
::::

::: notes
Two parts, one line each. But they are two different methods, and that is the
point.

Part one. You expect a dividend of sixteen cents and a price of sixty in a year,
and you discount at twelve percent. Sixty point one six over one point one two.
Fifty-three seventy-one.

Part two. The industry trades at sixteen times earnings, this firm earns a
dollar thirteen. Sixteen times one thirteen. Eighteen zero eight.

*[pause]*

Part one is a discounted cash flow. You forecast, then you discount.

Part two is relative valuation. You assume the market has priced similar
companies correctly, and you apply their multiple.

*[ask]* Which one do you trust more?

That argument is most of what equity analysts do for a living. Notice that the
first needs you to forecast, and the second needs the market to be right.
:::

## 3 · Problem 2.6

::: ask
The DJIA was **10,000** in 2010 and **20,000** in 2017. What was the annual
**geometric growth rate**?
:::

. . .

$$g = \left(\frac{20000}{10000}\right)^{1/7} - 1 = 2^{1/7} - 1 = 10.41\%$$

::: notes
Ten thousand in 2010, twenty thousand in 2017. What is the annual growth rate?

*[ask]* Careful. How many years is that?

Seven, not eight. Count the gaps, not the dates — that is the most common error
on this question.

So it is two to the power of one seventh, minus one. Ten point four one percent.

Notice the shape: the index doubled in seven years at about ten percent a year.
That is a useful number to carry around. Roughly seven years to double at ten
percent.
:::

## Next session {.center}

Session 4 finishes securities markets: **mortgages**, **foreign exchange** and
**derivatives**, then the green bonds discussion.

::: keyidea
Bring **Sartzetakis (2021)** on green bonds, read — and the security your group
wants to analyse. Not "a stock". A specific one, with a ticker.
:::

::: {.source}
Press **`O`** for the slide overview · **`C`** to draw · **`S`** for speaker notes · **`F`** for fullscreen · **`E`** then print for a PDF of the slides.
:::

::: notes
That is money markets, bonds and stocks. Next session we finish securities
markets: mortgages, foreign exchange, and derivatives, and then we discuss the
green bonds paper.

Two things to bring.

First, read Sartzetakis. Read it before the session, not after — the discussion
is part of what I assess, and it does not work if I am the only one who has read
it.

Second, your group's security. Not "a stock". A specific one, with a ticker,
that you can find real data on. Choose it now. Every year somebody chooses in
the week before the presentation and it shows.
:::

<!-- ===================================================================
     SESSION 4 - not rendered yet. Uncomment, or split into its own
     folder the way session 2 was, when you are ready.
     Sections 4 (mortgages), 5 (foreign exchange), 6 (derivatives)
     and 7 (Sartzetakis discussion).
     =================================================================== -->

<!--

# 4 · Mortgage markets {.divider .center background-color="#06302F"}
## The market that ate the world {.hook background-color="#FAF7F2"}

By 2006 an American mortgage was not a loan a bank held. It was an input:
pooled, tranched, rated, and sold to a German Landesbank that had never seen
Ohio.

::: incremental
-   A mortgage is a **long-term loan secured by real estate**.
-   It is **amortised**: one fixed payment covers both interest and principal
    every month.
:::
## 4 · The types

| Type | What it is |
|:---|:---|
| Conventional | Not guaranteed; usually needs private mortgage insurance and 5-20% down |
| Insured | Guaranteed by a government agency; low or zero down payment |
| Fixed-rate | The rate is fixed for the life of the mortgage |
| Adjustable-rate (ARM) | Rate tied to another security, adjusted periodically within annual limits |
| Graduated-payment (GPM) | Low initial payment rising each year; amortises in 30 years |
| Growing-equity (GEM) | Payment rises each year; amortises in less than 30 years |
| Second mortgage | Secured by a second lien; used for credit lines and home improvement |
| Reverse annuity (RAM) | The lender pays the borrower monthly; due when the property is sold |

::: ask
Which of these transfers **interest rate risk** from the lender to the borrower?
And who tends to be sold those?
:::
## 4 · Securitisation, in four steps

::: incremental
1.  The **secondary mortgage market** was created by the federal government
    after WWII, when it set up Fannie Mae. It grew quickly from the mid-1980s.
2.  Individual mortgages default unpredictably, so they are **pooled** —
    hundreds at a time — on the argument that diversification tames the risk.
3.  Rights to the pool's cash flows are sold as separate securities:
    **pass-throughs**, or mortgage-backed securities.
4.  The borrower's monthly payment passes through a trustee before reaching the
    investors.
:::

. . .

::: ask
Step 2 is the load-bearing one. Diversification removes risk that is
**independent** across loans. What happens to a pool of American mortgages if
house prices fall everywhere at once?
:::

::: notes
This is the hinge of the whole session and it pays to slow down. The 2008
answer: the correlation everyone assumed was near zero turned out to be near
one, and every tranche above the bottom was priced on that assumption.
:::
## 4 · Mortgage rates track long Treasuries

```{r}
#| fig-width: 12
#| fig-height: 4.4
ids <- c(MORTGAGE30US = "30-year fixed mortgage", GS10 = "10-year Treasury")
series <- lapply(names(ids), function(id) {
  d <- fred(id)
  if (is.null(d)) return(NULL)
  d$series <- ids[[id]]
  d
})
series <- Filter(Negate(is.null), series)

if (!length(series)) {
  no_data_note("MORTGAGE30US / GS10")
} else {
  d <- do.call(rbind, series)
  ggplot(d, aes(date, value, colour = series)) +
    recession_bands(from = min(d$date)) +
    geom_line(linewidth = .5) +
    scale_colour_manual(values = c(IF_COL$slate, IF_COL$clay)) +
    labs(x = NULL, y = "Percent",
         title = "The mortgage rate is a spread over the long Treasury",
         caption = "Shaded: NBER recessions. Source: Freddie Mac and the Treasury via FRED.") +
    theme_if()
}
```

::: ask
The gap between the two is not constant. What widens it?
:::
## 4 · Loan and annuity

:::: columns
::: {.column width="50%"}
What you can **borrow**, given what you can pay each period:

$$C_0 = a \times \left[\frac{1-(1+i)^{-T}}{i}\right]$$
:::

::: {.column width="50%"}
What you must **pay** each period, given what you borrowed:

$$a = C_0 \times \left[\frac{i}{1-(1+i)^{-T}}\right]$$
:::
::::

```{r}
#| fig-width: 10
#| fig-height: 3.3
i <- 0.05 / 12; T <- 20 * 12
d <- data.frame(a = seq(1000, 3000, by = 25))
d$loan <- d$a * (1 - (1 + i)^(-T)) / i
ggplot(d, aes(a, loan)) +
  geom_line(colour = IF_COL$slate, linewidth = .8) +
  scale_y_continuous(labels = function(x) format(x, big.mark = " ", scientific = FALSE)) +
  labs(x = "Monthly payment", y = "Loan you can support",
       title = "5% nominal, 20 years: i = 0.417% per month, T = 240 months",
       caption = "Linear in the payment, which is why affordability is really about the rate.") +
  theme_if()
```
## 4 · Problem 2.7

::: ask
You want to borrow **$100,000**. Choose between a **12% 30-year** mortgage and
an **11.5%** mortgage with **2 discount points**. Which?
:::

::: {.smaller}
A discount point is 1% of the loan paid up front: on a $200,000 loan, 1 point
costs $2,000, half a point $1,000, a quarter point $500.
:::

. . .

::: keyidea
Two points on $100,000 is $2,000 today against 50 basis points saved every month
for thirty years. Compute both monthly payments with the annuity formula, then
ask how long it takes the saving to repay the $2,000. **The answer depends
entirely on how long you stay in the house.**
:::

::: notes
This is the one problem in the session with a genuinely conditional answer.
If they move in four years the points are a bad deal; if they hold to maturity
they are a good one. Getting them to state the break-even horizon is the goal.
:::
# 5 · The foreign exchange market {.divider .center background-color="#06302F"}
## The biggest market nobody sees {.hook background-color="#FAF7F2"}

No building, no opening bell, no ticker on the news. Over-the-counter, bank to
bank, and turnover in the **trillions of dollars a day**.

::: incremental
-   What is traded is **bank deposits denominated in different currencies**.
-   Trades are large — typically in excess of USD 1 million.
-   We met the price in session 2. Now we meet the market.
:::
## 5 · Purchasing power parity

**Absolute PPP** — the exchange rate equals the ratio of price levels:

$$E_{\$/\mathrm{EUR}} = \frac{P_{US}}{P_{EUR}}$$

. . .

**Relative PPP** — the rate of depreciation equals the **inflation differential**:

$$\frac{\Delta E_{\$/\mathrm{EUR}}}{E_{\$/\mathrm{EUR}}} = \pi_{US} - \pi_{EUR}$$

. . .

::: ask
PPP works in the long run and fails in the short run. Given what it assumes,
which of its assumptions is the one that fails first?
:::

::: notes
Answers to steer towards: differentiated goods with brand names and legal
protection give firms pricing power, so identical goods do not have identical
prices; and prices are sticky, so they do not adjust quickly to market
conditions. Both break the arbitrage that PPP relies on.
:::
## 5 · Spot, forward, and what moves them

:::: columns
::: {.column width="46%"}
**Spot rate** — near-immediate exchange of bank deposits.

**Forward rate** — exchange at an agreed future date, at a price agreed now.
:::

::: {.column width="54%"}
Rates move when money demand or supply moves, which happens on expected changes in:

-   **relative price levels**
-   **trade barriers** — tariffs and quotas
-   **preferences** for domestic versus foreign goods
-   **relative productivity**
:::
::::

. . .

::: keyidea
Every item on that list is an **expected change**. The spot rate today is
mostly a statement about tomorrow.
:::
## 5 · What an appreciation does

::: incremental
-   Domestic goods become **more expensive abroad**; foreign goods become
    cheaper at home.
-   Domestic businesses become **less competitive**.
-   Domestic consumers gain **purchasing power**.
:::

. . .

::: ask
So is a strong currency good news? Ask a French exporter and a French
holidaymaker on the same day and you get two answers.
:::
# 6 · Derivative markets {.divider .center background-color="#06302F"}
## A price for a promise {.hook background-color="#FAF7F2"}

In 1995 Barings Bank, 233 years old, was destroyed in weeks by one trader's
position in Nikkei futures. The instruments were not exotic. The position was
simply unhedged and unsupervised.

::: incremental
-   **Hedging** means using a financial transaction to reduce or remove a risk.
-   The same instrument that hedges a position can create one many times larger.
:::
## 6 · Long, short, and offsetting

:::: columns
::: {.column width="50%"}
A **long** position: you own the asset and expect the price to rise.

A **short** position: you do not own it — you borrow it, sell it, and expect to
buy it back cheaper.
:::

::: {.column width="50%"}
To **hedge**:

-   a long position is offset by taking an additional **short** position
-   a short position is offset by taking an additional **long** position
:::
::::

::: keyidea
Hedging is not the absence of a position. It is **two** positions chosen so
that what one loses, the other gains.
:::
## 6 · The instruments

::: {.panel-tabset}

### Forwards

-   An agreement between two parties to transact at a **future point in time**.
-   Specifies the exact assets, the price, and the date of exchange.
-   **Long** = agree to buy later. **Short** = agree to sell later.
-   Private, bilateral, and therefore exposed to the other party defaulting.

### Futures

-   Same idea, but traded on an **organised exchange** rather than privately.
-   **Standardised**, and therefore far more liquid than forwards.
-   **Marked to market daily**, which is what removes most of the default risk.
-   No need to deliver: positions are netted in cash.
-   Harder to corner, because the deliverable range of securities is wide.
-   Regulated in the U.S. by the CFTC.

### Options

-   The **right, not the obligation**, to buy (**call**) or sell (**put**) at
    the strike price.
-   **American** options can be exercised up to expiry; **European** options only
    on the expiry date.
-   You pay a **premium** for that right. The premium rises with
    **time to expiry** and with **volatility**, and moves with the strike.

### Swaps

-   An **interest-rate swap** exchanges one stream of interest payments for
    another, in the same currency.
-   The **plain vanilla** version specifies the rates exchanged, the type of
    payments, and the notional amount.
-   No balance-sheet change, longer-dated than futures or options — but
    **illiquid** and exposed to **default risk**.

### Credit

-   **Credit derivatives** pay off on changes in credit events. The notional
    outstanding is in the trillions.
-   **Credit options** are tied to ratings or spreads.
-   **Credit swaps** exchange payments on loan portfolios, letting lenders
    specialise locally while diversifying.
-   A **credit default swap** pays off option-like when a basket of loans
    defaults. Remember that one.

:::
## 6 · Why the payoff diagram is the whole lesson

```{r}
#| fig-width: 12
#| fig-height: 4.2
S <- seq(105, 130, length.out = 300)
K <- 115; prem <- 2000; mult <- 400
call <- data.frame(S = S, profit = pmax(S - K, 0) * mult - prem,
                   who = "Buyer of call option", panel = "Call option and long future")
futl <- data.frame(S = S, profit = (S - K) * mult,
                   who = "Buyer of futures", panel = "Call option and long future")
put  <- data.frame(S = S, profit = pmax(K - S, 0) * mult - prem,
                   who = "Buyer of put option", panel = "Put option and short future")
futs <- data.frame(S = S, profit = (K - S) * mult,
                   who = "Seller of futures", panel = "Put option and short future")
d <- rbind(call, futl, put, futs)

ggplot(d, aes(S, profit, colour = who)) +
  geom_hline(yintercept = 0, colour = IF_COL$muted, linewidth = .3) +
  geom_line(linewidth = .8) +
  facet_wrap(~panel) +
  scale_colour_manual(values = unname(unlist(
    IF_COL[c("clay", "slate", "clay", "slate")]))) +
  scale_y_continuous(labels = function(x) format(x, big.mark = " ", scientific = FALSE)) +
  labs(x = "Price at expiration", y = "Profit",
       title = "The option has a floor. The future does not.",
       caption = "Strike 115, premium 2 000 per contract.") +
  theme_if() +
  theme(strip.text = element_text(colour = IF_COL$teal700, face = "bold"))
```

::: ask
The option line **bends** and the futures line does not. What did you pay for
that bend, and when is it worth it?
:::
## 6 · Problem 2.8

Rock Solid holds a **$100 million** stock portfolio that tracks the S&P 500.
The manager fears a decline and wants to hedge the downside completely.

::: ask
1.  The S&P is at **1,000**. How is the hedge done with **futures**?
2.  What happens if a year later the index is at **900**?
3.  And if it is at **1,100**?
:::

::: {.smaller}
An index futures contract has a multiplier of **250**, so one contract is worth
$250 \times 1000 = \$250{,}000$.
:::

::: notes
400 contracts sold short. At 900 the portfolio loses 10m and the futures gain
10m. At 1100 the portfolio gains 10m and the futures lose 10m. The hedge works
in both directions — which is the point students miss. They think a hedge
protects the downside and keeps the upside.
:::
## 6 · Problem 2.9

Same portfolio, same fear. Now the S&P is at **1,000** and a strike of **950**
costs a premium of **$200 per contract**.

::: ask
1.  How is the hedge done with **options**?
2.  What happens at **900**?
3.  What happens at **1,100**?
:::

::: {.smaller}
An index option contract has a multiplier of **100**, so one contract is worth
$100 \times 1000 = \$100{,}000$.
:::

. . .

::: keyidea
Compare with 2.8. The option costs money up front and leaves a gap down to 950
uncovered — but at 1,100 you keep the gain. **That is what the premium bought.**
:::
# 7 · Reading discussion {.divider .center background-color="#06302F"}
## Does a green label change the price? {.hook background-color="#FAF7F2"}

A green bond funds an environmental project. Otherwise it is an ordinary bond
from an ordinary issuer, with the same claim on the same balance sheet.

::: incremental
-   So if the cash flows and the credit risk are identical, should it trade at a
    different yield?
-   If it should not, what exactly is the instrument for?
:::
## 7 · Sartzetakis (2021), green bonds and the low-carbon transition

::: {.panel-tabset}

### The questions

Come with an answer to each of these. I will ask.

1.  What are the **advantages and disadvantages of enlarging the mandate of the
    central bank**?
2.  Which projects can be considered **"green"** and financed by green bonds?
3.  What is the process for **project evaluation and selection**?
4.  What are the **positive developments** in the green bond market, and the
    **limits** to its development?

### How to prepare

-   Question 1 is the contested one. A central bank with a climate mandate is
    making distributional choices it was not elected to make — against the
    argument that climate risk is financial risk and therefore already in its
    remit.
-   Questions 2 and 3 are where **greenwashing** lives. If the issuer defines
    "green" and the issuer picks the verifier, what is the label worth?
-   For question 4, separate market **growth** from market **effect**. A market
    can grow quickly and still finance nothing that would not have been financed
    anyway.

:::

::: notes
Run this as a discussion, not a lecture. Ten minutes in pairs on questions 2 and
3, then open the floor on 1. Question 4 is the one that connects back to the
bond pricing at the start of the session: if the green label does not change the
yield, the issuer has no incentive beyond reputation.
:::

-->
