---
title: "International Finance"
subtitle: "Sessions 3-4 — Securities Markets"
author: "Laura Contreras-Portela"
institute: "KEDGE Business School"
date: today
date-format: "MMMM YYYY"
format:
  revealjs:
    theme: [default, ../_theme/theme.scss]
    width: 1600
    height: 900
    title-slide-attributes:
      data-background-color: "#FAF7F2"
    slide-number: c/t
    show-slide-number: all
    progress: true
    history: true
    hash: true
    controls: true
    navigation-mode: linear      # right arrow always goes to the next slide
    preview-links: auto
    highlight-style: github
    code-line-numbers: true
    html-math-method: katex
    chalkboard:                     # C = draw on the slide, B = blank board
      buttons: true
      theme: whiteboard
      chalk-effect: 0
    footer: "International Finance · Sessions 3-4 · KEDGE"
    embed-resources: false          # true -> one portable file, no wifi needed
  beamer:                           # same source -> a real LaTeX PDF handout
    aspectratio: 169
    theme: metropolis
    incremental: false
    header-includes: |
      \definecolor{ifteal}{HTML}{0E5250}
      \definecolor{ifclay}{HTML}{B4573C}
      \setbeamercolor{frametitle}{bg=ifteal}
      \setbeamercolor{progress bar}{fg=ifteal}
      \setbeamercolor{title separator}{fg=ifteal}
      \setbeamercolor{structure}{fg=ifteal}
      \setbeamercolor{alerted text}{fg=ifclay}
execute:
  echo: false
  warning: false
  message: false
  fig-align: center
  fig-width: 8
  fig-height: 4.2
knitr:
  opts_chunk:
    dev: png
    dpi: 300
---

```{r setup}
source("../_R/setup.R")
```

## Where we are

Session 1 and 2 built four prices and one accounting identity. Now we open the
box: **who issues the paper, who buys it, and what happens to the price.**

:::: columns
::: {.column width="50%"}
1. **Money markets** — short, dull, enormous
2. **Bond market** — long-term debt
3. **Stock market** — ownership
:::

::: {.column width="50%"}
4. **Mortgage markets** — debt secured on a house
5. **Foreign exchange** — the price of a currency
6. **Derivatives** — the price of a promise
:::
::::

::: keyidea
Every market in this session answers the same two questions: **what is the
instrument**, and **what makes its price move**. If you can answer those two
for each, you can answer the exam.
:::

::: {.source}
Plus the reading discussion: **Sartzetakis (2021)** on green bonds.
:::

## Why a session on plumbing {.hook background-color="#FAF7F2"}

In August 2007, BNP Paribas froze three funds. The reason it gave was that it
could not value them: the **commercial paper** market had stopped rolling over.

::: incremental
-   Not a stock market crash. A **money market** that quietly refused to
    refinance itself.
-   Within a year the same plumbing took down Lehman, and with it the idea that
    short-term secured funding is always there.
:::

. . .

::: ask
The boring markets are the ones that break the system. Keep that in mind for
the next two sessions.
:::

::: notes
Worth two minutes. Students expect "securities markets" to mean stock-picking.
The point of opening with BNP Paribas is to set the opposite expectation: the
short, dull, enormous markets are where the accidents happen.
:::

# 1 · Money markets {.divider .center background-color="#06302F"}

## Where your company parks its cash {.hook background-color="#FAF7F2"}

A firm collects EUR 40 million on the 25th and pays it out on the 5th. For
eleven days that money cannot sit in a current account earning nothing, and it
cannot be locked up either.

::: incremental
-   That is the whole reason the money market exists: **cash-timing problems**.
-   Inflows and outflows are never synchronised, for corporations or for
    governments.
:::

## 1 · What makes a money market instrument

::: incremental
-   **Short term** — maturity under one year, and for most of it, under 120 days.
-   **Highly liquid** — you can get out before maturity without moving the price.
-   **Low default risk** — you are lending to Treasuries, large banks, big firms.
-   **Large denominations** — USD 1,000,000 and up. This is not a retail market.
:::

. . .

:::: columns
::: {.column width="50%"}
**Investors** want somewhere to put surplus funds for a short period.
:::

::: {.column width="50%"}
**Borrowers** want a low-cost source of temporary funds.
:::
::::

::: keyidea
Both sides are solving a **timing** problem, not an investment problem. Nobody
buys commercial paper because they believe in the company.
:::

## 1 · The instruments

::: {.panel-tabset}

### Treasury bills

-   Maturities from 28 days to 12 months, issued by the U.S. Treasury.
-   Sold at **auction**, which accepts both competitive bids (you name a price)
    and noncompetitive bids (you take the average and are always filled).
-   The benchmark for everything else: as close to default-free as the market gets.

### Fed funds

-   Short-term loans of **reserves** between financial institutions, usually
    **overnight**.
-   Banks use them to meet reserve requirements at the end of the day.
-   The rate on them is the one the central bank actually targets.

### Repos

-   A **repurchase agreement** is a sale with a promise to buy back, usually 3 to
    14 days later, at an agreed price.
-   Economically a secured loan: the securities are the collateral.
-   Like fed funds, except **nonbanks can play** — which is exactly why the repo
    market got so large, and so fragile.

### CDs and CP

-   **Negotiable certificates of deposit** are issued by banks, USD 100,000 to
    USD 10 million, at a discount or at a fixed rate payable at maturity.
-   **Commercial paper** is an *unsecured* promissory note issued by a
    corporation, maturity capped at **270 days** to avoid SEC registration.
-   Unsecured is the word to remember when we get to 2007.

### BAs and eurodollars

-   A **banker's acceptance** is a short-term debt instrument issued by a company
    and **guaranteed by its bank** — built for international trade.
-   The exporter is paid immediately, is shielded from FX risk, and never has to
    assess the importer's balance sheet. The importer's bank does that.
-   **Eurodollars** are dollar-denominated deposits held in banks outside the
    United States. They exist because contracts settle in dollars everywhere.

:::

## 1 · The instruments side by side

| Instrument | Issuer | Buyer | Usual maturity | Secondary market |
|:---|:---|:---|:---|:--:|
| Treasury bills | U.S. government | Consumers and companies | 4, 13, 26 weeks | Excellent |
| Federal funds | Banks | Banks | 1 to 7 days | None |
| Repurchase agreements | Businesses and banks | Businesses and banks | 1 to 15 days | Good |
| Negotiable CDs | Large money-centre banks | Businesses | 14 to 120 days | Good |
| Commercial paper | Finance companies and businesses | Businesses | 1 to 270 days | Poor |
| Banker's acceptance | Banks | Businesses | 30 to 180 days | Good |
| Eurodollar deposits | Non-US banks | Businesses, governments, banks | 1 day to 1 year | Poor |

::: {.source}
Read the last column as a liquidity ranking. Source: Mishkin and Eakins.
:::

## 1 · The rates move together

```{r}
#| fig-width: 12
#| fig-height: 4.4
ids <- c(TB3MS = "3-month Treasury bill",
         FEDFUNDS = "Federal funds",
         MPRIME = "Bank prime rate")
series <- lapply(names(ids), function(id) {
  d <- fred(id)
  if (is.null(d)) return(NULL)
  d$series <- ids[[id]]
  d
})
series <- Filter(Negate(is.null), series)

if (!length(series)) {
  no_data_note("TB3MS / FEDFUNDS / MPRIME")
} else {
  d <- do.call(rbind, series)
  d <- d[d$date >= as.Date("2000-01-01"), ]
  ggplot(d, aes(date, value, colour = series)) +
    recession_bands(from = as.Date("2000-01-01")) +
    geom_line(linewidth = .5) +
    scale_colour_manual(values = unname(unlist(
      IF_COL[c("teal700", "clay", "slate")]))) +
    labs(x = NULL, y = "Percent",
         title = "Money market rates move as one",
         caption = "Shaded: NBER recessions. Source: Federal Reserve via FRED.") +
    theme_if()
}
```

::: ask
They are almost on top of each other. So what is the **spread** between them
telling you in the weeks when they come apart?
:::

::: notes
The answer is credit and liquidity risk. In a calm market the spread is a few
basis points; in August 2007 and September 2008 it explodes. If the chart has
downloaded, zoom in on 2008 with the chalkboard.
:::

## 1 · Commercial paper outstanding

```{r}
#| fig-width: 12
#| fig-height: 4.2
cp <- fred("COMPOUT")
if (is.null(cp)) {
  no_data_note("COMPOUT")
} else {
  cp$value <- cp$value / 1000
  ggplot(cp, aes(date, value)) +
    recession_bands(from = min(cp$date)) +
    geom_line(colour = IF_COL$teal700, linewidth = .55) +
    labs(x = NULL, y = "USD trillion",
         title = "Commercial paper outstanding",
         caption = "Shaded: NBER recessions. Source: Federal Reserve Board via FRED.") +
    theme_if()
}
```

::: ask
Find 2007. The market did not decline — it **fell off a cliff** and never came
back to the old level. Who was funding themselves there, and where did they go
instead?
:::

## 1 · Price and rate

A money market instrument pays no coupon. You buy it below face value and it
matures at face value, so the whole return is in the **discount**.

:::: columns
::: {.column width="50%"}
The **discount rate** — how far below face value you paid:

$$i_{discount} = \frac{F-P}{F} \times \frac{360}{n}$$
:::

::: {.column width="50%"}
The **annualised yield** — what you actually earned:

$$i_{yt} = \frac{F-P}{P} \times \frac{365}{n}$$
:::
::::

. . .

::: keyidea
Same instrument, two numbers. The discount rate divides by **F** and uses a
360-day year; the yield divides by **P** and uses 365. The yield is always the
larger of the two, and it is the one you can compare with anything else.
:::

## 1 · The two rates, drawn

```{r}
#| fig-width: 10
#| fig-height: 4.0
P <- seq(97, 100, length.out = 200)
F <- 100; n <- 90
d <- rbind(
  data.frame(P = P, rate = (F - P) / F * 360 / n, which = "Discount rate"),
  data.frame(P = P, rate = (F - P) / P * 365 / n, which = "Annualised yield")
)
ggplot(d, aes(P, rate, colour = which)) +
  geom_line(linewidth = .8) +
  scale_colour_manual(values = c(IF_COL$clay, IF_COL$slate)) +
  scale_y_continuous(labels = function(x) paste0(round(x * 100), "%")) +
  labs(x = "Current price", y = "Rate",
       title = "F = 100, n = 90 days",
       caption = "The gap between the two is the 360 / 365 convention plus the choice of denominator.") +
  theme_if()
```

## 1 · Problem 2.1

::: ask
A 28-day T-bill costs **$996.73** and is worth **$1,000** at maturity.

1.  What is the **discount rate**?
2.  What is the **annualised yield**?
:::

. . .

:::: columns
::: {.column width="50%"}
$$i_{discount} = \frac{1000-996.73}{1000}\times\frac{360}{28} = 4.20\%$$
:::

::: {.column width="50%"}
$$i_{yt} = \frac{1000-996.73}{996.73}\times\frac{365}{28} = 4.28\%$$
:::
::::

::: notes
Make them do it on the calculator, not in their heads. The point of the exercise
is that the two answers differ by 8 basis points on an instrument that everyone
would describe as "the same rate".
:::

## 1 · Problem 2.2

The Treasury auctions **$2.5 billion** of 91-day bills. Competitive bids:

| Bidder | Bid amount | Bid price |
|:--:|:--:|:--:|
| 1 | $500 million | $0.9940 |
| 2 | $750 million | $0.9901 |
| 3 | $1.5 billion | $0.9925 |
| 4 | $1 billion | $0.9936 |
| 5 | $600 million | $0.9939 |

It also received **$750 million** of noncompetitive bids.

::: ask
Who receives bills, how many, and at what price?
:::

::: notes
Order by price descending: 1 (0.9940), 5 (0.9939), 4 (0.9936), 3 (0.9925),
2 (0.9901). Noncompetitive bids are filled first and come off the top: 2.5 - 0.75
= 1.75bn left for competitive bidders. Work down the list until it is exhausted.
Bidder 3 is partially filled and bidder 2 gets nothing.
:::

# 2 · The bond market {.divider .center background-color="#06302F"}

## A promise with a date on it {.hook background-color="#FAF7F2"}

In 2012 Greece restructured EUR 200 billion of bonds. Holders took a haircut of
roughly 53% on face value. The bonds had been rated investment grade a few
years earlier.

::: incremental
-   A bond is a **contract**: specified payments on specified dates.
-   Nothing about that contract guarantees you will be paid.
:::

## 2 · The vocabulary

| Term | What it means |
|:---|:---|
| Coupon interest rate | The annual interest rate, usually fixed for the whole life |
| Current yield | Coupon payment divided by the current price |
| Face amount / par value | The maturity value |
| Indenture | The contract that specifies the terms of the loan agreement |
| Market rate | The rate currently in effect on securities of like risk and maturity |
| Maturity | Number of periods until the security matures |
| Yield to maturity | The yield if you buy at the current price and hold to maturity |

::: keyidea
**Coupon rate is fixed. Market rate moves.** Everything in bond pricing follows
from the gap between those two.
:::

## 2 · The instruments

::: {.panel-tabset}

### Treasuries

-   **Treasury notes**: maturity 1 to 10 years. **Treasury bonds**: 10 to 30 years.
-   No default risk in the usual sense — the Treasury issues the currency it
    owes. **Inflation risk is still there**, and so is the risk of a price fall
    if rates rise.

### Municipals

-   Issued by local, county and state governments to fund public projects.
-   **Tax free**, which is the whole point. Compare like with like:

    $$\text{after-tax rate} = \text{taxable rate}\times(1-\text{marginal tax rate})$$

-   **General obligation** bonds are repaid from taxes; **revenue** bonds from
    user fees on the project itself.
-   They are **not** default free. Detroit, 2013.

### Corporates

-   Face value usually USD 1,000, interest paid semi-annually.
-   **Restrictive covenants** limit dividends, new debt, ratios — they exist
    because shareholder and bondholder interests conflict.
-   A **call provision** lets the issuer buy the bond back. A **conversion**
    feature lets the holder turn it into equity.
-   **Secured** bondholders have first claim on a specific asset; **unsecured**
    ones do not, and charge more for it.

### Ratings

-   Risk ranges from **Aaa/AAA** down. Below **BBB-/Baa3** a bond is
    **speculative grade** — "junk".
-   That line matters more than it should: many trusts and insurance companies
    are *not permitted* to hold debt below it.
-   Which means a downgrade across the line forces selling by people who have no
    view on the credit at all.

:::

## 2 · The rating scales

:::: columns
::: {.column width="52%"}
| Moody's | S&P | Fitch | Meaning |
|:--:|:--:|:--:|:---|
| Aaa | AAA | AAA | Best quality, least credit risk |
| Aa1-Aa3 | AA+/AA/AA- | AA+/AA/AA- | High quality, very strong capacity |
| A1-A3 | A+/A/A- | A+/A/A- | Upper-medium grade |
| Baa1-Baa3 | BBB+/BBB/BBB- | BBB+/BBB/BBB- | Medium quality, adequate capacity |
:::

::: {.column width="48%"}
| Moody's | S&P | Fitch | Meaning |
|:--:|:--:|:--:|:---|
| Ba1-Ba3 | BB+/BB/BB- | BB+/BB/BB- | Speculative |
| B1-B3 | B+/B/B- | B+/B/B- | Very speculative |
| Caa-C | CCC/CC/C | CCC/CC/C | Highly speculative, near or in default |
| — | D | DDD/DD/D | In default |
:::
::::

::: ask
The **investment grade** line sits between Baa3 and Ba1. Why should a single
notch matter so much to the price?
:::

## 2 · Rates by credit quality

```{r}
#| fig-width: 12
#| fig-height: 4.4
ids <- c(AAA = "Corporate Aaa", BAA = "Corporate Baa", GS20 = "20-year Treasury")
series <- lapply(names(ids), function(id) {
  d <- fred(id)
  if (is.null(d)) return(NULL)
  d$series <- ids[[id]]
  d
})
series <- Filter(Negate(is.null), series)

if (!length(series)) {
  no_data_note("AAA / BAA / GS20")
} else {
  d <- do.call(rbind, series)
  d <- d[d$date >= as.Date("1980-01-01"), ]
  ggplot(d, aes(date, value, colour = series)) +
    recession_bands(from = as.Date("1980-01-01")) +
    geom_line(linewidth = .5) +
    scale_colour_manual(values = unname(unlist(
      IF_COL[c("slate", "clay", "teal700")]))) +
    labs(x = NULL, y = "Percent",
         title = "The credit spread is the distance between these lines",
         caption = "Shaded: NBER recessions. Source: Moody's and the Treasury via FRED.") +
    theme_if()
}
```

::: ask
The lines are furthest apart exactly when you would most like to borrow. Why?
:::

## 2 · Price and yield

The price of a bond is the present value of everything it will pay you:

$$P = \frac{C}{(1+r)} + \frac{C}{(1+r)^2} + \dots + \frac{C}{(1+r)^n} + \frac{F}{(1+r)^n}$$

. . .

The **yield to maturity** is the $r$ that makes that equation hold at the price
you actually see:

$$0 = -P + \sum_{t=1}^{n}\frac{C}{(1+r)^t} + \frac{F}{(1+r)^n}$$

. . .

The **coupon yield** is a shortcut — the coupon over the market price:

$$r_c = \frac{C}{P}$$

::: keyidea
Price and yield move in **opposite directions**, always. The coupon never
changes; the only way a fixed payment can offer a higher return is if you pay
less for it.
:::

## 2 · Coupon yield is only an approximation

```{r}
#| fig-width: 10
#| fig-height: 4.0
Pr <- seq(50, 150, length.out = 300)
F <- 100; C <- 5; n <- 10
ytm <- sapply(Pr, function(p) {
  f <- function(r) -p + sum(C / (1 + r)^(1:n)) + F / (1 + r)^n
  tryCatch(uniroot(f, c(-0.5, 1))$root, error = function(e) NA_real_)
})
d <- rbind(
  data.frame(P = Pr, rate = ytm,    which = "Yield to maturity"),
  data.frame(P = Pr, rate = C / Pr, which = "Coupon yield")
)
ggplot(d, aes(P, rate, colour = which)) +
  geom_hline(yintercept = 0, colour = IF_COL$muted, linewidth = .3) +
  geom_vline(xintercept = 100, linetype = "dotted", colour = IF_COL$muted) +
  geom_line(linewidth = .8) +
  scale_colour_manual(values = c(IF_COL$clay, IF_COL$slate)) +
  scale_y_continuous(labels = function(x) paste0(round(x * 100), "%")) +
  labs(x = "Current price", y = NULL,
       title = "F = 100, coupon 5, 10 years to maturity",
       caption = "The two agree only at par. Away from par the coupon yield misleads.") +
  theme_if()
```

## 2 · Problem 2.3

::: ask
1.  A corporate bond yields **9%**, a municipal bond **6.75%**. Which do you buy?
2.  A corporate bond yields **5%**, a municipal **3.5%**, and your marginal tax
    rate is **28%**. Which now?
:::

. . .

:::: columns
::: {.column width="50%"}
**First case.** The comparison needs a tax rate to be meaningful. At a marginal
rate of 25%, the corporate is worth $9\% \times 0.75 = 6.75\%$ after tax —
exactly the municipal. Above 25% the municipal wins.
:::

::: {.column width="50%"}
**Second case.** $5\% \times (1-0.28) = 3.60\%$ after tax, against **3.5%**
tax-free. The corporate wins, narrowly.
:::
::::

::: notes
The first part has no numerical answer until you ask them for the tax rate.
That is deliberate: it is the question they should have asked. Let someone try
to answer before pointing it out.
:::

## 2 · Problem 2.4

::: ask
1.  What is the **coupon yield** on a bond with face value $1,000, current
    price $921.01 and a coupon rate of **10.95%**?
2.  What is the **price** of a two-year, 10% coupon bond paying **semi-annually**,
    face value $1,000, if the required return is **12%**?
:::

. . .

:::: columns
::: {.column width="46%"}
$$r_c = \frac{109.50}{921.01} = 11.89\%$$
:::

::: {.column width="54%"}
Four periods, coupon 50, discount rate 6% per period:

$$P = \sum_{t=1}^{4}\frac{50}{1.06^t} + \frac{1000}{1.06^4} = 965.35$$
:::
::::

::: notes
The semi-annual part is where they lose marks: halve the coupon AND halve the
rate AND double the number of periods. Three changes, not one.
:::

# 3 · The stock market {.divider .center background-color="#06302F"}

## Owning a piece of the argument {.hook background-color="#FAF7F2"}

A bondholder is owed a number. A shareholder owns whatever is left after
everyone else is paid — which can be everything, or nothing.

::: incremental
-   Stocks earn a return two ways: **price appreciation** and **dividends**.
-   Shareholders have a **residual** claim on assets, and a vote.
:::

## 3 · Common, preferred, and foreign

:::: columns
::: {.column width="50%"}
**Common stock** — the right to vote for directors and on certain issues, and
to receive dividends if the board declares them.

**Preferred stock** — usually **no vote**, but a **fixed dividend** paid before
common. Closer to a bond than most students expect.
:::

::: {.column width="50%"}
**Foreign stocks** are useful for diversification but awkward to buy if they are
not listed locally.

**American depository receipts (ADRs)** solve it: a bank buys the foreign
shares and issues receipts against them that trade on U.S. exchanges.
:::
::::

::: keyidea
An **index** — DJIA, S&P 500, NASDAQ composite — is a summary of a group of
stocks, not an investment in itself. How it is *weighted* changes what it tells you.
:::

## 3 · Where the trading happens

:::: columns
::: {.column width="50%"}
**Organised exchanges**

-   NYSE trades roughly 4 to 7 billion shares a day.
-   "Organised" once meant a physical floor; electronic networks have largely
    replaced that idea.
-   **Listing requirements** exclude small firms.
-   Auction market with floor specialists; around 25% of trades filled directly
    by the specialist.
:::

::: {.column width="50%"}
**Over the counter**

-   NASDAQ is the obvious example, with roughly 3,000 securities listed.
-   **Dealers** stand ready to make a market — without one, the equity would be
    hard to trade at all.
-   Multiple market makers post bid and ask prices for the same security.
:::
::::

::: ask
If several dealers quote the same stock, what stops one of them quoting a
terrible price?
:::

## 3 · Price-weighted versus value-weighted

The **DJIA** is a price-weighted average of 30 stocks. The **S&P 500** is a
market-value-weighted index of 500.

```{r}
#| fig-width: 12
#| fig-height: 3.9
dj <- fred("DJIA")
sp <- fred("SP500")
if (is.null(dj)) {
  no_data_note("DJIA")
} else {
  dj <- dj[dj$date >= as.Date("2015-01-01"), ]
  dj$index <- 100 * dj$value / dj$value[1]
  dj$series <- "Dow Jones Industrial Average"
  d <- dj[, c("date", "index", "series")]
  if (!is.null(sp)) {
    sp <- sp[sp$date >= as.Date("2015-01-01"), ]
    sp$index <- 100 * sp$value / sp$value[1]
    sp$series <- "S&P 500"
    d <- rbind(d, sp[, c("date", "index", "series")])
  }
  ggplot(d, aes(date, index, colour = series)) +
    geom_line(linewidth = .5) +
    scale_colour_manual(values = c(IF_COL$slate, IF_COL$teal500)) +
    labs(x = NULL, y = "Index, first observation = 100",
         title = "Thirty stocks by price, or five hundred by market value",
         caption = "Source: S&P Dow Jones Indices via FRED.") +
    theme_if()
}
```

::: ask
A price-weighted index gives a $400 stock thirteen times the weight of a $30
stock, whatever the companies are worth. Is that a sensible way to summarise a
market?
:::

## 3 · What a share is worth

The **Gordon growth model** — a share is the present value of its dividends,
growing forever at $g$:

$$P = \sum_{t=1}^{\infty}\frac{Div_t}{(1+k_e)^t} = \frac{Div_1}{k_e - g}$$

:::: columns
::: {.column width="50%"}
**Geometric growth**, for backing out $g$ from two observations:

$$g = \left(\frac{\text{final}}{\text{initial}}\right)^{1/n} - 1$$
:::

::: {.column width="50%"}
**Price-to-earnings**, the shortcut everyone actually quotes:

$$PE = \frac{\text{Price}}{\text{Earnings}}$$
:::
::::

::: keyidea
Look at the denominator: $k_e - g$. As growth approaches the required return,
the price goes to **infinity**. Every valuation argument about a fast-growing
company is a fight over that subtraction.
:::

## 3 · The denominator does all the work

```{r}
#| fig-width: 12
#| fig-height: 4.0
div1 <- 2
a <- data.frame(x = seq(.06, .24, length.out = 200))
a$price <- div1 / (a$x - .05); a$panel <- "Varying the required return (g = 5%)"
b <- data.frame(x = seq(.01, .19, length.out = 200))
b$price <- div1 / (.20 - b$x); b$panel <- "Varying the growth rate (k = 20%)"
d <- rbind(a, b)
ggplot(d, aes(x, price)) +
  geom_line(colour = IF_COL$clay, linewidth = .8) +
  facet_wrap(~panel, scales = "free_x") +
  scale_x_continuous(labels = function(x) paste0(round(x * 100), "%")) +
  coord_cartesian(ylim = c(0, 250)) +
  labs(x = NULL, y = "Stock price",
       title = "Gordon growth, with the first dividend fixed at 2",
       caption = "Both curves are the same asymptote seen from two directions.") +
  theme_if() +
  theme(strip.text = element_text(colour = IF_COL$teal700, face = "bold"))
```

## 3 · Problem 2.5

::: ask
1.  A stock is expected to pay a dividend of **$0.16** and to be worth **$60**
    in a year. At a **12%** discount rate, what is it worth now?
2.  If the industry PE ratio is **16** and the firm earns **$1.13 per share**,
    what is the price?
:::

. . .

:::: columns
::: {.column width="50%"}
$$P_0 = \frac{0.16 + 60}{1.12} = 53.71$$
:::

::: {.column width="50%"}
$$P = 16 \times 1.13 = 18.08$$
:::
::::

## 3 · Problem 2.6

::: ask
The DJIA was **10,000** in 2010 and **20,000** in 2017. What was the annual
**geometric growth rate**?
:::

. . .

$$g = \left(\frac{20000}{10000}\right)^{1/7} - 1 = 2^{1/7} - 1 = 10.41\%$$

::: notes
Watch for students using 8 years instead of 7. The exponent is the number of
*periods between* the observations, not the number of years mentioned.
:::

# 4 · Mortgage markets {.divider .center background-color="#06302F"}

## The market that ate the world {.hook background-color="#FAF7F2"}

By 2006 an American mortgage was not a loan a bank held. It was an input:
pooled, tranched, rated, and sold to a German Landesbank that had never seen
Ohio.

::: incremental
-   A mortgage is a **long-term loan secured by real estate**.
-   It is **amortised**: one fixed payment covers both interest and principal
    every month.
:::

## 4 · The types

| Type | What it is |
|:---|:---|
| Conventional | Not guaranteed; usually needs private mortgage insurance and 5-20% down |
| Insured | Guaranteed by a government agency; low or zero down payment |
| Fixed-rate | The rate is fixed for the life of the mortgage |
| Adjustable-rate (ARM) | Rate tied to another security, adjusted periodically within annual limits |
| Graduated-payment (GPM) | Low initial payment rising each year; amortises in 30 years |
| Growing-equity (GEM) | Payment rises each year; amortises in less than 30 years |
| Second mortgage | Secured by a second lien; used for credit lines and home improvement |
| Reverse annuity (RAM) | The lender pays the borrower monthly; due when the property is sold |

::: ask
Which of these transfers **interest rate risk** from the lender to the borrower?
And who tends to be sold those?
:::

## 4 · Securitisation, in four steps

::: incremental
1.  The **secondary mortgage market** was created by the federal government
    after WWII, when it set up Fannie Mae. It grew quickly from the mid-1980s.
2.  Individual mortgages default unpredictably, so they are **pooled** —
    hundreds at a time — on the argument that diversification tames the risk.
3.  Rights to the pool's cash flows are sold as separate securities:
    **pass-throughs**, or mortgage-backed securities.
4.  The borrower's monthly payment passes through a trustee before reaching the
    investors.
:::

. . .

::: ask
Step 2 is the load-bearing one. Diversification removes risk that is
**independent** across loans. What happens to a pool of American mortgages if
house prices fall everywhere at once?
:::

::: notes
This is the hinge of the whole session and it pays to slow down. The 2008
answer: the correlation everyone assumed was near zero turned out to be near
one, and every tranche above the bottom was priced on that assumption.
:::

## 4 · Mortgage rates track long Treasuries

```{r}
#| fig-width: 12
#| fig-height: 4.4
ids <- c(MORTGAGE30US = "30-year fixed mortgage", GS10 = "10-year Treasury")
series <- lapply(names(ids), function(id) {
  d <- fred(id)
  if (is.null(d)) return(NULL)
  d$series <- ids[[id]]
  d
})
series <- Filter(Negate(is.null), series)

if (!length(series)) {
  no_data_note("MORTGAGE30US / GS10")
} else {
  d <- do.call(rbind, series)
  ggplot(d, aes(date, value, colour = series)) +
    recession_bands(from = min(d$date)) +
    geom_line(linewidth = .5) +
    scale_colour_manual(values = c(IF_COL$slate, IF_COL$clay)) +
    labs(x = NULL, y = "Percent",
         title = "The mortgage rate is a spread over the long Treasury",
         caption = "Shaded: NBER recessions. Source: Freddie Mac and the Treasury via FRED.") +
    theme_if()
}
```

::: ask
The gap between the two is not constant. What widens it?
:::

## 4 · Loan and annuity

:::: columns
::: {.column width="50%"}
What you can **borrow**, given what you can pay each period:

$$C_0 = a \times \left[\frac{1-(1+i)^{-T}}{i}\right]$$
:::

::: {.column width="50%"}
What you must **pay** each period, given what you borrowed:

$$a = C_0 \times \left[\frac{i}{1-(1+i)^{-T}}\right]$$
:::
::::

```{r}
#| fig-width: 10
#| fig-height: 3.3
i <- 0.05 / 12; T <- 20 * 12
d <- data.frame(a = seq(1000, 3000, by = 25))
d$loan <- d$a * (1 - (1 + i)^(-T)) / i
ggplot(d, aes(a, loan)) +
  geom_line(colour = IF_COL$slate, linewidth = .8) +
  scale_y_continuous(labels = function(x) format(x, big.mark = " ", scientific = FALSE)) +
  labs(x = "Monthly payment", y = "Loan you can support",
       title = "5% nominal, 20 years: i = 0.417% per month, T = 240 months",
       caption = "Linear in the payment, which is why affordability is really about the rate.") +
  theme_if()
```

## 4 · Problem 2.7

::: ask
You want to borrow **$100,000**. Choose between a **12% 30-year** mortgage and
an **11.5%** mortgage with **2 discount points**. Which?
:::

::: {.smaller}
A discount point is 1% of the loan paid up front: on a $200,000 loan, 1 point
costs $2,000, half a point $1,000, a quarter point $500.
:::

. . .

::: keyidea
Two points on $100,000 is $2,000 today against 50 basis points saved every month
for thirty years. Compute both monthly payments with the annuity formula, then
ask how long it takes the saving to repay the $2,000. **The answer depends
entirely on how long you stay in the house.**
:::

::: notes
This is the one problem in the session with a genuinely conditional answer.
If they move in four years the points are a bad deal; if they hold to maturity
they are a good one. Getting them to state the break-even horizon is the goal.
:::

# 5 · The foreign exchange market {.divider .center background-color="#06302F"}

## The biggest market nobody sees {.hook background-color="#FAF7F2"}

No building, no opening bell, no ticker on the news. Over-the-counter, bank to
bank, and turnover in the **trillions of dollars a day**.

::: incremental
-   What is traded is **bank deposits denominated in different currencies**.
-   Trades are large — typically in excess of USD 1 million.
-   We met the price in session 2. Now we meet the market.
:::

## 5 · Purchasing power parity

**Absolute PPP** — the exchange rate equals the ratio of price levels:

$$E_{\$/\mathrm{EUR}} = \frac{P_{US}}{P_{EUR}}$$

. . .

**Relative PPP** — the rate of depreciation equals the **inflation differential**:

$$\frac{\Delta E_{\$/\mathrm{EUR}}}{E_{\$/\mathrm{EUR}}} = \pi_{US} - \pi_{EUR}$$

. . .

::: ask
PPP works in the long run and fails in the short run. Given what it assumes,
which of its assumptions is the one that fails first?
:::

::: notes
Answers to steer towards: differentiated goods with brand names and legal
protection give firms pricing power, so identical goods do not have identical
prices; and prices are sticky, so they do not adjust quickly to market
conditions. Both break the arbitrage that PPP relies on.
:::

## 5 · Spot, forward, and what moves them

:::: columns
::: {.column width="46%"}
**Spot rate** — near-immediate exchange of bank deposits.

**Forward rate** — exchange at an agreed future date, at a price agreed now.
:::

::: {.column width="54%"}
Rates move when money demand or supply moves, which happens on expected changes in:

-   **relative price levels**
-   **trade barriers** — tariffs and quotas
-   **preferences** for domestic versus foreign goods
-   **relative productivity**
:::
::::

. . .

::: keyidea
Every item on that list is an **expected change**. The spot rate today is
mostly a statement about tomorrow.
:::

## 5 · What an appreciation does

::: incremental
-   Domestic goods become **more expensive abroad**; foreign goods become
    cheaper at home.
-   Domestic businesses become **less competitive**.
-   Domestic consumers gain **purchasing power**.
:::

. . .

::: ask
So is a strong currency good news? Ask a French exporter and a French
holidaymaker on the same day and you get two answers.
:::

# 6 · Derivative markets {.divider .center background-color="#06302F"}

## A price for a promise {.hook background-color="#FAF7F2"}

In 1995 Barings Bank, 233 years old, was destroyed in weeks by one trader's
position in Nikkei futures. The instruments were not exotic. The position was
simply unhedged and unsupervised.

::: incremental
-   **Hedging** means using a financial transaction to reduce or remove a risk.
-   The same instrument that hedges a position can create one many times larger.
:::

## 6 · Long, short, and offsetting

:::: columns
::: {.column width="50%"}
A **long** position: you own the asset and expect the price to rise.

A **short** position: you do not own it — you borrow it, sell it, and expect to
buy it back cheaper.
:::

::: {.column width="50%"}
To **hedge**:

-   a long position is offset by taking an additional **short** position
-   a short position is offset by taking an additional **long** position
:::
::::

::: keyidea
Hedging is not the absence of a position. It is **two** positions chosen so
that what one loses, the other gains.
:::

## 6 · The instruments

::: {.panel-tabset}

### Forwards

-   An agreement between two parties to transact at a **future point in time**.
-   Specifies the exact assets, the price, and the date of exchange.
-   **Long** = agree to buy later. **Short** = agree to sell later.
-   Private, bilateral, and therefore exposed to the other party defaulting.

### Futures

-   Same idea, but traded on an **organised exchange** rather than privately.
-   **Standardised**, and therefore far more liquid than forwards.
-   **Marked to market daily**, which is what removes most of the default risk.
-   No need to deliver: positions are netted in cash.
-   Harder to corner, because the deliverable range of securities is wide.
-   Regulated in the U.S. by the CFTC.

### Options

-   The **right, not the obligation**, to buy (**call**) or sell (**put**) at
    the strike price.
-   **American** options can be exercised up to expiry; **European** options only
    on the expiry date.
-   You pay a **premium** for that right. The premium rises with
    **time to expiry** and with **volatility**, and moves with the strike.

### Swaps

-   An **interest-rate swap** exchanges one stream of interest payments for
    another, in the same currency.
-   The **plain vanilla** version specifies the rates exchanged, the type of
    payments, and the notional amount.
-   No balance-sheet change, longer-dated than futures or options — but
    **illiquid** and exposed to **default risk**.

### Credit

-   **Credit derivatives** pay off on changes in credit events. The notional
    outstanding is in the trillions.
-   **Credit options** are tied to ratings or spreads.
-   **Credit swaps** exchange payments on loan portfolios, letting lenders
    specialise locally while diversifying.
-   A **credit default swap** pays off option-like when a basket of loans
    defaults. Remember that one.

:::

## 6 · Why the payoff diagram is the whole lesson

```{r}
#| fig-width: 12
#| fig-height: 4.2
S <- seq(105, 130, length.out = 300)
K <- 115; prem <- 2000; mult <- 400
call <- data.frame(S = S, profit = pmax(S - K, 0) * mult - prem,
                   who = "Buyer of call option", panel = "Call option and long future")
futl <- data.frame(S = S, profit = (S - K) * mult,
                   who = "Buyer of futures", panel = "Call option and long future")
put  <- data.frame(S = S, profit = pmax(K - S, 0) * mult - prem,
                   who = "Buyer of put option", panel = "Put option and short future")
futs <- data.frame(S = S, profit = (K - S) * mult,
                   who = "Seller of futures", panel = "Put option and short future")
d <- rbind(call, futl, put, futs)

ggplot(d, aes(S, profit, colour = who)) +
  geom_hline(yintercept = 0, colour = IF_COL$muted, linewidth = .3) +
  geom_line(linewidth = .8) +
  facet_wrap(~panel) +
  scale_colour_manual(values = unname(unlist(
    IF_COL[c("clay", "slate", "clay", "slate")]))) +
  scale_y_continuous(labels = function(x) format(x, big.mark = " ", scientific = FALSE)) +
  labs(x = "Price at expiration", y = "Profit",
       title = "The option has a floor. The future does not.",
       caption = "Strike 115, premium 2 000 per contract.") +
  theme_if() +
  theme(strip.text = element_text(colour = IF_COL$teal700, face = "bold"))
```

::: ask
The option line **bends** and the futures line does not. What did you pay for
that bend, and when is it worth it?
:::

## 6 · Problem 2.8

Rock Solid holds a **$100 million** stock portfolio that tracks the S&P 500.
The manager fears a decline and wants to hedge the downside completely.

::: ask
1.  The S&P is at **1,000**. How is the hedge done with **futures**?
2.  What happens if a year later the index is at **900**?
3.  And if it is at **1,100**?
:::

::: {.smaller}
An index futures contract has a multiplier of **250**, so one contract is worth
$250 \times 1000 = \$250{,}000$.
:::

::: notes
400 contracts sold short. At 900 the portfolio loses 10m and the futures gain
10m. At 1100 the portfolio gains 10m and the futures lose 10m. The hedge works
in both directions — which is the point students miss. They think a hedge
protects the downside and keeps the upside.
:::

## 6 · Problem 2.9

Same portfolio, same fear. Now the S&P is at **1,000** and a strike of **950**
costs a premium of **$200 per contract**.

::: ask
1.  How is the hedge done with **options**?
2.  What happens at **900**?
3.  What happens at **1,100**?
:::

::: {.smaller}
An index option contract has a multiplier of **100**, so one contract is worth
$100 \times 1000 = \$100{,}000$.
:::

. . .

::: keyidea
Compare with 2.8. The option costs money up front and leaves a gap down to 950
uncovered — but at 1,100 you keep the gain. **That is what the premium bought.**
:::

# 7 · Reading discussion {.divider .center background-color="#06302F"}

## Does a green label change the price? {.hook background-color="#FAF7F2"}

A green bond funds an environmental project. Otherwise it is an ordinary bond
from an ordinary issuer, with the same claim on the same balance sheet.

::: incremental
-   So if the cash flows and the credit risk are identical, should it trade at a
    different yield?
-   If it should not, what exactly is the instrument for?
:::

## 7 · Sartzetakis (2021), green bonds and the low-carbon transition

::: {.panel-tabset}

### The questions

Come with an answer to each of these. I will ask.

1.  What are the **advantages and disadvantages of enlarging the mandate of the
    central bank**?
2.  Which projects can be considered **"green"** and financed by green bonds?
3.  What is the process for **project evaluation and selection**?
4.  What are the **positive developments** in the green bond market, and the
    **limits** to its development?

### How to prepare

-   Question 1 is the contested one. A central bank with a climate mandate is
    making distributional choices it was not elected to make — against the
    argument that climate risk is financial risk and therefore already in its
    remit.
-   Questions 2 and 3 are where **greenwashing** lives. If the issuer defines
    "green" and the issuer picks the verifier, what is the label worth?
-   For question 4, separate market **growth** from market **effect**. A market
    can grow quickly and still finance nothing that would not have been financed
    anyway.

:::

::: notes
Run this as a discussion, not a lecture. Ten minutes in pairs on questions 2 and
3, then open the floor on 1. Question 4 is the one that connects back to the
bond pricing at the start of the session: if the green label does not change the
yield, the issuer has no incentive beyond reputation.
:::

## Next session {.center}

Sessions 5 and 6 move to **financial institutions and firms**.

::: keyidea
Bring **Muñoz (2020)** on boards of trustees and the performance of socially
responsible mutual funds, read — and keep working on your group's security.
:::

::: {.source}
Press **`O`** for the slide overview · **`C`** to draw · **`S`** for speaker notes · **`F`** for fullscreen · **`E`** then print for a PDF of the slides.
:::
