---
title: "International Finance"
subtitle: "Session 3 — Securities Markets: Money, Bonds and Stocks"
author: "Laura Contreras-Portela"
institute: "KEDGE Business School"
date: today
date-format: "MMMM YYYY"
format:
  revealjs:
    theme: [default, ../_theme/theme.scss]
    width: 1600
    height: 900
    title-slide-attributes:
      data-background-color: "#FAF7F2"
    slide-number: c/t
    show-slide-number: all
    progress: true
    history: true
    hash: true
    controls: true
    navigation-mode: linear      # right arrow always goes to the next slide
    preview-links: auto
    highlight-style: github
    code-line-numbers: true
    html-math-method: katex
    chalkboard:                     # C = draw on the slide, B = blank board
      buttons: true
      theme: whiteboard
      chalk-effect: 0
    footer: "International Finance · Session 3 · KEDGE"
    embed-resources: false          # true -> one portable file, no wifi needed
  beamer:                           # same source -> a real LaTeX PDF handout
    aspectratio: 169
    theme: metropolis
    incremental: false
    header-includes: |
      \definecolor{ifteal}{HTML}{0E5250}
      \definecolor{ifclay}{HTML}{B4573C}
      \setbeamercolor{frametitle}{bg=ifteal}
      \setbeamercolor{progress bar}{fg=ifteal}
      \setbeamercolor{title separator}{fg=ifteal}
      \setbeamercolor{structure}{fg=ifteal}
      \setbeamercolor{alerted text}{fg=ifclay}
execute:
  echo: false
  warning: false
  message: false
  fig-align: center
  fig-width: 8
  fig-height: 4.2
knitr:
  opts_chunk:
    dev: png
    dpi: 300
---

```{r setup}
source("../_R/setup.R")
```

## Where we are

Sessions 1 and 2 established four prices and the balance-of-payments identity.
We now disaggregate the financial system by **instrument**, and ask how each
market transfers funds from surplus units to deficit units.

:::: columns
::: {.column width="50%"}
1. **Money markets** — short maturity, high liquidity
2. **Bond market** — long-term debt claims
3. **Stock market** — residual ownership claims
:::

::: {.column width="50%"}
4. **Mortgage markets** — debt secured on real property
5. **Foreign exchange** — claims in different currencies
6. **Derivatives** — claims contingent on other claims
:::
::::

::: keyidea
Each market is analysed through the same two questions: **what contractual
claim does the instrument represent**, and **what determines its price**. These
are the organising questions of the course, and of the examination.
:::

::: {.source}
Plus the reading discussion: **Sartzetakis (2021)** on green bonds.
:::

::: notes
Ninety seconds. The point of this slide is to set the promise: two questions per
market, repeated six times. Tell them explicitly that the exam asks those two
questions, because it makes the rest of the session feel like a checklist they
are filling in rather than a list of definitions to memorise.

Today covers the first three. Mortgages, FX and derivatives are next session.
:::

## 9 August 2007: a funding market stops {.hook background-color="#FAF7F2"}

BNP Paribas suspended redemptions on three funds, citing an evaporation of
liquidity that made it impossible to value the assets they held. Those assets
were **asset-backed commercial paper**, and the market for them had ceased to
refinance itself.

::: incremental
-   This was not an equity market event. It was a **money market** — the
    shortest, safest, largest segment of the system — that stopped rolling over.
-   The underlying practice is **maturity transformation**: funding long-dated
    assets with short-dated liabilities. It is profitable precisely because it
    is risky, and the risk is **rollover risk**.
-   The same mechanism propagated through Northern Rock, Bear Stearns and
    Lehman Brothers within thirteen months.
:::

. . .

::: ask
Market microstructure is usually treated as detail. Under what conditions does
it become the **binding constraint** on the whole financial system?
:::

::: notes
Worth two minutes. Students expect "securities markets" to mean stock-picking.
The point of opening with BNP Paribas is to set the opposite expectation: the
short, dull, enormous markets are where the accidents happen.
:::

# 1 · Money markets {.divider .center background-color="#06302F"}

## The corporate cash-timing problem {.hook background-color="#FAF7F2"}

A firm receives EUR 40 million on the 25th of the month and disburses it on the
5th. For eleven days those funds must earn a return without being committed:
the firm requires a **liquid**, short-dated, low-risk store of value.

::: incremental
-   Receipts and disbursements are **imperfectly synchronised** — for
    corporations, for banks, and for governments alike.
-   The money market exists to price that mismatch. It is a market for
    **liquidity over short horizons**, not for long-run investment returns.
:::

::: notes
Make the numbers concrete. Eleven days on EUR 40 million at 3% is roughly
EUR 36,000 - real money for doing nothing except not leaving it in a current
account. That is the entire commercial logic of the money market.

Ask who in the room has done a treasury internship. There is usually one, and
they will have watched exactly this.
:::

## 1 · What makes a money market instrument

::: incremental
-   **Short maturity** — under one year, and typically under 120 days. This
    limits both interest-rate risk and the horizon over which credit quality
    can deteriorate.
-   **High liquidity** — the instrument can be sold before maturity without a
    material price concession.
-   **Low default risk** — issuers are sovereigns, large banks and highly rated
    corporations.
-   **Large denominations** — USD 1 million and above. This is a **wholesale**
    market; households participate indirectly, through money market funds.
:::

. . .

::: keyidea
The four properties are not independent. Short maturity makes low default risk
credible; low default risk and standardisation are what sustain a liquid
secondary market. Both parties are managing a **liquidity-timing** problem
rather than seeking a risk-adjusted return — which is why these instruments
price close to the risk-free curve, and why small deviations from it are
informative.
:::

::: notes
The four characteristics are examinable and they hang together: short maturity
is what makes low default risk credible, and low default risk plus large
denominations is what makes the secondary market liquid.

Push on "large denominations". Retail investors reach this market only through
money market funds - which matters later, because those funds breaking the buck
in 2008 is how the plumbing reached ordinary savers.
:::

## 1 · The instruments

::: {.panel-tabset}

### Treasury bills

-   Zero-coupon obligations of the U.S. Treasury, 28 days to 12 months.
-   Issued by **auction**. *Competitive* bids specify a price and may be
    rationed; *noncompetitive* bids accept the clearing price and are filled in
    full.
-   They constitute the **benchmark risk-free asset** of the domestic market:
    every other short rate in this session is quoted relative to them.

### Fed funds

-   Uncollateralised loans of **reserve balances** between depository
    institutions, predominantly **overnight**.
-   Banks use them to satisfy reserve requirements at the close of the
    maintenance period.
-   The **federal funds rate** is the operating target of U.S. monetary policy,
    which makes it the anchor of the short end of the yield curve.

### Repos

-   A **repurchase agreement** is the sale of a security with a simultaneous
    commitment to repurchase it at a fixed price, typically within 3 to 14 days.
-   Economically it is a **collateralised loan**; the repurchase price less the
    sale price is the interest. Legally it is a sale, which is why repo
    receives privileged treatment in bankruptcy.
-   The lender applies a **haircut** — lending less than the collateral's market
    value — as protection against a fall in that value.
-   Unlike fed funds, **non-bank institutions participate**, which is what made
    the repo market both large and, in 2008, a channel for runs.

### CDs and CP

-   **Negotiable certificates of deposit**: bank-issued time deposits,
    USD 100,000 to USD 10 million, transferable before maturity.
-   **Commercial paper**: an **unsecured** promissory note issued by a
    corporation. Maturity is capped at **270 days** because that is the limit of
    the exemption from SEC registration.
-   Being unsecured, commercial paper is a direct claim on the issuer's
    creditworthiness — the feature that matters in section 1's charts.

### BAs and eurodollars

-   A **banker's acceptance** is a time draft on which a commercial bank has
    substituted its own credit for the importer's.
-   This resolves a problem of **asymmetric information** in cross-border trade:
    the exporter is paid at once, bears no exchange-rate exposure, and need not
    evaluate a counterparty it cannot observe. The importer's bank, which can
    observe it, performs that assessment.
-   **Eurodollars** are dollar-denominated deposits held at banks outside the
    United States — a response to the dollar's role as invoicing and settlement
    currency, not to any feature of U.S. regulation alone.

:::

::: notes
Do not read all five tabs. Pick Treasury bills and repos, and let them read the
rest in the handout.

Repos are the one to spend time on. Frame it as a pawn shop: I give you my
securities, you give me cash, and tomorrow I buy them back slightly dearer. The
difference is the interest. Everything about 2008 repo runs follows from asking
what happens when the lender suddenly wants a bigger haircut on the collateral.

Commercial paper being *unsecured* is the line to plant for later.
:::

## 1 · The instruments side by side

| Instrument | Issuer | Buyer | Usual maturity | Secondary market |
|:---|:---|:---|:---|:--:|
| Treasury bills | U.S. government | Consumers and companies | 4, 13, 26 weeks | Excellent |
| Federal funds | Banks | Banks | 1 to 7 days | None |
| Repurchase agreements | Businesses and banks | Businesses and banks | 1 to 15 days | Good |
| Negotiable CDs | Large money-centre banks | Businesses | 14 to 120 days | Good |
| Commercial paper | Finance companies and businesses | Businesses | 1 to 270 days | Poor |
| Banker's acceptance | Banks | Businesses | 30 to 180 days | Good |
| Eurodollar deposits | Non-US banks | Businesses, governments, banks | 1 day to 1 year | Poor |

::: {.source}
The final column is a ranking of secondary-market liquidity. Adapted from
Mishkin, F.S. and Eakins, S.G., *Financial Markets and Institutions*.
:::

::: notes
Do not read the table aloud. Give them thirty seconds to read it, then ask one
question: why does the secondary market column vary so much?

Answer to steer towards: standardisation. Treasury bills are fungible - one is
exactly like another, so anyone will buy yours. Commercial paper is a specific
claim on a specific firm with a specific maturity, so there is no ready buyer.
Fed funds are unsecured overnight loans between named banks - there is nothing
to sell.
:::

## 1 · The rates move together

```{r}
#| fig-width: 12
#| fig-height: 4.4
ids <- c(TB3MS = "3-month Treasury bill",
         FEDFUNDS = "Federal funds",
         MPRIME = "Bank prime rate")
series <- lapply(names(ids), function(id) {
  d <- fred(id)
  if (is.null(d)) return(NULL)
  d$series <- ids[[id]]
  d
})
series <- Filter(Negate(is.null), series)

if (!length(series)) {
  no_data_note("TB3MS / FEDFUNDS / MPRIME")
} else {
  d <- do.call(rbind, series)
  d <- d[d$date >= as.Date("2000-01-01"), ]
  ggplot(d, aes(date, value, colour = series)) +
    recession_bands(from = as.Date("2000-01-01")) +
    geom_line(linewidth = .5) +
    scale_colour_manual(values = unname(unlist(
      IF_COL[c("teal700", "clay", "slate")]))) +
    labs(x = NULL, y = "Percent",
         title = "Money market rates move as one",
         caption = "Shaded: NBER recessions. Source: Federal Reserve via FRED.") +
    theme_if()
}
```

::: ask
The series co-move closely because they share a common policy anchor. Under
what conditions does the **spread** between the federal funds rate and the
Treasury bill rate widen, and what is that spread measuring?
:::

::: notes
The answer is credit and liquidity risk. In a calm market the spread is a few
basis points; in August 2007 and September 2008 it explodes. If the chart has
downloaded, zoom in on 2008 with the chalkboard.
:::

## 1 · Commercial paper outstanding

```{r}
#| fig-width: 12
#| fig-height: 4.2
cp <- fred("COMPOUT")
if (is.null(cp)) {
  no_data_note("COMPOUT")
} else {
  cp$value <- cp$value / 1000
  ggplot(cp, aes(date, value)) +
    recession_bands(from = min(cp$date)) +
    geom_line(colour = IF_COL$teal700, linewidth = .55) +
    labs(x = NULL, y = "USD trillion",
         title = "Commercial paper outstanding",
         caption = "Shaded: NBER recessions. Source: Federal Reserve Board via FRED.") +
    theme_if()
}
```

::: ask
2007 marks a **structural break**, not a cyclical decline: the series steps
down and does not recover its prior level. Which class of issuer exited the
market, why could it not return, and what replaced that funding?
:::

::: notes
Let them look before you talk. Ask what they notice, and wait.

What to draw out: the fall in 2008 is not a dip, it is a step down to a level
the market never regained. The issuers who vanished were the ones funding
long-dated assets with 30-day paper - structured investment vehicles and
finance arms. When the paper stopped rolling, the assets had to be sold or the
sponsoring bank had to take them back on balance sheet. Both happened.

This is the slide that connects to the BNP Paribas opening.
:::

## 1 · Price and rate

Money market instruments are **zero-coupon**: they are purchased below face
value and redeemed at face value, so the entire return is the **discount**.
Two conventions are used to express it, and they are not interchangeable.

:::: columns
::: {.column width="50%"}
The **discount rate** — how far below face value you paid:

$$i_{discount} = \frac{F-P}{F} \times \frac{360}{n}$$
:::

::: {.column width="50%"}
The **annualised yield** — what you actually earned:

$$i_{yt} = \frac{F-P}{P} \times \frac{365}{n}$$
:::
::::

. . .

::: keyidea
One instrument, two measures. The **bank discount basis** deflates by face
value $F$ on a 360-day year — a convention inherited from pre-computational
market practice. The **annualised yield** deflates by the price actually paid,
$P$, on a 365-day year, and is therefore the true holding-period return
expressed annually. Since $P < F$, the yield always exceeds the discount rate.
Only the yield is comparable across instruments.
:::

::: notes
The two formulas are on the exam and students lose marks by mixing them.

Say out loud what makes them different: the discount rate divides by F, the
face value, and uses a 360-day year, because that is the convention the market
inherited. The yield divides by P, what you actually paid, and uses 365. Only
the yield is comparable with a deposit rate or a bond yield.

Ask them which one a seller would prefer to quote. The discount rate is always
the smaller number.
:::

## 1 · The two rates, drawn

```{r}
#| fig-width: 10
#| fig-height: 4.0
P <- seq(97, 100, length.out = 200)
F <- 100; n <- 90
d <- rbind(
  data.frame(P = P, rate = (F - P) / F * 360 / n, which = "Discount rate"),
  data.frame(P = P, rate = (F - P) / P * 365 / n, which = "Annualised yield")
)
ggplot(d, aes(P, rate, colour = which)) +
  geom_line(linewidth = .8) +
  scale_colour_manual(values = c(IF_COL$clay, IF_COL$slate)) +
  scale_y_continuous(labels = function(x) paste0(round(x * 100), "%")) +
  labs(x = "Current price", y = "Rate",
       title = "F = 100, n = 90 days",
       caption = "The gap between the two is the 360 / 365 convention plus the choice of denominator.") +
  theme_if()
```

::: notes
Thirty seconds. The only thing to notice is that the gap widens as the price
falls - the two conventions agree when there is nothing to discount and diverge
as the discount grows.

If nobody asks why the lines are not straight, do not volunteer it.
:::

## 1 · Problem 2.1

::: ask
A 28-day T-bill costs **$996.73** and is worth **$1,000** at maturity.

1.  What is the **discount rate**?
2.  What is the **annualised yield**?
:::

. . .

:::: columns
::: {.column width="50%"}
$$i_{discount} = \frac{1000-996.73}{1000}\times\frac{360}{28} = 4.20\%$$
:::

::: {.column width="50%"}
$$i_{yt} = \frac{1000-996.73}{996.73}\times\frac{365}{28} = 4.28\%$$
:::
::::

::: notes
Make them do it on the calculator, not in their heads. The point of the exercise
is that the two answers differ by 8 basis points on an instrument that everyone
would describe as "the same rate".
:::

## 1 · Problem 2.2

The Treasury auctions **$2.5 billion** of 91-day bills. Competitive bids:

| Bidder | Bid amount | Bid price |
|:--:|:--:|:--:|
| 1 | $500 million | $0.9940 |
| 2 | $750 million | $0.9901 |
| 3 | $1.5 billion | $0.9925 |
| 4 | $1 billion | $0.9936 |
| 5 | $600 million | $0.9939 |

It also received **$750 million** of noncompetitive bids.

::: ask
Who receives bills, how many, and at what price?
:::

::: notes
Order by price descending: 1 (0.9940), 5 (0.9939), 4 (0.9936), 3 (0.9925),
2 (0.9901). Noncompetitive bids are filled first and come off the top: 2.5 - 0.75
= 1.75bn left for competitive bidders. Work down the list until it is exhausted.
Bidder 3 is partially filled and bidder 2 gets nothing.
:::

# 2 · The bond market {.divider .center background-color="#06302F"}

## Athens, March 2012 {.hook background-color="#FAF7F2"}

Greece restructured approximately EUR 200 billion of sovereign bonds. Private
holders accepted a reduction of roughly 53% of face value — the largest
sovereign restructuring on record. The same bonds had carried investment-grade
ratings a few years earlier.

::: incremental
-   A bond is a **contractual claim**: specified payments on specified dates,
    enforceable under a governing law.
-   Enforceability is not payment. **Credit risk** is the probability that the
    contract is not performed, and it is priced.
-   For sovereign debt the issuer can also alter the contract: Greece inserted
    **collective action clauses** retroactively, binding dissenting holders to
    the majority. This has no corporate equivalent.
:::

::: notes
Short and blunt. The Greek restructuring is the counterexample to every
sentence in the next ten slides: the contract was explicit, the payments were
specified, and holders still took roughly 53% off face value.

Worth naming why it was called "voluntary" - collective action clauses were
inserted retroactively, so holdouts were bound by the majority. A sovereign
writing the rules of its own default is a risk with no equivalent in corporate
credit.
:::

## 2 · The vocabulary

| Term | What it means |
|:---|:---|
| Coupon interest rate | The annual interest rate, usually fixed for the whole life |
| Current yield | Coupon payment divided by the current price |
| Face amount / par value | The maturity value |
| Indenture | The contract that specifies the terms of the loan agreement |
| Market rate | The rate currently in effect on securities of like risk and maturity |
| Maturity | Number of periods until the security matures |
| Yield to maturity | The yield if you buy at the current price and hold to maturity |

::: keyidea
The **coupon rate** is fixed in the indenture at issue. The **market rate** on
comparable risk and maturity varies continuously. If the contractual payments
cannot adjust, the only variable that can reconcile the two is the **price**.
The whole of bond valuation follows from that single observation.
:::

::: notes
Do not lecture the table. Give them a minute to read it and take questions.

The line that matters is the key idea: the coupon rate is fixed in the
indenture and never changes; the market rate moves every day. Ask them what has
to give if the contract is fixed and the required return has risen. The answer -
the price - is the whole of bond pricing and they will have said it themselves.

Current yield versus yield to maturity is the distinction they will get wrong
in the exam. Flag it now and again in three slides.
:::

## 2 · The instruments

::: {.panel-tabset}

### Treasuries

-   **Treasury notes**: 1 to 10 years. **Treasury bonds**: 10 to 30 years.
-   Nominal default risk is negligible because the obligation is denominated in
    a currency the issuer creates. That is a statement about **nominal**
    default only.
-   Two risks remain and are examinable: **inflation risk**, which erodes the
    real value of fixed payments, and **interest-rate risk**, the capital loss
    on an existing bond when market rates rise. The latter increases with
    maturity.

### Municipals

-   Issued by local, county and state governments to fund public projects.
-   **Tax free**, which is the whole point. Compare like with like:

    $$\text{after-tax rate} = \text{taxable rate}\times(1-\text{marginal tax rate})$$

-   **General obligation** bonds are repaid from taxes; **revenue** bonds from
    user fees on the project itself.
-   They are **not** default free. Detroit, 2013.

### Corporates

-   Face value usually USD 1,000, interest paid semi-annually.
-   **Restrictive covenants** limit dividends, new debt, ratios — they exist
    because shareholder and bondholder interests conflict.
-   A **call provision** lets the issuer buy the bond back. A **conversion**
    feature lets the holder turn it into equity.
-   **Secured** bondholders have first claim on a specific asset; **unsecured**
    ones do not, and charge more for it.

### Ratings

-   Agencies map credit risk onto an ordinal scale from **Aaa/AAA** downwards.
    Below **Baa3/BBB-** an issue is **speculative grade**.
-   That threshold has effects beyond information. Many institutional mandates
    — insurers, pension funds, some trusts — **prohibit** holdings below it, so
    a one-notch downgrade can compel sales by investors with no independent
    view of the credit.
-   The result is **forced selling** into a falling market: a price movement
    caused by mandate design rather than by fundamentals.
-   Note also the **issuer-pays** model: the entity being rated selects and
    remunerates the agency. The conflict of interest this creates is central to
    the post-2008 literature on structured credit.

:::

::: notes
Municipals and ratings are the two tabs worth your time.

On municipals: the tax comparison is a favourite exam question and problem 2.3
is coming, so do the algebra once here. Detroit 2013 is the counterexample to
"municipal bonds are safe" - pensioners and bondholders in the same queue.

On ratings: the investment grade line is the thing to hammer. A downgrade from
Baa3 to Ba1 is one notch of credit opinion, but it forces selling by funds whose
mandates prohibit speculative grade. Price falls for a reason that has nothing
to do with the issuer's prospects. That is a market structure effect, not a
credit effect, and it is the kind of distinction that separates a good answer
from an average one.
:::

## 2 · The rating scales

:::: columns
::: {.column width="52%"}
| Moody's | S&P | Fitch | Meaning |
|:--:|:--:|:--:|:---|
| Aaa | AAA | AAA | Best quality, least credit risk |
| Aa1-Aa3 | AA+/AA/AA- | AA+/AA/AA- | High quality, very strong capacity |
| A1-A3 | A+/A/A- | A+/A/A- | Upper-medium grade |
| Baa1-Baa3 | BBB+/BBB/BBB- | BBB+/BBB/BBB- | Medium quality, adequate capacity |
:::

::: {.column width="48%"}
| Moody's | S&P | Fitch | Meaning |
|:--:|:--:|:--:|:---|
| Ba1-Ba3 | BB+/BB/BB- | BB+/BB/BB- | Speculative |
| B1-B3 | B+/B/B- | B+/B/B- | Very speculative |
| Caa-C | CCC/CC/C | CCC/CC/C | Highly speculative, near or in default |
| — | D | DDD/DD/D | In default |
:::
::::

::: ask
The investment-grade boundary lies between Baa3 and Ba1. Ratings are ordinal
and credit quality is continuous, so why should **one notch** produce a
discontinuity in price?
:::

::: notes
Two minutes maximum. They do not need to memorise the ladder; they need to know
where the line is and why it matters.

Ask the question on the slide and wait. If nobody bites: a pension fund
mandated to hold only investment grade must sell on the downgrade, whatever it
thinks. Forced sellers meeting no natural buyers is how a one-notch change moves
a price several points.

The term "fallen angel" is worth giving them.
:::

## 2 · Rates by credit quality

```{r}
#| fig-width: 12
#| fig-height: 4.4
ids <- c(AAA = "Corporate Aaa", BAA = "Corporate Baa", GS20 = "20-year Treasury")
series <- lapply(names(ids), function(id) {
  d <- fred(id)
  if (is.null(d)) return(NULL)
  d$series <- ids[[id]]
  d
})
series <- Filter(Negate(is.null), series)

if (!length(series)) {
  no_data_note("AAA / BAA / GS20")
} else {
  d <- do.call(rbind, series)
  d <- d[d$date >= as.Date("1980-01-01"), ]
  ggplot(d, aes(date, value, colour = series)) +
    recession_bands(from = as.Date("1980-01-01")) +
    geom_line(linewidth = .5) +
    scale_colour_manual(values = unname(unlist(
      IF_COL[c("slate", "clay", "teal700")]))) +
    labs(x = NULL, y = "Percent",
         title = "The credit spread is the distance between these lines",
         caption = "Shaded: NBER recessions. Source: Moody's and the Treasury via FRED.") +
    theme_if()
}
```

::: ask
The spread is widest precisely when firms most need external finance. Identify
the two distinct mechanisms behind this, and explain why their coincidence
makes credit supply **procyclical**.
:::

::: notes
The lines are the lesson: the distance between Baa and Treasury is the credit
spread, and it is not constant.

Point at 2008 and at 2020. Ask the question on the slide and let it sit -
spreads widen exactly when firms most need to borrow, because that is when
default probability rises and risk appetite falls at the same time. Credit is
procyclical, and that is a policy problem, not a market quirk.

If the chart has not downloaded, say so and use the numbers: Baa-Treasury went
from about 2 points to over 6 in late 2008.
:::

## 2 · Price and yield

The price of a bond is the present value of everything it will pay you:

$$P = \frac{C}{(1+r)} + \frac{C}{(1+r)^2} + \dots + \frac{C}{(1+r)^n} + \frac{F}{(1+r)^n}$$

. . .

The **yield to maturity** is the internal rate of return $r$ that equates the
present value of the contractual payments to the observed market price:

$$0 = -P + \sum_{t=1}^{n}\frac{C}{(1+r)^t} + \frac{F}{(1+r)^n}$$

. . .

The **current (coupon) yield** is an approximation that ignores any capital
gain or loss at redemption:

$$r_c = \frac{C}{P}$$

::: keyidea
Price and yield are **inversely related**, necessarily: the payments are fixed,
so a higher return can only be obtained at a lower price. Note the assumptions
embedded in the yield to maturity — that the bond is **held to maturity** and
that all coupons are **reinvested at $r$**. Neither holds in general, which is
why realised returns differ from quoted yields.
:::

::: notes
Build it, do not present it. Start with a one-year bond and a single cash flow,
then add periods until the formula on the slide appears.

The yield to maturity equation is implicit - there is no closed form for r
beyond two periods. Tell them that honestly: this is why the exam gives them
either a short maturity or a calculator, and why in practice a machine solves it.

End on the key idea and make them say it back: if the payments are fixed, the
only way to earn more is to pay less.
:::

## 2 · Coupon yield is only an approximation

```{r}
#| fig-width: 10
#| fig-height: 4.0
Pr <- seq(50, 150, length.out = 300)
F <- 100; C <- 5; n <- 10
ytm <- sapply(Pr, function(p) {
  f <- function(r) -p + sum(C / (1 + r)^(1:n)) + F / (1 + r)^n
  tryCatch(uniroot(f, c(-0.5, 1))$root, error = function(e) NA_real_)
})
d <- rbind(
  data.frame(P = Pr, rate = ytm,    which = "Yield to maturity"),
  data.frame(P = Pr, rate = C / Pr, which = "Coupon yield")
)
ggplot(d, aes(P, rate, colour = which)) +
  geom_hline(yintercept = 0, colour = IF_COL$muted, linewidth = .3) +
  geom_vline(xintercept = 100, linetype = "dotted", colour = IF_COL$muted) +
  geom_line(linewidth = .8) +
  scale_colour_manual(values = c(IF_COL$clay, IF_COL$slate)) +
  scale_y_continuous(labels = function(x) paste0(round(x * 100), "%")) +
  labs(x = "Current price", y = NULL,
       title = "F = 100, coupon 5, 10 years to maturity",
       caption = "The two agree only at par. Away from par the coupon yield misleads.") +
  theme_if()
```

::: notes
The two curves cross exactly at par, where price equals face value. Everywhere
else the coupon yield is wrong, and it is wrong in a predictable direction.

Ask which direction and why. Below par the coupon yield understates the return
because it ignores the capital gain you make when the bond redeems at 100.
Above par it overstates, because it ignores the capital loss.

That is problem 2.4 part one in a sentence.
:::

## 2 · Problem 2.3

::: ask
1.  A corporate bond yields **9%**, a municipal bond **6.75%**. Which do you buy?
2.  A corporate bond yields **5%**, a municipal **3.5%**, and your marginal tax
    rate is **28%**. Which now?
:::

. . .

:::: columns
::: {.column width="50%"}
**First case.** The comparison needs a tax rate to be meaningful. At a marginal
rate of 25%, the corporate is worth $9\% \times 0.75 = 6.75\%$ after tax —
exactly the municipal. Above 25% the municipal wins.
:::

::: {.column width="50%"}
**Second case.** $5\% \times (1-0.28) = 3.60\%$ after tax, against **3.5%**
tax-free. The corporate wins, narrowly.
:::
::::

::: notes
The first part has no numerical answer until you ask them for the tax rate.
That is deliberate: it is the question they should have asked. Let someone try
to answer before pointing it out.
:::

## 2 · Problem 2.4

::: ask
1.  What is the **coupon yield** on a bond with face value $1,000, current
    price $921.01 and a coupon rate of **10.95%**?
2.  What is the **price** of a two-year, 10% coupon bond paying **semi-annually**,
    face value $1,000, if the required return is **12%**?
:::

. . .

:::: columns
::: {.column width="46%"}
$$r_c = \frac{109.50}{921.01} = 11.89\%$$
:::

::: {.column width="54%"}
Four periods, coupon 50, discount rate 6% per period:

$$P = \sum_{t=1}^{4}\frac{50}{1.06^t} + \frac{1000}{1.06^4} = 965.35$$
:::
::::

::: notes
The semi-annual part is where they lose marks: halve the coupon AND halve the
rate AND double the number of periods. Three changes, not one.
:::

# 3 · The stock market {.divider .center background-color="#06302F"}

## The residual claim {.hook background-color="#FAF7F2"}

A bondholder holds a claim to a fixed sum. A shareholder holds a claim to
whatever remains once all prior claims are satisfied — a quantity bounded below
by zero and unbounded above.

::: incremental
-   Equity returns arise from **dividends** and **capital appreciation**.
-   Shareholders are **residual claimants** on the firm's assets and cash flows,
    and hold **control rights** through the vote.
-   Ranking last in liquidation is what gives equity its asymmetric payoff, and
    what makes the **agency relationship** between managers and shareholders a
    central problem of corporate finance.
:::

::: notes
Short slide, one idea: the bondholder is owed a number, the shareholder owns the
remainder. Everything that follows - voting rights, dividend policy, valuation
by discounting dividends - comes from being last in the queue.

Ask which they would rather hold in a firm about to be liquidated, and which in
a firm about to triple in size. The answer differing is the point.
:::

## 3 · Common, preferred, and foreign

:::: columns
::: {.column width="50%"}
**Common stock** — the right to vote for directors and on certain issues, and
to receive dividends if the board declares them.

**Preferred stock** — usually **no vote**, but a **fixed dividend** paid before
common. Closer to a bond than most students expect.
:::

::: {.column width="50%"}
**Foreign stocks** are useful for diversification but awkward to buy if they are
not listed locally.

**American depository receipts (ADRs)** solve it: a bank buys the foreign
shares and issues receipts against them that trade on U.S. exchanges.
:::
::::

::: keyidea
An **index** is a summary statistic for a defined set of securities, not an
asset. Its **weighting scheme** determines what the statistic measures — and
therefore what a given movement in it means.
:::

::: notes
Preferred stock is the one worth a minute: fixed dividend, paid ahead of common,
usually no vote. Ask whether that is equity or debt. The honest answer is that
it sits between them, which is why accounting and ratings treat it
inconsistently.

ADRs connect back to session 1 - a claim on a foreign asset, repackaged so it
settles in dollars under domestic law. Same engineering as a eurodollar deposit.

On indexes: an index is a summary, not an investment. The weighting decides what
it summarises, which is the next slide.
:::

## 3 · Where the trading happens

:::: columns
::: {.column width="50%"}
**Organised exchanges**

-   NYSE trades roughly 4 to 7 billion shares a day.
-   "Organised" once meant a physical floor; electronic networks have largely
    replaced that idea.
-   **Listing requirements** exclude small firms.
-   Auction market with floor specialists; around 25% of trades filled directly
    by the specialist.
:::

::: {.column width="50%"}
**Over the counter**

-   NASDAQ is the obvious example, with roughly 3,000 securities listed.
-   **Dealers** stand ready to make a market — without one, the equity would be
    hard to trade at all.
-   Multiple market makers post bid and ask prices for the same security.
:::
::::

::: ask
Dealers quote a **bid-ask spread** and profit from it. Competition constrains
that spread — but a spread of zero would not be sustainable either. What costs
must the spread compensate the dealer for, and what happens to liquidity when
those costs rise?
:::

::: notes
The exchange-versus-OTC distinction is less sharp than the textbook makes it -
say so. Most NYSE volume is electronic and NASDAQ has a central book.

The question on the slide is the useful one. A dealer quoting a bad price simply
gets no trades, because the buyer goes to another market maker. Competition
between dealers is what replaces the auction. Then push: what if there is only
one dealer willing to quote? That is a thin market, and it is why liquidity
disappears exactly when it is needed.
:::

## 3 · Price-weighted versus value-weighted

The **DJIA** is a price-weighted average of 30 stocks. The **S&P 500** is a
market-value-weighted index of 500.

```{r}
#| fig-width: 12
#| fig-height: 3.9
dj <- fred("DJIA")
sp <- fred("SP500")
if (is.null(dj)) {
  no_data_note("DJIA")
} else {
  dj <- dj[dj$date >= as.Date("2015-01-01"), ]
  dj$index <- 100 * dj$value / dj$value[1]
  dj$series <- "Dow Jones Industrial Average"
  d <- dj[, c("date", "index", "series")]
  if (!is.null(sp)) {
    sp <- sp[sp$date >= as.Date("2015-01-01"), ]
    sp$index <- 100 * sp$value / sp$value[1]
    sp$series <- "S&P 500"
    d <- rbind(d, sp[, c("date", "index", "series")])
  }
  ggplot(d, aes(date, index, colour = series)) +
    geom_line(linewidth = .5) +
    scale_colour_manual(values = c(IF_COL$slate, IF_COL$teal500)) +
    labs(x = NULL, y = "Index, first observation = 100",
         title = "Thirty stocks by price, or five hundred by market value",
         caption = "Source: S&P Dow Jones Indices via FRED.") +
    theme_if()
}
```

::: ask
Under price weighting, a USD 400 constituent carries roughly thirteen times the
weight of a USD 30 constituent, irrespective of the firms' market values. State
the implication for a **stock split**, and hence whether the index measures
anything economically meaningful.
:::

::: notes
The arithmetic first: in a price-weighted index a 10% move in a USD 400 stock
moves the index thirteen times as much as a 10% move in a USD 30 stock, whatever
the two companies are worth. In a value-weighted index the weights are market
capitalisations, which is what an investor holding the market would actually own.

The DJIA is a historical artifact kept for continuity. Ask why anyone still
quotes it. The honest answer is journalism, not finance.

Stock splits are the killer detail: a split changes a price-weighted index's
composition without anything happening to any company.
:::

## 3 · What a share is worth

The **Gordon growth model** — a special case of the dividend discount model, in
which dividends grow at a constant rate $g$ in perpetuity and $k_e > g$:

$$P = \sum_{t=1}^{\infty}\frac{Div_t}{(1+k_e)^t} = \frac{Div_1}{k_e - g}$$

:::: columns
::: {.column width="50%"}
**Geometric growth**, for backing out $g$ from two observations:

$$g = \left(\frac{\text{final}}{\text{initial}}\right)^{1/n} - 1$$
:::

::: {.column width="50%"}
**Price-to-earnings**, a relative valuation measure:

$$PE = \frac{\text{Price}}{\text{Earnings}}$$
:::
::::

::: keyidea
The denominator $k_e - g$ governs everything. As $g \to k_e$ the valuation
diverges, so near that boundary the price is **highly sensitive** to small
revisions in either parameter. This is not a defect of the model but a property
of perpetual growth — and it explains why estimates of high-growth firms
disperse so widely, and why such valuations reprice violently on modest news.
:::

::: notes
Derive the Gordon model rather than asserting it - a growing perpetuity, three
lines of algebra, and they have seen it in corporate finance.

The condition k > g is not a technicality. If a firm could grow faster than the
required return forever it would eventually be the whole economy. Say that out
loud; it is the sanity check that catches a wrong answer in an exam.

PE ratio: a shortcut whose numerator is a market price and whose denominator is
an accounting number. Ask what a high PE means. Both "expensive" and "expected
to grow" are right, and you cannot tell which from the ratio alone.
:::

## 3 · The denominator does all the work

```{r}
#| fig-width: 12
#| fig-height: 4.0
div1 <- 2
a <- data.frame(x = seq(.06, .24, length.out = 200))
a$price <- div1 / (a$x - .05); a$panel <- "Varying the required return (g = 5%)"
b <- data.frame(x = seq(.01, .19, length.out = 200))
b$price <- div1 / (.20 - b$x); b$panel <- "Varying the growth rate (k = 20%)"
d <- rbind(a, b)
ggplot(d, aes(x, price)) +
  geom_line(colour = IF_COL$clay, linewidth = .8) +
  facet_wrap(~panel, scales = "free_x") +
  scale_x_continuous(labels = function(x) paste0(round(x * 100), "%")) +
  coord_cartesian(ylim = c(0, 250)) +
  labs(x = NULL, y = "Stock price",
       title = "Gordon growth, with the first dividend fixed at 2",
       caption = "Both curves are the same asymptote seen from two directions.") +
  theme_if() +
  theme(strip.text = element_text(colour = IF_COL$teal700, face = "bold"))
```

::: notes
Both panels are the same asymptote seen from two directions. As k falls towards
g, or g rises towards k, the price goes vertical.

The lesson is about fragility, not maths: near the asymptote a small revision to
a growth assumption produces an enormous revision to the price. That is why
high-growth valuations move so violently on small news, and why two analysts
with almost identical assumptions can publish wildly different target prices.

Connect it forward: this is the same sensitivity that made the dot-com and the
2021 growth repricings so severe.
:::

## 3 · Problem 2.5

::: ask
1.  A stock is expected to pay a dividend of **$0.16** and to be worth **$60**
    in a year. At a **12%** discount rate, what is it worth now?
2.  If the industry PE ratio is **16** and the firm earns **$1.13 per share**,
    what is the price?
:::

. . .

:::: columns
::: {.column width="50%"}
$$P_0 = \frac{0.16 + 60}{1.12} = 53.71$$
:::

::: {.column width="50%"}
$$P = 16 \times 1.13 = 18.08$$
:::
::::

::: notes
Both parts are one line each - the point is that they are different methods
answering the same question.

Part 1 is a discounted cash flow: you forecast, then discount. Part 2 is
relative valuation: you assume the market has priced comparable firms correctly
and apply their multiple.

Ask which they trust more, and why they disagree. That argument is the whole of
equity research.
:::

## 3 · Problem 2.6

::: ask
The DJIA was **10,000** in 2010 and **20,000** in 2017. What was the annual
**geometric growth rate**?
:::

. . .

$$g = \left(\frac{20000}{10000}\right)^{1/7} - 1 = 2^{1/7} - 1 = 10.41\%$$

::: notes
Watch for students using 8 years instead of 7. The exponent is the number of
*periods between* the observations, not the number of years mentioned.
:::

## Next session {.center}

Session 4 finishes securities markets: **mortgages**, **foreign exchange** and
**derivatives**, then the green bonds discussion.

::: keyidea
Bring **Sartzetakis (2021)** on green bonds, read — and the security your group
wants to analyse. Not "a stock". A specific one, with a ticker.
:::

::: {.source}
Press **`O`** for the slide overview · **`C`** to draw · **`S`** for speaker notes · **`F`** for fullscreen · **`E`** then print for a PDF of the slides.
:::

::: notes
Leave two minutes for this. Read the paper before the session, not after -
the discussion is the assessment-relevant part.

The group work reminder is the one that saves you grief later: a specific
security with a ticker, chosen now, not the week before session 8.
:::

<!-- ===================================================================
     SESSION 4 - not rendered yet. Uncomment, or split into its own
     folder the way session 2 was, when you are ready.
     Sections 4 (mortgages), 5 (foreign exchange), 6 (derivatives)
     and 7 (Sartzetakis discussion).
     =================================================================== -->

<!--

# 4 · Mortgage markets {.divider .center background-color="#06302F"}
## The market that ate the world {.hook background-color="#FAF7F2"}

By 2006 an American mortgage was not a loan a bank held. It was an input:
pooled, tranched, rated, and sold to a German Landesbank that had never seen
Ohio.

::: incremental
-   A mortgage is a **long-term loan secured by real estate**.
-   It is **amortised**: one fixed payment covers both interest and principal
    every month.
:::
## 4 · The types

| Type | What it is |
|:---|:---|
| Conventional | Not guaranteed; usually needs private mortgage insurance and 5-20% down |
| Insured | Guaranteed by a government agency; low or zero down payment |
| Fixed-rate | The rate is fixed for the life of the mortgage |
| Adjustable-rate (ARM) | Rate tied to another security, adjusted periodically within annual limits |
| Graduated-payment (GPM) | Low initial payment rising each year; amortises in 30 years |
| Growing-equity (GEM) | Payment rises each year; amortises in less than 30 years |
| Second mortgage | Secured by a second lien; used for credit lines and home improvement |
| Reverse annuity (RAM) | The lender pays the borrower monthly; due when the property is sold |

::: ask
Which of these transfers **interest rate risk** from the lender to the borrower?
And who tends to be sold those?
:::
## 4 · Securitisation, in four steps

::: incremental
1.  The **secondary mortgage market** was created by the federal government
    after WWII, when it set up Fannie Mae. It grew quickly from the mid-1980s.
2.  Individual mortgages default unpredictably, so they are **pooled** —
    hundreds at a time — on the argument that diversification tames the risk.
3.  Rights to the pool's cash flows are sold as separate securities:
    **pass-throughs**, or mortgage-backed securities.
4.  The borrower's monthly payment passes through a trustee before reaching the
    investors.
:::

. . .

::: ask
Step 2 is the load-bearing one. Diversification removes risk that is
**independent** across loans. What happens to a pool of American mortgages if
house prices fall everywhere at once?
:::

::: notes
This is the hinge of the whole session and it pays to slow down. The 2008
answer: the correlation everyone assumed was near zero turned out to be near
one, and every tranche above the bottom was priced on that assumption.
:::
## 4 · Mortgage rates track long Treasuries

```{r}
#| fig-width: 12
#| fig-height: 4.4
ids <- c(MORTGAGE30US = "30-year fixed mortgage", GS10 = "10-year Treasury")
series <- lapply(names(ids), function(id) {
  d <- fred(id)
  if (is.null(d)) return(NULL)
  d$series <- ids[[id]]
  d
})
series <- Filter(Negate(is.null), series)

if (!length(series)) {
  no_data_note("MORTGAGE30US / GS10")
} else {
  d <- do.call(rbind, series)
  ggplot(d, aes(date, value, colour = series)) +
    recession_bands(from = min(d$date)) +
    geom_line(linewidth = .5) +
    scale_colour_manual(values = c(IF_COL$slate, IF_COL$clay)) +
    labs(x = NULL, y = "Percent",
         title = "The mortgage rate is a spread over the long Treasury",
         caption = "Shaded: NBER recessions. Source: Freddie Mac and the Treasury via FRED.") +
    theme_if()
}
```

::: ask
The gap between the two is not constant. What widens it?
:::
## 4 · Loan and annuity

:::: columns
::: {.column width="50%"}
What you can **borrow**, given what you can pay each period:

$$C_0 = a \times \left[\frac{1-(1+i)^{-T}}{i}\right]$$
:::

::: {.column width="50%"}
What you must **pay** each period, given what you borrowed:

$$a = C_0 \times \left[\frac{i}{1-(1+i)^{-T}}\right]$$
:::
::::

```{r}
#| fig-width: 10
#| fig-height: 3.3
i <- 0.05 / 12; T <- 20 * 12
d <- data.frame(a = seq(1000, 3000, by = 25))
d$loan <- d$a * (1 - (1 + i)^(-T)) / i
ggplot(d, aes(a, loan)) +
  geom_line(colour = IF_COL$slate, linewidth = .8) +
  scale_y_continuous(labels = function(x) format(x, big.mark = " ", scientific = FALSE)) +
  labs(x = "Monthly payment", y = "Loan you can support",
       title = "5% nominal, 20 years: i = 0.417% per month, T = 240 months",
       caption = "Linear in the payment, which is why affordability is really about the rate.") +
  theme_if()
```
## 4 · Problem 2.7

::: ask
You want to borrow **$100,000**. Choose between a **12% 30-year** mortgage and
an **11.5%** mortgage with **2 discount points**. Which?
:::

::: {.smaller}
A discount point is 1% of the loan paid up front: on a $200,000 loan, 1 point
costs $2,000, half a point $1,000, a quarter point $500.
:::

. . .

::: keyidea
Two points on $100,000 is $2,000 today against 50 basis points saved every month
for thirty years. Compute both monthly payments with the annuity formula, then
ask how long it takes the saving to repay the $2,000. **The answer depends
entirely on how long you stay in the house.**
:::

::: notes
This is the one problem in the session with a genuinely conditional answer.
If they move in four years the points are a bad deal; if they hold to maturity
they are a good one. Getting them to state the break-even horizon is the goal.
:::
# 5 · The foreign exchange market {.divider .center background-color="#06302F"}
## The biggest market nobody sees {.hook background-color="#FAF7F2"}

No building, no opening bell, no ticker on the news. Over-the-counter, bank to
bank, and turnover in the **trillions of dollars a day**.

::: incremental
-   What is traded is **bank deposits denominated in different currencies**.
-   Trades are large — typically in excess of USD 1 million.
-   We met the price in session 2. Now we meet the market.
:::
## 5 · Purchasing power parity

**Absolute PPP** — the exchange rate equals the ratio of price levels:

$$E_{\$/\mathrm{EUR}} = \frac{P_{US}}{P_{EUR}}$$

. . .

**Relative PPP** — the rate of depreciation equals the **inflation differential**:

$$\frac{\Delta E_{\$/\mathrm{EUR}}}{E_{\$/\mathrm{EUR}}} = \pi_{US} - \pi_{EUR}$$

. . .

::: ask
PPP works in the long run and fails in the short run. Given what it assumes,
which of its assumptions is the one that fails first?
:::

::: notes
Answers to steer towards: differentiated goods with brand names and legal
protection give firms pricing power, so identical goods do not have identical
prices; and prices are sticky, so they do not adjust quickly to market
conditions. Both break the arbitrage that PPP relies on.
:::
## 5 · Spot, forward, and what moves them

:::: columns
::: {.column width="46%"}
**Spot rate** — near-immediate exchange of bank deposits.

**Forward rate** — exchange at an agreed future date, at a price agreed now.
:::

::: {.column width="54%"}
Rates move when money demand or supply moves, which happens on expected changes in:

-   **relative price levels**
-   **trade barriers** — tariffs and quotas
-   **preferences** for domestic versus foreign goods
-   **relative productivity**
:::
::::

. . .

::: keyidea
Every item on that list is an **expected change**. The spot rate today is
mostly a statement about tomorrow.
:::
## 5 · What an appreciation does

::: incremental
-   Domestic goods become **more expensive abroad**; foreign goods become
    cheaper at home.
-   Domestic businesses become **less competitive**.
-   Domestic consumers gain **purchasing power**.
:::

. . .

::: ask
So is a strong currency good news? Ask a French exporter and a French
holidaymaker on the same day and you get two answers.
:::
# 6 · Derivative markets {.divider .center background-color="#06302F"}
## A price for a promise {.hook background-color="#FAF7F2"}

In 1995 Barings Bank, 233 years old, was destroyed in weeks by one trader's
position in Nikkei futures. The instruments were not exotic. The position was
simply unhedged and unsupervised.

::: incremental
-   **Hedging** means using a financial transaction to reduce or remove a risk.
-   The same instrument that hedges a position can create one many times larger.
:::
## 6 · Long, short, and offsetting

:::: columns
::: {.column width="50%"}
A **long** position: you own the asset and expect the price to rise.

A **short** position: you do not own it — you borrow it, sell it, and expect to
buy it back cheaper.
:::

::: {.column width="50%"}
To **hedge**:

-   a long position is offset by taking an additional **short** position
-   a short position is offset by taking an additional **long** position
:::
::::

::: keyidea
Hedging is not the absence of a position. It is **two** positions chosen so
that what one loses, the other gains.
:::
## 6 · The instruments

::: {.panel-tabset}

### Forwards

-   An agreement between two parties to transact at a **future point in time**.
-   Specifies the exact assets, the price, and the date of exchange.
-   **Long** = agree to buy later. **Short** = agree to sell later.
-   Private, bilateral, and therefore exposed to the other party defaulting.

### Futures

-   Same idea, but traded on an **organised exchange** rather than privately.
-   **Standardised**, and therefore far more liquid than forwards.
-   **Marked to market daily**, which is what removes most of the default risk.
-   No need to deliver: positions are netted in cash.
-   Harder to corner, because the deliverable range of securities is wide.
-   Regulated in the U.S. by the CFTC.

### Options

-   The **right, not the obligation**, to buy (**call**) or sell (**put**) at
    the strike price.
-   **American** options can be exercised up to expiry; **European** options only
    on the expiry date.
-   You pay a **premium** for that right. The premium rises with
    **time to expiry** and with **volatility**, and moves with the strike.

### Swaps

-   An **interest-rate swap** exchanges one stream of interest payments for
    another, in the same currency.
-   The **plain vanilla** version specifies the rates exchanged, the type of
    payments, and the notional amount.
-   No balance-sheet change, longer-dated than futures or options — but
    **illiquid** and exposed to **default risk**.

### Credit

-   **Credit derivatives** pay off on changes in credit events. The notional
    outstanding is in the trillions.
-   **Credit options** are tied to ratings or spreads.
-   **Credit swaps** exchange payments on loan portfolios, letting lenders
    specialise locally while diversifying.
-   A **credit default swap** pays off option-like when a basket of loans
    defaults. Remember that one.

:::
## 6 · Why the payoff diagram is the whole lesson

```{r}
#| fig-width: 12
#| fig-height: 4.2
S <- seq(105, 130, length.out = 300)
K <- 115; prem <- 2000; mult <- 400
call <- data.frame(S = S, profit = pmax(S - K, 0) * mult - prem,
                   who = "Buyer of call option", panel = "Call option and long future")
futl <- data.frame(S = S, profit = (S - K) * mult,
                   who = "Buyer of futures", panel = "Call option and long future")
put  <- data.frame(S = S, profit = pmax(K - S, 0) * mult - prem,
                   who = "Buyer of put option", panel = "Put option and short future")
futs <- data.frame(S = S, profit = (K - S) * mult,
                   who = "Seller of futures", panel = "Put option and short future")
d <- rbind(call, futl, put, futs)

ggplot(d, aes(S, profit, colour = who)) +
  geom_hline(yintercept = 0, colour = IF_COL$muted, linewidth = .3) +
  geom_line(linewidth = .8) +
  facet_wrap(~panel) +
  scale_colour_manual(values = unname(unlist(
    IF_COL[c("clay", "slate", "clay", "slate")]))) +
  scale_y_continuous(labels = function(x) format(x, big.mark = " ", scientific = FALSE)) +
  labs(x = "Price at expiration", y = "Profit",
       title = "The option has a floor. The future does not.",
       caption = "Strike 115, premium 2 000 per contract.") +
  theme_if() +
  theme(strip.text = element_text(colour = IF_COL$teal700, face = "bold"))
```

::: ask
The option line **bends** and the futures line does not. What did you pay for
that bend, and when is it worth it?
:::
## 6 · Problem 2.8

Rock Solid holds a **$100 million** stock portfolio that tracks the S&P 500.
The manager fears a decline and wants to hedge the downside completely.

::: ask
1.  The S&P is at **1,000**. How is the hedge done with **futures**?
2.  What happens if a year later the index is at **900**?
3.  And if it is at **1,100**?
:::

::: {.smaller}
An index futures contract has a multiplier of **250**, so one contract is worth
$250 \times 1000 = \$250{,}000$.
:::

::: notes
400 contracts sold short. At 900 the portfolio loses 10m and the futures gain
10m. At 1100 the portfolio gains 10m and the futures lose 10m. The hedge works
in both directions — which is the point students miss. They think a hedge
protects the downside and keeps the upside.
:::
## 6 · Problem 2.9

Same portfolio, same fear. Now the S&P is at **1,000** and a strike of **950**
costs a premium of **$200 per contract**.

::: ask
1.  How is the hedge done with **options**?
2.  What happens at **900**?
3.  What happens at **1,100**?
:::

::: {.smaller}
An index option contract has a multiplier of **100**, so one contract is worth
$100 \times 1000 = \$100{,}000$.
:::

. . .

::: keyidea
Compare with 2.8. The option costs money up front and leaves a gap down to 950
uncovered — but at 1,100 you keep the gain. **That is what the premium bought.**
:::
# 7 · Reading discussion {.divider .center background-color="#06302F"}
## Does a green label change the price? {.hook background-color="#FAF7F2"}

A green bond funds an environmental project. Otherwise it is an ordinary bond
from an ordinary issuer, with the same claim on the same balance sheet.

::: incremental
-   So if the cash flows and the credit risk are identical, should it trade at a
    different yield?
-   If it should not, what exactly is the instrument for?
:::
## 7 · Sartzetakis (2021), green bonds and the low-carbon transition

::: {.panel-tabset}

### The questions

Come with an answer to each of these. I will ask.

1.  What are the **advantages and disadvantages of enlarging the mandate of the
    central bank**?
2.  Which projects can be considered **"green"** and financed by green bonds?
3.  What is the process for **project evaluation and selection**?
4.  What are the **positive developments** in the green bond market, and the
    **limits** to its development?

### How to prepare

-   Question 1 is the contested one. A central bank with a climate mandate is
    making distributional choices it was not elected to make — against the
    argument that climate risk is financial risk and therefore already in its
    remit.
-   Questions 2 and 3 are where **greenwashing** lives. If the issuer defines
    "green" and the issuer picks the verifier, what is the label worth?
-   For question 4, separate market **growth** from market **effect**. A market
    can grow quickly and still finance nothing that would not have been financed
    anyway.

:::

::: notes
Run this as a discussion, not a lecture. Ten minutes in pairs on questions 2 and
3, then open the floor on 1. Question 4 is the one that connects back to the
bond pricing at the start of the session: if the green label does not change the
yield, the issuer has no incentive beyond reputation.
:::

-->
